---
title: "control-plot-species"
author: "Anjum K. Gujral"
date: "2026-07-28"
output: html_document
---

# load data
```{r}
# all COND, PLOT, TREE, and SEEDLING data should be loaded
```

# CA
## whole plot network
```{r}
# create a unique PLOT ID in COND and in PLOT tables 
CA_COND_IDs <- CA_COND %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )
CA_PLOT_IDs <- CA_PLOT %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )
# create unique plot ids for CA_TREE
CA_TREE_IDs <- CA_TREE %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )
# create unique plot ids for CA_SEEDLING
CA_SEEDLING_IDs <- CA_SEEDLING %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

# filter CA_COND_IDs for forest types only, no disturbance filtering
CA_COND_forest_type <- CA_COND_IDs %>%
  filter(FORTYPCD %in% c(185, 182, 184,
201, 202, 221, 222, 224, 225, 226, 
241, 261, 262, 263, 264, 265, 267, 268, 270,
271, 281, 301, 304, 305, 321, 
341, 342, 361, 365, 371,
901, 911, 912, 922, 923, 933, 941, 961))
           

#(371, 201, 221, 222, 224, 225, 261, 922))

# get the unique plot_ids that qualify
CA_forest_type_plot_id <- CA_COND_forest_type %>%
  distinct(plot_id) %>%
  pull(plot_id)

length(CA_forest_type_plot_id)

# filter CA_PLOT_IDs for those plot_ids
CA_PLOT_forest_type <- CA_PLOT_IDs %>%
  filter(plot_id %in% CA_forest_type_plot_id)

# filter CA_TREE_IDs for those plot_ids
CA_TREE_forest_type <- CA_TREE_IDs %>%
  filter(plot_id %in% CA_forest_type_plot_id)

nrow(CA_PLOT_forest_type)
n_distinct(CA_TREE_forest_type$plot_id)

CA_TREE_forest_type_species <- CA_TREE_forest_type %>%
  left_join(
    REF_SPECIES %>%
      mutate(species = paste(GENUS, SPECIES, sep = "_")) %>%
      dplyr::select(SPCD, species),
    by = "SPCD"
  )
sum(is.na(CA_TREE_forest_type_species$species))
n_distinct(CA_TREE_forest_type_species$species)

# species from your new forest-type-only workflow
CA_forest_type_all_species <- CA_TREE_forest_type_species %>%
  distinct(species) %>%
  pull(species)

length(CA_forest_type_all_species)

# make sure the species coming out of SEEDLING match TREE
# filter CA_SEEDLING_IDs for those plot_ids
CA_SEEDLING_forest_type <- CA_SEEDLING_IDs %>%
  filter(plot_id %in% CA_forest_type_plot_id)

CA_SEEDLING_forest_type_species <- CA_SEEDLING_forest_type %>%
  left_join(
    REF_SPECIES %>%
      mutate(species = paste(GENUS, SPECIES, sep = "_")) %>%
      dplyr::select(SPCD, species),
    by = "SPCD"
  )

# species list from seedling data (forest-type-only workflow)
CA_forest_type_all_seedling_species <- CA_SEEDLING_forest_type_species %>%
  distinct(species) %>%
  pull(species)

length(CA_forest_type_all_species)
length(CA_forest_type_all_seedling_species)

# species found in both tree and seedling data
intersect(CA_forest_type_all_species, CA_forest_type_all_seedling_species)

# species found only in tree data (mature trees, no seedlings recorded)
setdiff(CA_forest_type_all_species, CA_forest_type_all_seedling_species)

# species found only in seedling data (recruiting, but no mature trees recorded)
setdiff(CA_forest_type_all_seedling_species, CA_forest_type_all_species)

# create a shared species list
CA_forest_type_all_species
CA_forest_type_all_seedling_species

CA_species <- data.frame(
  species = unique(c(
    CA_forest_type_all_species,
    CA_forest_type_all_seedling_species
  ))
)

```
## dominance
```{r}

CA_tree_count <- CA_TREE_forest_type_species %>%
  group_by(plot_id, INVYR) %>%
  summarise(total_trees_in_plot = n(), .groups = "drop")

CA_tree_species_dominance <- CA_TREE_forest_type_species %>%
  group_by(plot_id, INVYR, species) %>%
  summarise(species_count = n(), .groups = "drop") %>%
  left_join(CA_tree_count, by = c("plot_id", "INVYR")) %>%
  mutate(relative_abundance = species_count / total_trees_in_plot)

CA_tree_species_summary <- CA_tree_species_dominance %>%
  group_by(species) %>%
  summarise(
    n_plots_present  = n_distinct(plot_id),
    avg_dominance    = mean(relative_abundance),
    median_dominance = median(relative_abundance),
    min_dominance    = min(relative_abundance),
    max_dominance    = max(relative_abundance),
    .groups = "drop"
  ) %>%
  mutate(frequency = n_plots_present / n_distinct(CA_TREE_forest_type_species$plot_id)) %>%
  arrange(desc(frequency))

CA_tree_species_summary

# calculate dominance and commonness for seedling species too 

CA_seedling_count <- CA_SEEDLING_forest_type_species %>%
  group_by(plot_id, INVYR) %>%
  summarise(total_seedlings_in_plot = sum(TREECOUNT, na.rm = TRUE),
            .groups = "drop")

CA_seedling_species_dominance <- CA_SEEDLING_forest_type_species %>%
  group_by(plot_id, INVYR, species) %>%
  summarise(species_count = sum(TREECOUNT, na.rm = TRUE), .groups = "drop") %>%
  left_join(CA_seedling_count, by = c("plot_id", "INVYR")) %>%
  filter(total_seedlings_in_plot > 0) %>%   # excludes plots with no seedling count data
  mutate(relative_abundance = species_count / total_seedlings_in_plot)

CA_seedling_species_summary <- CA_seedling_species_dominance %>%
  group_by(species) %>%
  summarise(
    n_plots_present  = n_distinct(plot_id),
    avg_dominance    = mean(relative_abundance),
    median_dominance = median(relative_abundance),
    min_dominance    = min(relative_abundance),
    max_dominance    = max(relative_abundance),
    .groups = "drop"
  ) %>%
  mutate(
    frequency = n_plots_present / n_distinct(CA_seedling_species_dominance$plot_id)  # note: denominator now uses the filtered plot set
  ) %>%
  arrange(desc(frequency))

CA_seedling_species_summary
```

# OR
## whole plot network
```{r}
# create a unique PLOT ID in COND and in PLOT tables 
OR_COND_IDs <- OR_COND %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_"))

OR_PLOT_IDs <- OR_PLOT %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_"))

# create unique plot ids for OR_TREE
OR_TREE_IDs <- OR_TREE %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_"))

# create unique plot ids for OR_SEEDLING
OR_SEEDLING_IDs <- OR_SEEDLING %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_"))

# filter CA_COND_IDs for forest types only, no disturbance filtering
OR_COND_forest_type <- OR_COND_IDs %>%
  filter(FORTYPCD %in% c(185, 182, 184,
201, 202, 221, 222, 224, 225, 226, 
241, 261, 262, 263, 264, 265, 267, 268, 270,
271, 281, 301, 304, 305, 321, 
341, 342, 361, 365, 371,
901, 911, 912, 922, 923, 933, 941, 961))
         
#(371, 201, 221, 222, 224, 225, 261, 922))


# get the unique plot_ids that qualify
OR_forest_type_plot_id <- OR_COND_forest_type %>%
  distinct(plot_id) %>%
  pull(plot_id)

length(OR_forest_type_plot_id)

# filter OR_PLOT_IDs for those plot_ids
OR_PLOT_forest_type <- OR_PLOT_IDs %>%
  filter(plot_id %in% OR_forest_type_plot_id)

# filter OR_TREE_IDs for those plot_ids
OR_TREE_forest_type <- OR_TREE_IDs %>%
  filter(plot_id %in% OR_forest_type_plot_id)

nrow(OR_PLOT_forest_type)
n_distinct(OR_TREE_forest_type$plot_id)

OR_TREE_forest_type_species <- OR_TREE_forest_type %>%
  left_join(
    REF_SPECIES %>%
      mutate(species = paste(GENUS, SPECIES, sep = "_")) %>%
      select(SPCD, species),
    by = "SPCD"
  )

sum(is.na(OR_TREE_forest_type_species$species))
n_distinct(OR_TREE_forest_type_species$species)

# species from your new forest-type-only workflow
OR_forest_type_all_species <- OR_TREE_forest_type_species %>%
  distinct(species) %>%
  pull(species)

length(OR_forest_type_all_species)

# make sure the species coming out of SEEDLING match TREE
# filter OR_SEEDLING_IDs for those plot_ids
OR_SEEDLING_forest_type <- OR_SEEDLING_IDs %>%
  filter(plot_id %in% OR_forest_type_plot_id)

OR_SEEDLING_forest_type_species <- OR_SEEDLING_forest_type %>%
  left_join(
    REF_SPECIES %>%
      mutate(species = paste(GENUS, SPECIES, sep = "_")) %>%
      select(SPCD, species),
    by = "SPCD"
  )

# species list from seedling data (forest-type-only workflow)
OR_forest_type_all_seedling_species <- OR_SEEDLING_forest_type_species %>%
  distinct(species) %>%
  pull(species)

length(OR_forest_type_all_species)
length(OR_forest_type_all_seedling_species)

# species found in both tree and seedling data
intersect(OR_forest_type_all_species, OR_forest_type_all_seedling_species)

# species found only in tree data (mature trees, no seedlings recorded)
setdiff(OR_forest_type_all_species, OR_forest_type_all_seedling_species)

# species found only in seedling data (recruiting, but no mature trees recorded)
setdiff(OR_forest_type_all_seedling_species, OR_forest_type_all_species)

# create a shared species list
OR_forest_type_all_species
OR_forest_type_all_seedling_species

OR_species <- data.frame(
  species = unique(c(
    OR_forest_type_all_species,
    OR_forest_type_all_seedling_species
  ))
)
```

## dominance
```{r}
OR_tree_count <- OR_TREE_forest_type_species %>%
  group_by(plot_id, INVYR) %>%
  summarise(total_trees_in_plot = n(), .groups = "drop")

OR_tree_species_dominance <- OR_TREE_forest_type_species %>%
  group_by(plot_id, INVYR, species) %>%
  summarise(species_count = n(), .groups = "drop") %>%
  left_join(OR_tree_count, by = c("plot_id", "INVYR")) %>%
  mutate(relative_abundance = species_count / total_trees_in_plot)

OR_tree_species_summary <- OR_tree_species_dominance %>%
  group_by(species) %>%
  summarise(
    n_plots_present  = n_distinct(plot_id),
    avg_dominance    = mean(relative_abundance),
    median_dominance = median(relative_abundance),
    min_dominance    = min(relative_abundance),
    max_dominance    = max(relative_abundance),
    .groups = "drop"
  ) %>%
  mutate(frequency = n_plots_present / n_distinct(OR_TREE_forest_type_species$plot_id)) %>%
  arrange(desc(frequency))

OR_tree_species_summary


# calculate dominance and commonness for seedling species too 

OR_seedling_count <- OR_SEEDLING_forest_type_species %>%
  group_by(plot_id, INVYR) %>%
  summarise(
    total_seedlings_in_plot = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop"
  )

OR_seedling_species_dominance <- OR_SEEDLING_forest_type_species %>%
  group_by(plot_id, INVYR, species) %>%
  summarise(species_count = sum(TREECOUNT, na.rm = TRUE), .groups = "drop") %>%
  left_join(OR_seedling_count, by = c("plot_id", "INVYR")) %>%
  filter(total_seedlings_in_plot > 0) %>%   # excludes plots with no seedling count data
  mutate(relative_abundance = species_count / total_seedlings_in_plot)

OR_seedling_species_summary <- OR_seedling_species_dominance %>%
  group_by(species) %>%
  summarise(
    n_plots_present  = n_distinct(plot_id),
    avg_dominance    = mean(relative_abundance),
    median_dominance = median(relative_abundance),
    min_dominance    = min(relative_abundance),
    max_dominance    = max(relative_abundance),
    .groups = "drop"
  ) %>%
  mutate(
    frequency = n_plots_present / n_distinct(OR_seedling_species_dominance$plot_id)  # note: denominator now uses the filtered plot set
  ) %>%
  arrange(desc(frequency))

OR_seedling_species_summary
```

#WA
## whole plot network
```{r}
# create a unique PLOT ID in COND and in PLOT tables 
WA_COND_IDs <- WA_COND %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_"))

WA_PLOT_IDs <- WA_PLOT %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_"))

# create unique plot ids for OR_TREE
WA_TREE_IDs <- WA_TREE %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_"))

# create unique plot ids for OR_SEEDLING
WA_SEEDLING_IDs <- WA_SEEDLING %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_"))

# filter CA_COND_IDs for forest types only, no disturbance filtering
WA_COND_forest_type <- WA_COND_IDs %>%
  filter(FORTYPCD %in% c(185, 182, 184,
201, 202, 221, 222, 224, 225, 226, 
241, 261, 262, 263, 264, 265, 267, 268, 270,
271, 281, 301, 304, 305, 321, 
341, 342, 361, 365, 371,
901, 911, 912, 922, 923, 933, 941, 961))
           
#c(201, 202, 221, 222, 224, 225, 226, 241, 261, 262, 263, 264, 265, 267, 268, 270, 271, 281, 301, 304, 305, 321, 341, 342, 361, 365, 371, 722, 911, 912, 922, 923, 933, 941, 961))
         
         
#(371, 201, 221, 222, 224, 225, 261, 922))

# get the unique plot_ids that qualify
WA_forest_type_plot_id <- WA_COND_forest_type %>%
  distinct(plot_id) %>%
  pull(plot_id)

length(WA_forest_type_plot_id)

# filter CA_PLOT_IDs for those plot_ids
WA_PLOT_forest_type <- WA_PLOT_IDs %>%
  filter(plot_id %in% WA_forest_type_plot_id)

# filter CA_TREE_IDs for those plot_ids
WA_TREE_forest_type <- WA_TREE_IDs %>%
  filter(plot_id %in% WA_forest_type_plot_id)

nrow(WA_PLOT_forest_type)
n_distinct(WA_TREE_forest_type$plot_id)

WA_TREE_forest_type_species <- WA_TREE_forest_type %>%
  left_join(
    REF_SPECIES %>%
      mutate(species = paste(GENUS, SPECIES, sep = "_")) %>%
      select(SPCD, species),
    by = "SPCD"
  )
sum(is.na(WA_TREE_forest_type_species$species))
n_distinct(WA_TREE_forest_type_species$species) # 49 species

# species from your new forest-type-only workflow
WA_forest_type_all_species <- WA_TREE_forest_type_species %>%
  distinct(species) %>%
  pull(species)

length(WA_forest_type_all_species)

# make sure the species coming out of SEEDLING match TREE
# filter CA_SEEDLING_IDs for those plot_ids
WA_SEEDLING_forest_type <- WA_SEEDLING_IDs %>%
  filter(plot_id %in% WA_forest_type_plot_id)

WA_SEEDLING_forest_type_species <- WA_SEEDLING_forest_type %>%
  left_join(
    REF_SPECIES %>%
      mutate(species = paste(GENUS, SPECIES, sep = "_")) %>%
      select(SPCD, species),
    by = "SPCD"
  )

# species list from seedling data (forest-type-only workflow)
WA_forest_type_all_seedling_species <- WA_SEEDLING_forest_type_species %>%
  distinct(species) %>%
  pull(species)

length(WA_forest_type_all_species)
length(WA_forest_type_all_seedling_species)

# species found in both tree and seedling data
intersect(WA_forest_type_all_species, WA_forest_type_all_seedling_species)

# species found only in tree data (mature trees, no seedlings recorded)
setdiff(WA_forest_type_all_species, WA_forest_type_all_seedling_species)

# species found only in seedling data (recruiting, but no mature trees recorded)
setdiff(WA_forest_type_all_seedling_species, WA_forest_type_all_species)

# create a shared species list
WA_forest_type_all_species
WA_forest_type_all_seedling_species

WA_species <- data.frame(
  species = unique(c(
    WA_forest_type_all_species,
    WA_forest_type_all_seedling_species
  ))
)
```

## dominance
```{r}
WA_tree_count <- WA_TREE_forest_type_species %>%
  group_by(plot_id, INVYR) %>%
  summarise(total_trees_in_plot = n(), .groups = "drop")

WA_tree_species_dominance <- WA_TREE_forest_type_species %>%
  group_by(plot_id, INVYR, species) %>%
  summarise(species_count = n(), .groups = "drop") %>%
  left_join(WA_tree_count, by = c("plot_id", "INVYR")) %>%
  mutate(relative_abundance = species_count / total_trees_in_plot)

WA_tree_species_summary <- WA_tree_species_dominance %>%
  group_by(species) %>%
  summarise(
    n_plots_present  = n_distinct(plot_id),
    avg_dominance    = mean(relative_abundance),
    median_dominance = median(relative_abundance),
    min_dominance    = min(relative_abundance),
    max_dominance    = max(relative_abundance),
    .groups = "drop"
  ) %>%
  mutate(frequency = n_plots_present / n_distinct(WA_TREE_forest_type_species$plot_id)) %>%
  arrange(desc(frequency))

WA_tree_species_summary


# calculate dominance and commonness for seedling species too 

WA_seedling_count <- WA_SEEDLING_forest_type_species %>%
  group_by(plot_id, INVYR) %>%
  summarise(
    total_seedlings_in_plot = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop"
  )

WA_seedling_species_dominance <- WA_SEEDLING_forest_type_species %>%
  group_by(plot_id, INVYR, species) %>%
  summarise(species_count = sum(TREECOUNT, na.rm = TRUE), .groups = "drop") %>%
  left_join(WA_seedling_count, by = c("plot_id", "INVYR")) %>%
  filter(total_seedlings_in_plot > 0) %>%   # excludes plots with no actual seedling count data
  mutate(relative_abundance = species_count / total_seedlings_in_plot)

WA_seedling_species_summary <- WA_seedling_species_dominance %>%
  group_by(species) %>%
  summarise(
    n_plots_present  = n_distinct(plot_id),
    avg_dominance    = mean(relative_abundance),
    median_dominance = median(relative_abundance),
    min_dominance    = min(relative_abundance),
    max_dominance    = max(relative_abundance),
    .groups = "drop"
  ) %>%
  mutate(
    frequency = n_plots_present / n_distinct(WA_seedling_species_dominance$plot_id)  # note: denominator now uses the filtered plot set
  ) %>%
  arrange(desc(frequency))

WA_seedling_species_summary
```

# CA, WA, OR species list

```{r}

CA_species_vect <- CA_species$species
OR_species_vect <- OR_species$species
WA_species_vect <- WA_species$species

CA_OR_WA_forest_type_species_lists <- list(
  CA = CA_species_vect,
  OR = OR_species_vect,
  WA = WA_species_vect
)
CA_OR_WA_forest_type_species_overlap <- Reduce(union, CA_OR_WA_forest_type_species_lists)

length(CA_OR_WA_forest_type_species_overlap)

# species found in ALL THREE states
CA_OR_WA_forest_type_species <- Reduce(intersect, CA_OR_WA_forest_type_species_lists)
length(unique(CA_OR_WA_forest_type_species)) 
# 47 species shared in all 3 states


write.csv(CA_OR_WA_forest_type_species_overlap, "updated_species_list_07182026.csv", row.names = FALSE)

# species found in exactly one state only (not shared with either other state)
length(setdiff(CA_species$species, union(OR_species$species, WA_species$species))) # 21 in CA only
length(setdiff(OR_species$species, union(CA_species$species, WA_species$species))) # 11 in OR only
length(setdiff(WA_species$species, union(CA_species$species, OR_species$species))) # 5 in WA only

# Shared by CA and OR, but NOT WA
setdiff(intersect(CA_species$species, OR_species$species), WA_species$species)
# 15 species

# Shared by CA and WA, but NOT OR
setdiff(intersect(CA_species$species, WA_species$species), OR_species$species)
# 1 species

# Shared by OR and WA, but NOT CA
setdiff(intersect(OR_species$species, WA_species$species), CA_species$species)
# 7 species 

# 100 unique species across all states, 40 shared and 37 unique

# Install and load the package if you haven't already
install.packages("ggvenn")
library(ggvenn)

# Combine your separate state lists into a single named list
venn_data <- list(
  California = CA_species$species,
  Oregon     = OR_species$species,
  Washington = WA_species$species
)

# Plot the diagram
ggvenn(venn_data, fill_color = c("#3A86FF", "#FF006E", "#FFBE0B"))



CA_OR_WA_species_lists <- list(
  CA = CA_all_species,
  OR = OR_all_species,
  WA = WA_all_species
)

# species found in ALL THREE states
CA_OR_WA_species <- Reduce(intersect, CA_OR_WA_species_lists)
length(CA_OR_WA_species)

# total unique species across all three states combined
CA_OR_WA_species_overlap <- Reduce(union, CA_OR_WA_species_lists)
length(CA_OR_WA_species_overlap)

# species found in exactly one state only (not shared with either other state)
length(setdiff(CA_all_species, union(OR_all_species, WA_all_species)))
length(setdiff(OR_all_species, union(CA_all_species, WA_all_species)))
length(setdiff(WA_all_species, union(CA_all_species, OR_all_species)))

# 78 unique species across all states, 48 shared and 30 unique


# are burned plot species much different than control plot species
# CA
CA_species_burned  <- CA_TREE_species_filtered %>% filter(condition == "burned")  %>% distinct(species) %>% pull(species)
CA_species_control <- CA_TREE_species_filtered %>% filter(condition == "control") %>% distinct(species) %>% pull(species)

length(CA_species_burned) # 49 species
length(CA_species_control) # 60 species
length(intersect(CA_species_burned, CA_species_control)) # 44 species shared
length(setdiff(CA_species_burned, CA_species_control)) # 5 in burned only
length(setdiff(CA_species_control, CA_species_burned)) # 16 in control-only

# OR
OR_species_burned  <- OR_TREE_species_filtered %>% filter(condition == "burned")  %>% distinct(species) %>% pull(species)
OR_species_control <- OR_TREE_species_filtered %>% filter(condition == "control") %>% distinct(species) %>% pull(species)

length(OR_species_burned) # 44 species
length(OR_species_control) # 51 species
length(intersect(OR_species_burned, OR_species_control)) # 42 species shared
length(setdiff(OR_species_burned, OR_species_control)) # 2 in burned only
length(setdiff(OR_species_control, OR_species_burned)) # 9 in control only

# WA
WA_species_burned  <- WA_TREE_species_filtered %>% filter(condition == "burned")  %>% distinct(species) %>% pull(species)
WA_species_control <- WA_TREE_species_filtered %>% filter(condition == "control") %>% distinct(species) %>% pull(species)

length(WA_species_burned) # 25 species
length(WA_species_control) # 31 species
length(intersect(WA_species_burned, WA_species_control)) # 23 species shared
length(setdiff(WA_species_burned, WA_species_control)) # 2 in burned only
length(setdiff(WA_species_control, WA_species_burned)) # 8 in control only 
```

#ID
## whole plot network
```{r}
ID_COND_IDs <- ID_COND %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )
ID_PLOT_IDs <- ID_PLOT %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )
ID_TREE_IDs <- ID_TREE %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )
# create unique plot ids for CA_SEEDLING
ID_SEEDLING_IDs <- ID_SEEDLING %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

# filter CA_COND_IDs for forest types only, no disturbance filtering
ID_COND_forest_type <- ID_COND_IDs %>%
  filter(FORTYPCD %in% c(185, 182, 184,
201, 202, 221, 222, 224, 225, 226, 
241, 261, 262, 263, 264, 265, 267, 268, 270,
271, 281, 301, 304, 305, 321, 
341, 342, 361, 365, 371,
901, 911, 912, 922, 923, 933, 941, 961))

#c(201, 202, 221, 222, 224, 225, 226, 241, 261, 262, 263, 264, 265, 267, 268, 270, 271, 281, 301, 304, 305, 321, 341, 342, 361, 365, 371, 722, 911, 912, 922, 923, 933, 941, 961))
         
#(371, 201, 221, 222, 224, 225, 261, 922))

# get the unique plot_ids that qualify
ID_forest_type_plot_id <- ID_COND_forest_type %>%
  distinct(plot_id) %>%
  pull(plot_id)

length(ID_forest_type_plot_id)

# filter CA_PLOT_IDs for those plot_ids
ID_PLOT_forest_type <- ID_PLOT_IDs %>%
  filter(plot_id %in% ID_forest_type_plot_id)

# filter CA_TREE_IDs for those plot_ids
ID_TREE_forest_type <- ID_TREE_IDs %>%
  filter(plot_id %in% ID_forest_type_plot_id)

nrow(ID_PLOT_forest_type)
n_distinct(ID_TREE_forest_type$plot_id)

ID_TREE_forest_type_species <- ID_TREE_forest_type %>%
  left_join(
    REF_SPECIES %>%
      mutate(species = paste(GENUS, SPECIES, sep = "_")) %>%
      select(SPCD, species),
    by = "SPCD"
  )
sum(is.na(ID_TREE_forest_type_species$species))
n_distinct(ID_TREE_forest_type_species$species)

# how different are the species from this workflow and the previous control plot filtering by climate?


# species from your new forest-type-only workflow
ID_forest_type_all_species <- ID_TREE_forest_type_species %>%
  distinct(species) %>%
  pull(species)

# make sure the species coming out of SEEDLING match TREE
# filter CA_SEEDLING_IDs for those plot_ids
ID_SEEDLING_forest_type <- ID_SEEDLING_IDs %>%
  filter(plot_id %in% ID_forest_type_plot_id)

ID_SEEDLING_forest_type_species <- ID_SEEDLING_forest_type %>%
  left_join(
    REF_SPECIES %>%
      mutate(species = paste(GENUS, SPECIES, sep = "_")) %>%
      select(SPCD, species),
    by = "SPCD"
  )

# species list from seedling data (forest-type-only workflow)
ID_forest_type_all_seedling_species <- ID_SEEDLING_forest_type_species %>%
  distinct(species) %>%
  pull(species)

length(ID_forest_type_all_species)
length(ID_forest_type_all_seedling_species)

# species found in both tree and seedling data
intersect(ID_forest_type_all_species, ID_forest_type_all_seedling_species)

# species found only in tree data (mature trees, no seedlings recorded)
setdiff(ID_forest_type_all_species, ID_forest_type_all_seedling_species)

# species found only in seedling data (recruiting, but no mature trees recorded)
setdiff(ID_forest_type_all_seedling_species, ID_forest_type_all_species)

# create a shared species list
ID_forest_type_all_species
ID_forest_type_all_seedling_species

ID_species <- data.frame(
  species = unique(c(
    ID_forest_type_all_species,
    ID_forest_type_all_seedling_species
  ))
)

```

## dominance
```{r}
ID_tree_count <- ID_TREE_forest_type_species %>%
  group_by(plot_id, INVYR) %>%
  summarise(total_trees_in_plot = n(), .groups = "drop")

ID_tree_species_dominance <- ID_TREE_forest_type_species %>%
  group_by(plot_id, INVYR, species) %>%
  summarise(species_count = n(), .groups = "drop") %>%
  left_join(ID_tree_count, by = c("plot_id", "INVYR")) %>%
  mutate(relative_abundance = species_count / total_trees_in_plot)

ID_tree_species_summary <- ID_tree_species_dominance %>%
  group_by(species) %>%
  summarise(
    n_plots_present  = n_distinct(plot_id),
    avg_dominance    = mean(relative_abundance),
    median_dominance = median(relative_abundance),
    min_dominance    = min(relative_abundance),
    max_dominance    = max(relative_abundance),
    .groups = "drop"
  ) %>%
  mutate(frequency = n_plots_present / n_distinct(ID_TREE_forest_type_species$plot_id)) %>%
  arrange(desc(frequency))

ID_tree_species_summary

ID_seedling_count <- ID_SEEDLING_forest_type_species %>%
  group_by(plot_id, INVYR) %>%
  summarise(
    total_seedlings_in_plot = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop"
  )

ID_seedling_species_dominance <- ID_SEEDLING_forest_type_species %>%
  group_by(plot_id, INVYR, species) %>%
  summarise(species_count = sum(TREECOUNT, na.rm = TRUE), .groups = "drop") %>%
  left_join(ID_seedling_count, by = c("plot_id", "INVYR")) %>%
  filter(total_seedlings_in_plot > 0) %>%   # excludes plots with no actual seedling count data
  mutate(relative_abundance = species_count / total_seedlings_in_plot)

ID_seedling_species_summary <- ID_seedling_species_dominance %>%
  group_by(species) %>%
  summarise(
    n_plots_present  = n_distinct(plot_id),
    avg_dominance    = mean(relative_abundance),
    median_dominance = median(relative_abundance),
    min_dominance    = min(relative_abundance),
    max_dominance    = max(relative_abundance),
    .groups = "drop"
  ) %>%
  mutate(
    frequency = n_plots_present / n_distinct(ID_seedling_species_dominance$plot_id)  # note: denominator now uses the filtered plot set
  ) %>%
  arrange(desc(frequency))

ID_seedling_species_summary
```


#MT
## whole plot network
```{r}
MT_COND_IDs <- MT_COND %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )
MT_PLOT_IDs <- MT_PLOT %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )
# create unique plot ids for CA_TREE
MT_TREE_IDs <- MT_TREE %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

# create unique plot ids for CA_SEEDLING
MT_SEEDLING_IDs <- MT_SEEDLING %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )
# filter CA_COND_IDs for forest types only, no disturbance filtering
MT_COND_forest_type <- MT_COND_IDs %>%
  filter(FORTYPCD %in% c(185, 182, 184,
201, 202, 221, 222, 224, 225, 226, 
241, 261, 262, 263, 264, 265, 267, 268, 270,
271, 281, 301, 304, 305, 321, 
341, 342, 361, 365, 371,
901, 911, 912, 922, 923, 933, 941, 961))
           
#c(201, 202, 221, 222, 224, 225, 226, 241, 261, 262, 263, 264, 265, 267, 268, 270, 271, 281, 301, 304, 305, 321, 341, 342, 361, 365, 371, 722, 911, 912, 922, 923, 933, 941, 961))
         
#(371, 201, 221, 222, 224, 225, 261, 922))

# get the unique plot_ids that qualify
MT_forest_type_plot_id <- MT_COND_forest_type %>%
  distinct(plot_id) %>%
  pull(plot_id)

length(MT_forest_type_plot_id)

# filter CA_PLOT_IDs for those plot_ids
MT_PLOT_forest_type <- MT_PLOT_IDs %>%
  filter(plot_id %in% MT_forest_type_plot_id)

# filter CA_TREE_IDs for those plot_ids
MT_TREE_forest_type <- MT_TREE_IDs %>%
  filter(plot_id %in% MT_forest_type_plot_id)

nrow(MT_PLOT_forest_type)
n_distinct(MT_TREE_forest_type$plot_id)

MT_TREE_forest_type_species <- MT_TREE_forest_type %>%
  left_join(
    REF_SPECIES %>%
      mutate(species = paste(GENUS, SPECIES, sep = "_")) %>%
      select(SPCD, species),
    by = "SPCD"
  )
sum(is.na(MT_TREE_forest_type_species$species))
n_distinct(MT_TREE_forest_type_species$species)



# species from your new forest-type-only workflow
MT_forest_type_all_species <- MT_TREE_forest_type_species %>%
  distinct(species) %>%
  pull(species)

# make sure the species coming out of SEEDLING match TREE
# filter MT_SEEDLING_IDs for those plot_ids
MT_SEEDLING_forest_type <- MT_SEEDLING_IDs %>%
  filter(plot_id %in% MT_forest_type_plot_id)

MT_SEEDLING_forest_type_species <- MT_SEEDLING_forest_type %>%
  left_join(
    REF_SPECIES %>%
      mutate(species = paste(GENUS, SPECIES, sep = "_")) %>%
      select(SPCD, species),
    by = "SPCD"
  )

# species list from seedling data (forest-type-only workflow)
MT_forest_type_all_seedling_species <- MT_SEEDLING_forest_type_species %>%
  distinct(species) %>%
  pull(species)

length(MT_forest_type_all_species)
length(MT_forest_type_all_seedling_species)

# species found in both tree and seedling data
intersect(MT_forest_type_all_species, MT_forest_type_all_seedling_species)

# species found only in tree data (mature trees, no seedlings recorded)
setdiff(MT_forest_type_all_species, MT_forest_type_all_seedling_species)

# species found only in seedling data (recruiting, but no mature trees recorded)
setdiff(MT_forest_type_all_seedling_species, MT_forest_type_all_species)

# create a shared species list
MT_forest_type_all_species
MT_forest_type_all_seedling_species

MT_species <- data.frame(
  species = unique(c(
    MT_forest_type_all_species,
    MT_forest_type_all_seedling_species
  ))
)
```

##dominance
```{r}

MT_tree_count <- MT_TREE_forest_type_species %>%
  group_by(plot_id, INVYR) %>%
  summarise(total_trees_in_plot = n(), .groups = "drop")

MT_tree_species_dominance <- MT_TREE_forest_type_species %>%
  group_by(plot_id, INVYR, species) %>%
  summarise(species_count = n(), .groups = "drop") %>%
  left_join(MT_tree_count, by = c("plot_id", "INVYR")) %>%
  mutate(relative_abundance = species_count / total_trees_in_plot)

MT_tree_species_summary <- MT_tree_species_dominance %>%
  group_by(species) %>%
  summarise(
    n_plots_present  = n_distinct(plot_id),
    avg_dominance    = mean(relative_abundance),
    median_dominance = median(relative_abundance),
    min_dominance    = min(relative_abundance),
    max_dominance    = max(relative_abundance),
    .groups = "drop"
  ) %>%
  mutate(frequency = n_plots_present / n_distinct(MT_TREE_forest_type_species$plot_id)) %>%
  arrange(desc(frequency))

MT_tree_species_summary


# calculate dominance and commonness for seedling species too 

MT_seedling_count <- MT_SEEDLING_forest_type_species %>%
  group_by(plot_id, INVYR) %>%
  summarise(
    total_seedlings_in_plot = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop"
  )

MT_seedling_species_dominance <- MT_SEEDLING_forest_type_species %>%
  group_by(plot_id, INVYR, species) %>%
  summarise(species_count = sum(TREECOUNT, na.rm = TRUE), .groups = "drop") %>%
  left_join(MT_seedling_count, by = c("plot_id", "INVYR")) %>%
  filter(total_seedlings_in_plot > 0) %>%   # excludes plots with no actual seedling count data
  mutate(relative_abundance = species_count / total_seedlings_in_plot)

MT_seedling_species_summary <- MT_seedling_species_dominance %>%
  group_by(species) %>%
  summarise(
    n_plots_present  = n_distinct(plot_id),
    avg_dominance    = mean(relative_abundance),
    median_dominance = median(relative_abundance),
    min_dominance    = min(relative_abundance),
    max_dominance    = max(relative_abundance),
    .groups = "drop"
  ) %>%
  mutate(
    frequency = n_plots_present / n_distinct(MT_seedling_species_dominance$plot_id)  # note: denominator now uses the filtered plot set
  ) %>%
  arrange(desc(frequency))

MT_seedling_species_summary
```

# NV
## whole plot network
```{r}
# NV
## whole plot network

# create a unique PLOT ID in COND and in PLOT tables 
NV_COND_IDs <- NV_COND %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

NV_PLOT_IDs <- NV_PLOT %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

# create unique plot ids for NV_TREE
NV_TREE_IDs <- NV_TREE %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

# create unique plot ids for NV_SEEDLING
NV_SEEDLING_IDs <- NV_SEEDLING %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

# filter NV_COND_IDs for forest types only, no disturbance filtering
NV_COND_forest_type <- NV_COND_IDs %>%
  filter(FORTYPCD %in% c(185, 182, 184,
201, 202, 221, 222, 224, 225, 226, 
241, 261, 262, 263, 264, 265, 267, 268, 270,
271, 281, 301, 304, 305, 321, 
341, 342, 361, 365, 371,
901, 911, 912, 922, 923, 933, 941, 961))

#c(201, 202, 221, 222, 224, 225, 226, 241, 261, 262,263, 264, 265, 267, 268, 270, 271, 281, 301, 304, 305, 321, 341, 342, 361, 365, 371, 722, 911, 912,922, 923, 933, 941, 961))

# get the unique plot_ids that qualify
NV_forest_type_plot_id <- NV_COND_forest_type %>%
  distinct(plot_id) %>%
  pull(plot_id)

length(NV_forest_type_plot_id)

# filter NV_PLOT_IDs for those plot_ids
NV_PLOT_forest_type <- NV_PLOT_IDs %>%
  filter(plot_id %in% NV_forest_type_plot_id)

# filter NV_TREE_IDs for those plot_ids
NV_TREE_forest_type <- NV_TREE_IDs %>%
  filter(plot_id %in% NV_forest_type_plot_id)

nrow(NV_PLOT_forest_type)
n_distinct(NV_TREE_forest_type$plot_id)

NV_TREE_forest_type_species <- NV_TREE_forest_type %>%
  left_join(
    REF_SPECIES %>%
      mutate(species = paste(GENUS, SPECIES, sep = "_")) %>%
      select(SPCD, species),
    by = "SPCD"
  )

sum(is.na(NV_TREE_forest_type_species$species))
n_distinct(NV_TREE_forest_type_species$species)


# species from your new forest-type-only workflow
NV_forest_type_all_species <- NV_TREE_forest_type_species %>%
  distinct(species) %>%
  pull(species)

length(NV_forest_type_all_species)


# make sure the species coming out of SEEDLING match TREE
# filter NV_SEEDLING_IDs for those plot_ids
NV_SEEDLING_forest_type <- NV_SEEDLING_IDs %>%
  filter(plot_id %in% NV_forest_type_plot_id)

NV_SEEDLING_forest_type_species <- NV_SEEDLING_forest_type %>%
  left_join(
    REF_SPECIES %>%
      mutate(species = paste(GENUS, SPECIES, sep = "_")) %>%
      select(SPCD, species),
    by = "SPCD"
  )


# species list from seedling data (forest-type-only workflow)
NV_forest_type_all_seedling_species <- NV_SEEDLING_forest_type_species %>%
  distinct(species) %>%
  pull(species)

length(NV_forest_type_all_species)
length(NV_forest_type_all_seedling_species)


# species found in both tree and seedling data
intersect(NV_forest_type_all_species, NV_forest_type_all_seedling_species)

# species found only in tree data (mature trees, no seedlings recorded)
setdiff(NV_forest_type_all_species, NV_forest_type_all_seedling_species)

# species found only in seedling data (recruiting, but no mature trees recorded)
setdiff(NV_forest_type_all_seedling_species, NV_forest_type_all_species)


# create a shared species list
NV_forest_type_all_species
NV_forest_type_all_seedling_species

NV_species <- data.frame(
  species = unique(c(
    NV_forest_type_all_species,
    NV_forest_type_all_seedling_species
  ))
)
```
## dominance
```{r}
NV_tree_count <- NV_TREE_forest_type_species %>%
  group_by(plot_id, INVYR) %>%
  summarise(total_trees_in_plot = n(), .groups = "drop")

NV_tree_species_dominance <- NV_TREE_forest_type_species %>%
  group_by(plot_id, INVYR, species) %>%
  summarise(species_count = n(), .groups = "drop") %>%
  left_join(NV_tree_count, by = c("plot_id", "INVYR")) %>%
  mutate(relative_abundance = species_count / total_trees_in_plot)

NV_tree_species_summary <- NV_tree_species_dominance %>%
  group_by(species) %>%
  summarise(
    n_plots_present  = n_distinct(plot_id),
    avg_dominance    = mean(relative_abundance),
    median_dominance = median(relative_abundance),
    min_dominance    = min(relative_abundance),
    max_dominance    = max(relative_abundance),
    .groups = "drop"
  ) %>%
  mutate(frequency = n_plots_present / n_distinct(NV_TREE_forest_type_species$plot_id)) %>%
  arrange(desc(frequency))

NV_tree_species_summary


# calculate dominance and commonness for seedling species too 

NV_seedling_count <- NV_SEEDLING_forest_type_species %>%
  group_by(plot_id, INVYR) %>%
  summarise(
    total_seedlings_in_plot = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop"
  )

NV_seedling_species_dominance <- NV_SEEDLING_forest_type_species %>%
  group_by(plot_id, INVYR, species) %>%
  summarise(species_count = sum(TREECOUNT, na.rm = TRUE), .groups = "drop") %>%
  left_join(NV_seedling_count, by = c("plot_id", "INVYR")) %>%
  filter(total_seedlings_in_plot > 0) %>%   # excludes plots with no actual seedling count data
  mutate(relative_abundance = species_count / total_seedlings_in_plot)

NV_seedling_species_summary <- NV_seedling_species_dominance %>%
  group_by(species) %>%
  summarise(
    n_plots_present  = n_distinct(plot_id),
    avg_dominance    = mean(relative_abundance),
    median_dominance = median(relative_abundance),
    min_dominance    = min(relative_abundance),
    max_dominance    = max(relative_abundance),
    .groups = "drop"
  ) %>%
  mutate(
    frequency = n_plots_present / n_distinct(NV_seedling_species_dominance$plot_id)  # note: denominator now uses the filtered plot set
  ) %>%
  arrange(desc(frequency))

NV_seedling_species_summary

```

# UT
## whole plot network
```{r}
# UT
## whole plot network

# create a unique PLOT ID in COND and in PLOT tables 
UT_COND_IDs <- UT_COND %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

UT_PLOT_IDs <- UT_PLOT %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

# create unique plot ids for UT_TREE
UT_TREE_IDs <- UT_TREE %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

# create unique plot ids for UT_SEEDLING
UT_SEEDLING_IDs <- UT_SEEDLING %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

# filter UT_COND_IDs for forest types only, no disturbance filtering
UT_COND_forest_type <- UT_COND_IDs %>%
  filter(FORTYPCD %in% c(185, 182, 184,
201, 202, 221, 222, 224, 225, 226, 
241, 261, 262, 263, 264, 265, 267, 268, 270,
271, 281, 301, 304, 305, 321, 
341, 342, 361, 365, 371,
901, 911, 912, 922, 923, 933, 941, 961))
           
#c(201, 202, 221, 222, 224, 225, 226, 241, 261, 262, 263, 264, 265, 267, 268, 270, 271, 281, 301, 304,305, 321, 341, 342, 361, 365, 371, 722, 911, 912,922, 923, 933, 941, 961))

# get the unique plot_ids that qualify
UT_forest_type_plot_id <- UT_COND_forest_type %>%
  distinct(plot_id) %>%
  pull(plot_id)

length(UT_forest_type_plot_id)

# filter UT_PLOT_IDs for those plot_ids
UT_PLOT_forest_type <- UT_PLOT_IDs %>%
  filter(plot_id %in% UT_forest_type_plot_id)

# filter UT_TREE_IDs for those plot_ids
UT_TREE_forest_type <- UT_TREE_IDs %>%
  filter(plot_id %in% UT_forest_type_plot_id)

nrow(UT_PLOT_forest_type)
n_distinct(UT_TREE_forest_type$plot_id)

UT_TREE_forest_type_species <- UT_TREE_forest_type %>%
  left_join(
    REF_SPECIES %>%
      mutate(species = paste(GENUS, SPECIES, sep = "_")) %>%
      select(SPCD, species),
    by = "SPCD"
  )

sum(is.na(UT_TREE_forest_type_species$species))
n_distinct(UT_TREE_forest_type_species$species)


# species from your new forest-type-only workflow
UT_forest_type_all_species <- UT_TREE_forest_type_species %>%
  distinct(species) %>%
  pull(species)

length(UT_forest_type_all_species)


# make sure the species coming out of SEEDLING match TREE
# filter UT_SEEDLING_IDs for those plot_ids
UT_SEEDLING_forest_type <- UT_SEEDLING_IDs %>%
  filter(plot_id %in% UT_forest_type_plot_id)

UT_SEEDLING_forest_type_species <- UT_SEEDLING_forest_type %>%
  left_join(
    REF_SPECIES %>%
      mutate(species = paste(GENUS, SPECIES, sep = "_")) %>%
      select(SPCD, species),
    by = "SPCD"
  )


# species list from seedling data (forest-type-only workflow)
UT_forest_type_all_seedling_species <- UT_SEEDLING_forest_type_species %>%
  distinct(species) %>%
  pull(species)

length(UT_forest_type_all_species)
length(UT_forest_type_all_seedling_species)


# species found in both tree and seedling data
intersect(UT_forest_type_all_species, UT_forest_type_all_seedling_species)

# species found only in tree data (mature trees, no seedlings recorded)
setdiff(UT_forest_type_all_species, UT_forest_type_all_seedling_species)

# species found only in seedling data (recruiting, but no mature trees recorded)
setdiff(UT_forest_type_all_seedling_species, UT_forest_type_all_species)


# create a shared species list
UT_forest_type_all_species
UT_forest_type_all_seedling_species

UT_species <- data.frame(
  species = unique(c(
    UT_forest_type_all_species,
    UT_forest_type_all_seedling_species
  ))
)
```

## dominance
```{r}
UT_tree_count <- UT_TREE_forest_type_species %>%
  group_by(plot_id, INVYR) %>%
  summarise(total_trees_in_plot = n(), .groups = "drop")

UT_tree_species_dominance <- UT_TREE_forest_type_species %>%
  group_by(plot_id, INVYR, species) %>%
  summarise(species_count = n(), .groups = "drop") %>%
  left_join(UT_tree_count, by = c("plot_id", "INVYR")) %>%
  mutate(relative_abundance = species_count / total_trees_in_plot)

UT_tree_species_summary <- UT_tree_species_dominance %>%
  group_by(species) %>%
  summarise(
    n_plots_present  = n_distinct(plot_id),
    avg_dominance    = mean(relative_abundance),
    median_dominance = median(relative_abundance),
    min_dominance    = min(relative_abundance),
    max_dominance    = max(relative_abundance),
    .groups = "drop"
  ) %>%
  mutate(frequency = n_plots_present / n_distinct(UT_TREE_forest_type_species$plot_id)) %>%
  arrange(desc(frequency))

UT_tree_species_summary


# calculate dominance and commonness for seedling species too 

UT_seedling_count <- UT_SEEDLING_forest_type_species %>%
  group_by(plot_id, INVYR) %>%
  summarise(
    total_seedlings_in_plot = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop"
  )

UT_seedling_species_dominance <- UT_SEEDLING_forest_type_species %>%
  group_by(plot_id, INVYR, species) %>%
  summarise(species_count = sum(TREECOUNT, na.rm = TRUE), .groups = "drop") %>%
  left_join(UT_seedling_count, by = c("plot_id", "INVYR")) %>%
  filter(total_seedlings_in_plot > 0) %>%
  mutate(relative_abundance = species_count / total_seedlings_in_plot)

UT_seedling_species_summary <- UT_seedling_species_dominance %>%
  group_by(species) %>%
  summarise(
    n_plots_present  = n_distinct(plot_id),
    avg_dominance    = mean(relative_abundance),
    median_dominance = median(relative_abundance),
    min_dominance    = min(relative_abundance),
    max_dominance    = max(relative_abundance),
    .groups = "drop"
  ) %>%
  mutate(
    frequency = n_plots_present / n_distinct(UT_seedling_species_dominance$plot_id)
  ) %>%
  arrange(desc(frequency))

UT_seedling_species_summary
```

# CO
## whole plot network
```{r}
# create a unique PLOT ID in COND and in PLOT tables 
CO_COND_IDs <- CO_COND %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

CO_PLOT_IDs <- CO_PLOT %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

# create unique plot ids for CO_TREE
CO_TREE_IDs <- CO_TREE %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

# create unique plot ids for CO_SEEDLING
CO_SEEDLING_IDs <- CO_SEEDLING %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

# filter CO_COND_IDs for forest types only, no disturbance filtering
CO_COND_forest_type <- CO_COND_IDs %>%
  filter(FORTYPCD %in% c(185, 182, 184,
201, 202, 221, 222, 224, 225, 226, 
241, 261, 262, 263, 264, 265, 267, 268, 270,
271, 281, 301, 304, 305, 321, 
341, 342, 361, 365, 371,
901, 911, 912, 922, 923, 933, 941, 961))
           
#c(201, 202, 221, 222, 224, 225, 226, 241, 261, 262, 263, 264, 265, 267, 268, 270, 271, 281, 301, 304, 305, 321, 341, 342, 361, 365, 371, 722, 911, 912,922, 923, 933, 941, 961))

# get the unique plot_ids that qualify
CO_forest_type_plot_id <- CO_COND_forest_type %>%
  distinct(plot_id) %>%
  pull(plot_id)

length(CO_forest_type_plot_id)

# filter CO_PLOT_IDs for those plot_ids
CO_PLOT_forest_type <- CO_PLOT_IDs %>%
  filter(plot_id %in% CO_forest_type_plot_id)

# filter CO_TREE_IDs for those plot_ids
CO_TREE_forest_type <- CO_TREE_IDs %>%
  filter(plot_id %in% CO_forest_type_plot_id)

nrow(CO_PLOT_forest_type)
n_distinct(CO_TREE_forest_type$plot_id)

CO_TREE_forest_type_species <- CO_TREE_forest_type %>%
  left_join(
    REF_SPECIES %>%
      mutate(species = paste(GENUS, SPECIES, sep = "_")) %>%
      select(SPCD, species),
    by = "SPCD"
  )

sum(is.na(CO_TREE_forest_type_species$species))
n_distinct(CO_TREE_forest_type_species$species)


# species from your new forest-type-only workflow
CO_forest_type_all_species <- CO_TREE_forest_type_species %>%
  distinct(species) %>%
  pull(species)

length(CO_forest_type_all_species)


# make sure the species coming out of SEEDLING match TREE
# filter CO_SEEDLING_IDs for those plot_ids
CO_SEEDLING_forest_type <- CO_SEEDLING_IDs %>%
  filter(plot_id %in% CO_forest_type_plot_id)

CO_SEEDLING_forest_type_species <- CO_SEEDLING_forest_type %>%
  left_join(
    REF_SPECIES %>%
      mutate(species = paste(GENUS, SPECIES, sep = "_")) %>%
      select(SPCD, species),
    by = "SPCD"
  )


# species list from seedling data (forest-type-only workflow)
CO_forest_type_all_seedling_species <- CO_SEEDLING_forest_type_species %>%
  distinct(species) %>%
  pull(species)

length(CO_forest_type_all_species)
length(CO_forest_type_all_seedling_species)


# species found in both tree and seedling data
intersect(CO_forest_type_all_species, CO_forest_type_all_seedling_species)

# species found only in tree data (mature trees, no seedlings recorded)
setdiff(CO_forest_type_all_species, CO_forest_type_all_seedling_species)

# species found only in seedling data (recruiting, but no mature trees recorded)
setdiff(CO_forest_type_all_seedling_species, CO_forest_type_all_species)


# create a shared species list
CO_forest_type_all_species
CO_forest_type_all_seedling_species

CO_species <- data.frame(
  species = unique(c(
    CO_forest_type_all_species,
    CO_forest_type_all_seedling_species
  ))
)
```
## dominance 

```{r}

CO_tree_count <- CO_TREE_forest_type_species %>%
  group_by(plot_id, INVYR) %>%
  summarise(total_trees_in_plot = n(), .groups = "drop")

CO_tree_species_dominance <- CO_TREE_forest_type_species %>%
  group_by(plot_id, INVYR, species) %>%
  summarise(species_count = n(), .groups = "drop") %>%
  left_join(CO_tree_count, by = c("plot_id", "INVYR")) %>%
  mutate(relative_abundance = species_count / total_trees_in_plot)

CO_tree_species_summary <- CO_tree_species_dominance %>%
  group_by(species) %>%
  summarise(
    n_plots_present  = n_distinct(plot_id),
    avg_dominance    = mean(relative_abundance),
    median_dominance = median(relative_abundance),
    min_dominance    = min(relative_abundance),
    max_dominance    = max(relative_abundance),
    .groups = "drop"
  ) %>%
  mutate(frequency = n_plots_present / n_distinct(CO_TREE_forest_type_species$plot_id)) %>%
  arrange(desc(frequency))

CO_tree_species_summary


# calculate dominance and commonness for seedling species too 

CO_seedling_count <- CO_SEEDLING_forest_type_species %>%
  group_by(plot_id, INVYR) %>%
  summarise(
    total_seedlings_in_plot = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop"
  )

CO_seedling_species_dominance <- CO_SEEDLING_forest_type_species %>%
  group_by(plot_id, INVYR, species) %>%
  summarise(species_count = sum(TREECOUNT, na.rm = TRUE), .groups = "drop") %>%
  left_join(CO_seedling_count, by = c("plot_id", "INVYR")) %>%
  filter(total_seedlings_in_plot > 0) %>%
  mutate(relative_abundance = species_count / total_seedlings_in_plot)

CO_seedling_species_summary <- CO_seedling_species_dominance %>%
  group_by(species) %>%
  summarise(
    n_plots_present  = n_distinct(plot_id),
    avg_dominance    = mean(relative_abundance),
    median_dominance = median(relative_abundance),
    min_dominance    = min(relative_abundance),
    max_dominance    = max(relative_abundance),
    .groups = "drop"
  ) %>%
  mutate(
    frequency = n_plots_present / n_distinct(CO_seedling_species_dominance$plot_id)
  ) %>%
  arrange(desc(frequency))

CO_seedling_species_summary

```


# WY
## whole plot network
```{r}

WY_COND_IDs <- WY_COND %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

WY_PLOT_IDs <- WY_PLOT %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

WY_TREE_IDs <- WY_TREE %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

WY_SEEDLING_IDs <- WY_SEEDLING %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

WY_COND_forest_type <- WY_COND_IDs %>%
  filter(FORTYPCD %in% c(185, 182, 184,
201, 202, 221, 222, 224, 225, 226, 
241, 261, 262, 263, 264, 265, 267, 268, 270,
271, 281, 301, 304, 305, 321, 
341, 342, 361, 365, 371,
901, 911, 912, 922, 923, 933, 941, 961))

WY_forest_type_plot_id <- WY_COND_forest_type %>%
  distinct(plot_id) %>%
  pull(plot_id)

length(WY_forest_type_plot_id)

WY_PLOT_forest_type <- WY_PLOT_IDs %>%
  filter(plot_id %in% WY_forest_type_plot_id)

WY_TREE_forest_type <- WY_TREE_IDs %>%
  filter(plot_id %in% WY_forest_type_plot_id)

nrow(WY_PLOT_forest_type)
n_distinct(WY_TREE_forest_type$plot_id)

WY_TREE_forest_type_species <- WY_TREE_forest_type %>%
  left_join(
    REF_SPECIES %>%
      mutate(species = paste(GENUS, SPECIES, sep = "_")) %>%
      select(SPCD, species),
    by = "SPCD"
  )

sum(is.na(WY_TREE_forest_type_species$species))
n_distinct(WY_TREE_forest_type_species$species)

WY_forest_type_all_species <- WY_TREE_forest_type_species %>%
  distinct(species) %>%
  pull(species)


WY_SEEDLING_forest_type <- WY_SEEDLING_IDs %>%
  filter(plot_id %in% WY_forest_type_plot_id)

WY_SEEDLING_forest_type_species <- WY_SEEDLING_forest_type %>%
  left_join(
    REF_SPECIES %>%
      mutate(species = paste(GENUS, SPECIES, sep = "_")) %>%
      select(SPCD, species),
    by = "SPCD"
  )

WY_forest_type_all_seedling_species <- WY_SEEDLING_forest_type_species %>%
  distinct(species) %>%
  pull(species)

length(WY_forest_type_all_species)
length(WY_forest_type_all_seedling_species)

intersect(WY_forest_type_all_species, WY_forest_type_all_seedling_species)

setdiff(WY_forest_type_all_species, WY_forest_type_all_seedling_species)

setdiff(WY_forest_type_all_seedling_species, WY_forest_type_all_species)

WY_species <- data.frame(
  species = unique(c(
    WY_forest_type_all_species,
    WY_forest_type_all_seedling_species
  ))
)
```

## dominance  
```{r}

WY_tree_count <- WY_TREE_forest_type_species %>%
  group_by(plot_id, INVYR) %>%
  summarise(total_trees_in_plot = n(), .groups = "drop")

WY_tree_species_dominance <- WY_TREE_forest_type_species %>%
  group_by(plot_id, INVYR, species) %>%
  summarise(species_count = n(), .groups = "drop") %>%
  left_join(WY_tree_count, by = c("plot_id", "INVYR")) %>%
  mutate(relative_abundance = species_count / total_trees_in_plot)

WY_tree_species_summary <- WY_tree_species_dominance %>%
  group_by(species) %>%
  summarise(
    n_plots_present = n_distinct(plot_id),
    avg_dominance = mean(relative_abundance),
    median_dominance = median(relative_abundance),
    min_dominance = min(relative_abundance),
    max_dominance = max(relative_abundance),
    .groups = "drop"
  ) %>%
  mutate(frequency = n_plots_present / n_distinct(WY_TREE_forest_type_species$plot_id)) %>%
  arrange(desc(frequency))


WY_seedling_count <- WY_SEEDLING_forest_type_species %>%
  group_by(plot_id, INVYR) %>%
  summarise(
    total_seedlings_in_plot = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop"
  )

WY_seedling_species_dominance <- WY_SEEDLING_forest_type_species %>%
  group_by(plot_id, INVYR, species) %>%
  summarise(species_count = sum(TREECOUNT, na.rm = TRUE), .groups = "drop") %>%
  left_join(WY_seedling_count, by = c("plot_id", "INVYR")) %>%
  filter(total_seedlings_in_plot > 0) %>%
  mutate(relative_abundance = species_count / total_seedlings_in_plot)

WY_seedling_species_summary <- WY_seedling_species_dominance %>%
  group_by(species) %>%
  summarise(
    n_plots_present = n_distinct(plot_id),
    avg_dominance = mean(relative_abundance),
    median_dominance = median(relative_abundance),
    min_dominance = min(relative_abundance),
    max_dominance = max(relative_abundance),
    .groups = "drop"
  ) %>%
  mutate(
    frequency = n_plots_present / n_distinct(WY_seedling_species_dominance$plot_id)
  ) %>%
  arrange(desc(frequency))

```

#AZ
## whole plot network
```{r}

AZ_COND_IDs <- AZ_COND %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

AZ_PLOT_IDs <- AZ_PLOT %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

AZ_TREE_IDs <- AZ_TREE %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

AZ_SEEDLING_IDs <- AZ_SEEDLING %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

AZ_COND_forest_type <- AZ_COND_IDs %>%
  filter(FORTYPCD %in% c(185, 182, 184,
201, 202, 221, 222, 224, 225, 226, 
241, 261, 262, 263, 264, 265, 267, 268, 270,
271, 281, 301, 304, 305, 321, 
341, 342, 361, 365, 371,
901, 911, 912, 922, 923, 933, 941, 961))

AZ_forest_type_plot_id <- AZ_COND_forest_type %>%
  distinct(plot_id) %>%
  pull(plot_id)

length(AZ_forest_type_plot_id)

AZ_PLOT_forest_type <- AZ_PLOT_IDs %>%
  filter(plot_id %in% AZ_forest_type_plot_id)

AZ_TREE_forest_type <- AZ_TREE_IDs %>%
  filter(plot_id %in% AZ_forest_type_plot_id)

nrow(AZ_PLOT_forest_type)
n_distinct(AZ_TREE_forest_type$plot_id)

AZ_TREE_forest_type_species <- AZ_TREE_forest_type %>%
  left_join(
    REF_SPECIES %>%
      mutate(species = paste(GENUS, SPECIES, sep = "_")) %>%
      select(SPCD, species),
    by = "SPCD"
  )

sum(is.na(AZ_TREE_forest_type_species$species))
n_distinct(AZ_TREE_forest_type_species$species)

AZ_forest_type_all_species <- AZ_TREE_forest_type_species %>%
  distinct(species) %>%
  pull(species)


AZ_SEEDLING_forest_type <- AZ_SEEDLING_IDs %>%
  filter(plot_id %in% AZ_forest_type_plot_id)

AZ_SEEDLING_forest_type_species <- AZ_SEEDLING_forest_type %>%
  left_join(
    REF_SPECIES %>%
      mutate(species = paste(GENUS, SPECIES, sep = "_")) %>%
      select(SPCD, species),
    by = "SPCD"
  )

AZ_forest_type_all_seedling_species <- AZ_SEEDLING_forest_type_species %>%
  distinct(species) %>%
  pull(species)

length(AZ_forest_type_all_species)
length(AZ_forest_type_all_seedling_species)

intersect(AZ_forest_type_all_species, AZ_forest_type_all_seedling_species)

setdiff(AZ_forest_type_all_species, AZ_forest_type_all_seedling_species)

setdiff(AZ_forest_type_all_seedling_species, AZ_forest_type_all_species)

AZ_species <- data.frame(
  species = unique(c(
    AZ_forest_type_all_species,
    AZ_forest_type_all_seedling_species
  ))
)
```

## dominance
```{r}
AZ_tree_count <- AZ_TREE_forest_type_species %>%
  group_by(plot_id, INVYR) %>%
  summarise(total_trees_in_plot = n(), .groups = "drop")

AZ_tree_species_dominance <- AZ_TREE_forest_type_species %>%
  group_by(plot_id, INVYR, species) %>%
  summarise(species_count = n(), .groups = "drop") %>%
  left_join(AZ_tree_count, by = c("plot_id", "INVYR")) %>%
  mutate(relative_abundance = species_count / total_trees_in_plot)

AZ_tree_species_summary <- AZ_tree_species_dominance %>%
  group_by(species) %>%
  summarise(
    n_plots_present = n_distinct(plot_id),
    avg_dominance = mean(relative_abundance),
    median_dominance = median(relative_abundance),
    min_dominance = min(relative_abundance),
    max_dominance = max(relative_abundance),
    .groups = "drop"
  ) %>%
  mutate(frequency = n_plots_present / n_distinct(AZ_TREE_forest_type_species$plot_id)) %>%
  arrange(desc(frequency))


AZ_seedling_count <- AZ_SEEDLING_forest_type_species %>%
  group_by(plot_id, INVYR) %>%
  summarise(
    total_seedlings_in_plot = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop"
  )

AZ_seedling_species_dominance <- AZ_SEEDLING_forest_type_species %>%
  group_by(plot_id, INVYR, species) %>%
  summarise(species_count = sum(TREECOUNT, na.rm = TRUE), .groups = "drop") %>%
  left_join(AZ_seedling_count, by = c("plot_id", "INVYR")) %>%
  filter(total_seedlings_in_plot > 0) %>%
  mutate(relative_abundance = species_count / total_seedlings_in_plot)

AZ_seedling_species_summary <- AZ_seedling_species_dominance %>%
  group_by(species) %>%
  summarise(
    n_plots_present = n_distinct(plot_id),
    avg_dominance = mean(relative_abundance),
    median_dominance = median(relative_abundance),
    min_dominance = min(relative_abundance),
    max_dominance = max(relative_abundance),
    .groups = "drop"
  ) %>%
  mutate(
    frequency = n_plots_present / n_distinct(AZ_seedling_species_dominance$plot_id)
  ) %>%
  arrange(desc(frequency))
```

# NM
## whole plot network
```{r}

NM_COND_IDs <- NM_COND %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

NM_PLOT_IDs <- NM_PLOT %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

NM_TREE_IDs <- NM_TREE %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

NM_SEEDLING_IDs <- NM_SEEDLING %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

NM_COND_forest_type <- NM_COND_IDs %>%
  filter(FORTYPCD %in% c(185, 182, 184,
201, 202, 221, 222, 224, 225, 226, 
241, 261, 262, 263, 264, 265, 267, 268, 270,
271, 281, 301, 304, 305, 321, 
341, 342, 361, 365, 371,
901, 911, 912, 922, 923, 933, 941, 961))

NM_forest_type_plot_id <- NM_COND_forest_type %>%
  distinct(plot_id) %>%
  pull(plot_id)

length(NM_forest_type_plot_id)

NM_PLOT_forest_type <- NM_PLOT_IDs %>%
  filter(plot_id %in% NM_forest_type_plot_id)

NM_TREE_forest_type <- NM_TREE_IDs %>%
  filter(plot_id %in% NM_forest_type_plot_id)

nrow(NM_PLOT_forest_type)
n_distinct(NM_TREE_forest_type$plot_id)

NM_TREE_forest_type_species <- NM_TREE_forest_type %>%
  left_join(
    REF_SPECIES %>%
      mutate(species = paste(GENUS, SPECIES, sep = "_")) %>%
      select(SPCD, species),
    by = "SPCD"
  )

sum(is.na(NM_TREE_forest_type_species$species))
n_distinct(NM_TREE_forest_type_species$species)

NM_forest_type_all_species <- NM_TREE_forest_type_species %>%
  distinct(species) %>%
  pull(species)


NM_SEEDLING_forest_type <- NM_SEEDLING_IDs %>%
  filter(plot_id %in% NM_forest_type_plot_id)

NM_SEEDLING_forest_type_species <- NM_SEEDLING_forest_type %>%
  left_join(
    REF_SPECIES %>%
      mutate(species = paste(GENUS, SPECIES, sep = "_")) %>%
      select(SPCD, species),
    by = "SPCD"
  )

NM_forest_type_all_seedling_species <- NM_SEEDLING_forest_type_species %>%
  distinct(species) %>%
  pull(species)

length(NM_forest_type_all_species)
length(NM_forest_type_all_seedling_species)

intersect(NM_forest_type_all_species, NM_forest_type_all_seedling_species)

setdiff(NM_forest_type_all_species, NM_forest_type_all_seedling_species)

setdiff(NM_forest_type_all_seedling_species, NM_forest_type_all_species)

NM_species <- data.frame(
  species = unique(c(
    NM_forest_type_all_species,
    NM_forest_type_all_seedling_species
  ))
)

```

## dominance
```{r}
NM_tree_count <- NM_TREE_forest_type_species %>%
  group_by(plot_id, INVYR) %>%
  summarise(total_trees_in_plot = n(), .groups = "drop")

NM_tree_species_dominance <- NM_TREE_forest_type_species %>%
  group_by(plot_id, INVYR, species) %>%
  summarise(species_count = n(), .groups = "drop") %>%
  left_join(NM_tree_count, by = c("plot_id", "INVYR")) %>%
  mutate(relative_abundance = species_count / total_trees_in_plot)

NM_tree_species_summary <- NM_tree_species_dominance %>%
  group_by(species) %>%
  summarise(
    n_plots_present = n_distinct(plot_id),
    avg_dominance = mean(relative_abundance),
    median_dominance = median(relative_abundance),
    min_dominance = min(relative_abundance),
    max_dominance = max(relative_abundance),
    .groups = "drop"
  ) %>%
  mutate(frequency = n_plots_present / n_distinct(NM_TREE_forest_type_species$plot_id)) %>%
  arrange(desc(frequency))


NM_seedling_count <- NM_SEEDLING_forest_type_species %>%
  group_by(plot_id, INVYR) %>%
  summarise(
    total_seedlings_in_plot = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop"
  )

NM_seedling_species_dominance <- NM_SEEDLING_forest_type_species %>%
  group_by(plot_id, INVYR, species) %>%
  summarise(species_count = sum(TREECOUNT, na.rm = TRUE), .groups = "drop") %>%
  left_join(NM_seedling_count, by = c("plot_id", "INVYR")) %>%
  filter(total_seedlings_in_plot > 0) %>%
  mutate(relative_abundance = species_count / total_seedlings_in_plot)

NM_seedling_species_summary <- NM_seedling_species_dominance %>%
  group_by(species) %>%
  summarise(
    n_plots_present = n_distinct(plot_id),
    avg_dominance = mean(relative_abundance),
    median_dominance = median(relative_abundance),
    min_dominance = min(relative_abundance),
    max_dominance = max(relative_abundance),
    .groups = "drop"
  ) %>%
  mutate(
    frequency = n_plots_present / n_distinct(NM_seedling_species_dominance$plot_id)
  ) %>%
  arrange(desc(frequency))

```



# merge all states
```{r}

# every plot x inventory year combination of relative abundance for each species for each state

all_states_tree_dominance <- bind_rows(
  CA_tree_species_dominance %>% mutate(state = "CA"),
  OR_tree_species_dominance %>% mutate(state = "OR"),
  WA_tree_species_dominance %>% mutate(state = "WA"),
  ID_tree_species_dominance %>% mutate(state = "ID"),
  MT_tree_species_dominance %>% mutate(state = "MT"),
  CO_tree_species_dominance %>% mutate(state = "CO"),
  WY_tree_species_dominance %>% mutate(state = "WY"),
  AZ_tree_species_dominance %>% mutate(state = "AZ"),
  NM_tree_species_dominance %>% mutate(state = "NM"),
  UT_tree_species_dominance %>% mutate(state = "UT"),
  NV_tree_species_dominance %>% mutate(state = "NV")
)

# each plot should have multiple entries for different inventory years
all_states_tree_dominance %>%
  distinct(state, plot_id, INVYR) %>%
  count(state, plot_id, name = "n_surveys") %>%
  count(n_surveys, name = "n_plots")

# n_surveys n_plots
# 1   23406
# 2   34290
# 3    4856
# 4      19



# and for seedlings separately 
all_states_seedling_dominance <- bind_rows(
  CA_seedling_species_dominance %>% mutate(state = "CA"),
  OR_seedling_species_dominance %>% mutate(state = "OR"),
  WA_seedling_species_dominance %>% mutate(state = "WA"),
  ID_seedling_species_dominance %>% mutate(state = "ID"),
  MT_seedling_species_dominance %>% mutate(state = "MT"),
  CO_seedling_species_dominance %>% mutate(state = "CO"),
  WY_seedling_species_dominance %>% mutate(state = "WY"),
  AZ_seedling_species_dominance %>% mutate(state = "AZ"),
  NM_seedling_species_dominance %>% mutate(state = "NM"),
  UT_seedling_species_dominance %>% mutate(state = "UT"),
  NV_seedling_species_dominance %>% mutate(state = "NV")
)

all_states_seedling_dominance %>%
  distinct(state, plot_id, INVYR) %>%
  count(state, plot_id, name = "n_surveys") %>%
  count(n_surveys, name = "n_plots")

# n_surveys n_plots
# 1   10395
# 2   21079
# 3    1537
# 4       8

# create a summary table for dominance for each species across all states for each plot x inventory year combination that the species is present in 
all_states_tree_species_summary <- all_states_tree_dominance %>%
  group_by(species) %>%
  summarise(
    n_plots_present  = n_distinct(paste(state, plot_id)),
    avg_dominance    = mean(relative_abundance),
    median_dominance = median(relative_abundance),
    min_dominance    = min(relative_abundance),
    max_dominance    = max(relative_abundance),
    .groups = "drop"
  ) %>%
  mutate(frequency = n_plots_present / n_distinct(paste(all_states_tree_dominance$state, all_states_tree_dominance$plot_id))) %>%
  arrange(desc(median_dominance))

all_states_seedling_species_summary <- all_states_seedling_dominance %>%
  group_by(species) %>%
  summarise(
    n_plots_present  = n_distinct(paste(state, plot_id)),
    avg_dominance    = mean(relative_abundance),
    median_dominance = median(relative_abundance),
    min_dominance    = min(relative_abundance),
    max_dominance    = max(relative_abundance),
    .groups = "drop"
  ) %>%
  mutate(
    frequency = n_plots_present / n_distinct(paste(all_states_seedling_dominance$state, all_states_seedling_dominance$plot_id))
  ) %>%
  arrange(desc(median_dominance))


# remove unhelpful names
all_states_seedling_species_summary <- all_states_seedling_species_summary %>%
  filter(!species %in% c(
    # FIA unknown categories
    "Tree_unknown",
    "Tree_broadleaf",

    # genus-level / non-specific categories
    "Fraxinus_spp.",
    "Juglans_spp.",
    "Populus_spp.",
    "Prunus_spp."
  ))

all_states_tree_species_summary <- all_states_tree_species_summary %>%
  filter(!species %in% c(
    # FIA unknown categories
    "Tree_unknown",
    "Tree_broadleaf",
    "Tree_evergreen",

    # genus-level / non-specific categories
    "Acer_spp.",
    "Quercus_spp.",
    "Fraxinus_spp.",
    "Juglans_spp.",
    "Populus_spp.",
    "Salix_spp.",
    "Betula_spp.",
    "Prunus_spp.",
    "Crataegus_spp.",
    "Malus_spp.",
    "Aesculus_spp.",
    "Eucalyptus_spp.",
    "Acacia_spp.",
    "Prosopis_spp."
  ))

all_states_tree_species_summary <- all_states_tree_species_summary %>%
  mutate(
    dominance_rank = rank(-median_dominance, ties.method = "average"),
    frequency_rank = rank(-frequency, ties.method = "average"),
    combined_rank = (dominance_rank + frequency_rank) / 2
  ) %>%
  arrange(combined_rank)

all_states_seedling_species_summary <- all_states_seedling_species_summary %>%
  mutate(
    dominance_rank = rank(-median_dominance, ties.method = "average"),
    frequency_rank = rank(-frequency, ties.method = "average"),
    combined_rank = (dominance_rank + frequency_rank) / 2
  ) %>%
  arrange(combined_rank)


# are there species in seedling dataset not in tree dataset?
setdiff(
  unique(all_states_seedling_species_summary$species),
  unique(all_states_tree_species_summary$species)
)


# how different is this list from CA_OR_WA_forest_type_species_overlap
setdiff(
  all_states_tree_species_summary$species,
  CA_OR_WA_forest_type_species_overlap
)

# is everthing in CA_OR_WA in all species?
setdiff(
  CA_OR_WA_forest_type_species_overlap,
  all_states_tree_species_summary$species
)

intersect(all_states_tree_species_summary$species, CA_OR_WA_forest_type_species_overlap) 

# assign each species to one ecological forest category
species <- all_states_tree_species_summary$species

category_map <- c(
  # Core mixed conifer forests
  "Pseudotsuga_menziesii" = "Core mixed conifer forests",
  "Abies_concolor" = "Core mixed conifer forests",
  "Pinus_ponderosa" = "Core mixed conifer forests",
  "Pinus_lambertiana" = "Core mixed conifer forests",
  "Calocedrus_decurrens" = "Core mixed conifer forests",
  "Pinus_jeffreyi" = "Core mixed conifer forests",
  "Abies_magnifica" = "Core mixed conifer forests",

  # Mixed conifer hardwood associates
  "Quercus_chrysolepis" = "Mixed conifer hardwood associates",
  "Quercus_kelloggii" = "Mixed conifer hardwood associates",
  "Quercus_wislizeni" = "Mixed conifer hardwood associates",
  "Lithocarpus_densiflorus" = "Mixed conifer hardwood associates",
  "Arbutus_menziesii" = "Mixed conifer hardwood associates",
  "Chrysolepis_chrysophylla" = "Mixed conifer hardwood associates",
  "Cornus_nuttallii" = "Mixed conifer hardwood associates",
  "Acer_macrophyllum" = "Mixed conifer hardwood associates",
  "Umbellularia_californica" = "Mixed conifer hardwood associates",
  "Quercus_garryana" = "Mixed conifer hardwood associates",
  "Quercus_douglasii" = "Mixed conifer hardwood associates",

  # Pacific Northwest moist conifer forests
  "Tsuga_heterophylla" = "Pacific Northwest moist conifer forests",
  "Thuja_plicata" = "Pacific Northwest moist conifer forests",
  "Picea_sitchensis" = "Pacific Northwest moist conifer forests",
  "Abies_amabilis" = "Pacific Northwest moist conifer forests",
  "Abies_grandis" = "Pacific Northwest moist conifer forests",
  "Abies_procera" = "Pacific Northwest moist conifer forests",
  "Chamaecyparis_lawsoniana" = "Pacific Northwest moist conifer forests",
  "Chamaecyparis_nootkatensis" = "Pacific Northwest moist conifer forests",
  "Larix_occidentalis" = "Pacific Northwest moist conifer forests",
  "Alnus_rubra" = "Pacific Northwest moist conifer forests",
  "Picea_breweriana" = "Pacific Northwest moist conifer forests",
  "Taxus_brevifolia" = "Pacific Northwest moist conifer forests",

  # Southwestern dry mixed conifer forests
  "Pinus_arizonica" = "Southwestern dry mixed conifer forests",
  "Pinus_engelmannii" = "Southwestern dry mixed conifer forests",
  "Pinus_strobiformis" = "Southwestern dry mixed conifer forests",
  "Pseudotsuga_macrocarpa" = "Southwestern dry mixed conifer forests",
  "Quercus_arizonica" = "Southwestern dry mixed conifer forests",
  "Quercus_grisea" = "Southwestern dry mixed conifer forests",
  "Quercus_rugosa" = "Southwestern dry mixed conifer forests",

  # Pinyon-juniper and dry woodland forests
  "Pinus_edulis" = "Pinyon-juniper and dry woodland forests",
  "Pinus_monophylla" = "Pinyon-juniper and dry woodland forests",
  "Pinus_cembroides" = "Pinyon-juniper and dry woodland forests",
  "Pinus_discolor" = "Pinyon-juniper and dry woodland forests",
  "Juniperus_osteosperma" = "Pinyon-juniper and dry woodland forests",
  "Juniperus_monosperma" = "Pinyon-juniper and dry woodland forests",
  "Juniperus_coahuilensis" = "Pinyon-juniper and dry woodland forests",
  "Juniperus_deppeana" = "Pinyon-juniper and dry woodland forests",
  "Juniperus_pinchotii" = "Pinyon-juniper and dry woodland forests",
  "Juniperus_californica" = "Pinyon-juniper and dry woodland forests",
  "Juniperus_occidentalis" = "Pinyon-juniper and dry woodland forests",
  "Juniperus_scopulorum" = "Pinyon-juniper and dry woodland forests",
  "Cupressus_arizonica" = "Pinyon-juniper and dry woodland forests",
  "Quercus_oblongifolia" = "Pinyon-juniper and dry woodland forests",

  # Subalpine / high elevation conifer forests
  "Pinus_albicaulis" = "Subalpine/high elevation conifer forests",
  "Pinus_flexilis" = "Subalpine/high elevation conifer forests",
  "Pinus_longaeva" = "Subalpine/high elevation conifer forests",
  "Pinus_aristata" = "Subalpine/high elevation conifer forests",
  "Pinus_balfouriana" = "Subalpine/high elevation conifer forests",
  "Abies_lasiocarpa" = "Subalpine/high elevation conifer forests",
  "Abies_shastensis" = "Subalpine/high elevation conifer forests",
  "Picea_engelmannii" = "Subalpine/high elevation conifer forests",
  "Tsuga_mertensiana" = "Subalpine/high elevation conifer forests",
  "Larix_lyallii" = "Subalpine/high elevation conifer forests",
  "Pinus_washoensis" = "Subalpine/high elevation conifer forests",
  "Picea_glauca" = "Subalpine/high elevation conifer forests",

  # Aspen / post-fire deciduous forests
  "Populus_tremuloides" = "Aspen/post-fire deciduous forests",
  "Populus_angustifolia" = "Aspen/post-fire deciduous forests",
  "Betula_papyrifera" = "Aspen/post-fire deciduous forests",

  # Coastal California fire-prone forests
  "Sequoia_sempervirens" = "Coastal California fire-prone forests",
  "Sequoiadendron_giganteum" = "Coastal California fire-prone forests",
  "Pinus_muricata" = "Coastal California fire-prone forests",
  "Pinus_radiata" = "Coastal California fire-prone forests",
  "Pinus_attenuata" = "Coastal California fire-prone forests",
  "Pinus_sabiniana" = "Coastal California fire-prone forests",

  # Non-native / exclude
  "Pyrus_calleryana" = "Non-native / exclude",
  "Robinia_pseudoacacia" = "Non-native / exclude",
  "Eucalyptus_globulus" = "Non-native / exclude",
  "Prunus_domestica" = "Non-native / exclude",
  "Prunus_avium" = "Non-native / exclude",
  "Ulmus_pumila" = "Non-native / exclude",
  "Acer_platanoides" = "Non-native / exclude",
  "Ilex_aquifolium" = "Non-native / exclude",
  "Ailanthus_altissima" = "Non-native / exclude",
  "Salix_alba" = "Non-native / exclude",
  "Salix_x sepulcralis" = "Non-native / exclude",
  "Pinus_sylvestris" = "Non-native / exclude"
)

# Anything not explicitly mapped above falls into "Other native woody species"
all_states_tree_species_summary$forest_category <- ifelse(
  species %in% names(category_map),
  category_map[species],
  "Other native woody species"
)

write.csv(all_states_tree_species_summary, "all_states_species_list_07_31_2026.csv", row.names = FALSE)



all_states_plot_ids <- bind_rows(
  CA_TREE_forest_type_species %>% distinct(plot_id) %>% mutate(state = "CA"),
  OR_TREE_forest_type_species %>% distinct(plot_id) %>% mutate(state = "OR"),
  WA_TREE_forest_type_species %>% distinct(plot_id) %>% mutate(state = "WA"),
  ID_TREE_forest_type_species %>% distinct(plot_id) %>% mutate(state = "ID"),
  MT_TREE_forest_type_species %>% distinct(plot_id) %>% mutate(state = "MT"),
  CO_TREE_forest_type_species %>% distinct(plot_id) %>% mutate(state = "CO"),
  WY_TREE_forest_type_species %>% distinct(plot_id) %>% mutate(state = "WY"),
  AZ_TREE_forest_type_species %>% distinct(plot_id) %>% mutate(state = "AZ"),
  NM_TREE_forest_type_species %>% distinct(plot_id) %>% mutate(state = "NM"),
  UT_TREE_forest_type_species %>% distinct(plot_id) %>% mutate(state = "UT"),
  NV_TREE_forest_type_species %>% distinct(plot_id) %>% mutate(state = "NV")
)

n_distinct(paste(all_states_plot_ids$state, all_states_plot_ids$plot_id))


```

```{r}
mixed_conifer_species <- c(
  # core mixed conifer conifers
  "Pinus_contorta",
  "Pinus_ponderosa",
  "Pseudotsuga_menziesii",
  "Abies_concolor",
  "Abies_magnifica",
  "Abies_grandis",
  "Abies_lasiocarpa",
  "Abies_amabilis",
  "Abies_procera",
  "Abies_shastensis",
  "Tsuga_heterophylla",
  "Tsuga_mertensiana",
  "Picea_engelmannii",
  "Picea_pungens",
  "Picea_breweriana",
  "Larix_occidentalis",
  "Larix_lyallii",
  "Calocedrus_decurrens",
  "Thuja_plicata",
  "Chamaecyparis_lawsoniana",
  "Chamaecyparis_nootkatensis",
  "Pinus_lambertiana",
  "Pinus_jeffreyi",
  "Pinus_coulteri",
  "Pinus_strobiformis",
  "Pinus_monticola",
  "Pinus_flexilis",
  "Pinus_albicaulis",
  "Pinus_balfouriana",
  "Pinus_washoensis",
  "Sequoiadendron_giganteum",

  # hardwood associates / mixed conifer associates
  "Quercus_kelloggii",
  "Quercus_chrysolepis",
  "Quercus_gambelii",
  "Populus_tremuloides",
  "Acer_glabrum",
  "Acer_grandidentatum",
  "Populus_spp.",
  "Populus_angustifolia",
  "Crataegus_chrysocarpa",
  "Prunus_emarginata",
  "Prunus_virginiana",
  "Cercocarpus_ledifolius",
  "Arbutus_menziesii",
  "Arbutus_arizonica",
  "Pinus_arizonica",
  "Pinus_engelmannii"
)

mixed_conifer_species_summary <- all_states_tree_species_summary %>%
  filter(species %in% mixed_conifer_species)

n_distinct(mixed_conifer_species_summary$species)



# looking at seedling dominance per state and across states

CA_seedling_species_summary %>%
  filter(median_dominance > 0.01) %>%
  filter(!species %in% (CA_species_summary %>% filter(median_dominance > 0.01) %>% pull(species)))

OR_seedling_species_summary %>%
  filter(median_dominance > 0.01) %>%
  filter(!species %in% (OR_species_summary %>% filter(median_dominance > 0.01) %>% pull(species)))

WA_seedling_species_summary %>%
  filter(median_dominance > 0.01) %>%
  filter(!species %in% (WA_species_summary %>% filter(median_dominance > 0.01) %>% pull(species)))

ID_seedling_species_summary %>%
  filter(median_dominance > 0.01) %>%
  filter(!species %in% (ID_tree_species_summary %>% filter(median_dominance > 0.01) %>% pull(species)))


```

