#load packages
```{r}
library(sf)
library(terra)
library(dplyr)
library(raster)
library(terra)
install.packages('tidyverse')
library(tidyverse)
library(here)
library(rnaturalearth)
library(rnaturalearthdata)
install.packages('rnaturalearthhires')
install.packages("remotes")
remotes::install_github("ropensci/rnaturalearthhires")
library(gridExtra)
install.packages('geosphere')
library(geosphere)
library(fields)

```
#load data
```{r}
# load in FIA tree data for western North America (California, Oregon, Washington, Colorado, Utah, Nevada, Wyoming) to encompass the Sierra Nevada, Northern and Southern Coast Ranges, maybe also include New Mexico and Montana.

# Load points for fuzzed FIA plots
CA_PLOT <- read.csv("./data/FIA/CA_PLOT.csv")
CO_PLOT <- read.csv("./data/FIA/CO_PLOT.csv")
NV_PLOT <- read.csv("./data/FIA/NV_PLOT.csv")
OR_PLOT <- read.csv("./data/FIA/OR_PLOT.csv")
UT_PLOT <- read.csv("./data/FIA/UT_PLOT.csv")
WA_PLOT <- read.csv("./data/FIA/WA_PLOT.csv")
WY_PLOT <- read.csv("./data/FIA/WY_PLOT.csv")
ID_PLOT <- read.csv("./data/FIA/ID_PLOT.csv")
MT_PLOT <- read.csv("./data/FIA/MT_PLOT.csv")

# then load in condition code files for each state
CA_COND <- read.csv("./data/FIA/CA_COND.csv")
CO_COND <- read.csv("./data/FIA/CO_COND.csv")
NV_COND <- read.csv("./data/FIA/NV_COND.csv")
OR_COND <- read.csv("./data/FIA/OR_COND.csv")
UT_COND <- read.csv("./data/FIA/UT_COND.csv")
WA_COND <- read.csv("./data/FIA/WA_COND.csv")
WY_COND <- read.csv("./data/FIA/WY_COND.csv")
ID_COND <- read.csv("./data/FIA/ID_COND.csv")
MT_COND <- read.csv("./data/FIA/MT_COND.csv")

# then load in TREE level data for each state
CA_TREE <- read.csv("./data/FIA/CA_TREE.csv")
CO_TREE <- read.csv("./data/FIA/CO_TREE.csv")
NV_TREE <- read.csv("./data/FIA/NV_TREE.csv")
OR_TREE <- read.csv("./data/FIA/OR_TREE.csv")
UT_TREE <- read.csv("./data/FIA/UT_TREE.csv")
WA_TREE <- read.csv("./data/FIA/WA_TREE.csv")
WY_TREE <- read.csv("./data/FIA/WY_TREE.csv")
ID_TREE <- read.csv("./data/FIA/ID_TREE.csv")
MT_TREE <- read.csv("./data/FIA/MT_TREE.csv")

CA_SEEDLING <- read.csv("./data/FIA/CA_SEEDLING.csv")
CO_SEEDLING <- read.csv("./data/FIA/CO_SEEDLING.csv")
NV_SEEDLING <- read.csv("./data/FIA/NV_SEEDLING.csv")
OR_SEEDLING <- read.csv("./data/FIA/OR_SEEDLING.csv")
UT_SEEDLING <- read.csv("./data/FIA/UT_SEEDLING.csv")
WA_SEEDLING <- read.csv("./data/FIA/WA_SEEDLING.csv")
WY_SEEDLING <- read.csv("./data/FIA/WY_SEEDLING.csv")
ID_SEEDLING <- read.csv("./data/FIA/ID_SEEDLING.csv")
MT_SEEDLING <- read.csv("./data/FIA/MT_SEEDLING.csv")


# Load in reference species names
REF_SPECIES <- read.csv("./data/FIA/REF_SPECIES.csv")

# load terra climate data
ppt_mean <- rast("data/terra_climate/ppt_30yr_mean.tif")
def_mean <- rast("data/terra_climate/def_30yr_mean.tif")
vpd_mean <- rast("data/terra_climate/vpd_30yr_mean.tif")
aet_mean <- rast("data/terra_climate/aet_30yr_mean.tif")
pet_mean <- rast("data/terra_climate/pet_30yr_mean.tif")
tmax_mean <- rast("data/terra_climate/tmax_30yr_mean.tif")
tmin_mean <- rast("data/terra_climate/tmin_30yr_mean.tif")

```

# filter by disturbance and forest type
```{r}
# filter plots to DSTRBCD1 == 30, 31, or 32, this is selecting all plots with any fire damage [30], (from crown and ground fire, either prescribed or natural).
# [31] = ground fire, [32] = canopy fire
# and then filter CA burned plots to focal forest types including mixed conifer, douglas-fir, ponderosa pine, incense cedar, sugar pine, jefferyi, pine groups, and white fir, and CA black oak groups

CA_COND_filtered <- CA_COND %>%
  filter(if_any(
    c(DSTRBCD1, DSTRBCD2, DSTRBCD3),
    ~ .x %in% c(30, 31, 32)
  )) %>%
  filter(FORTYPCD %in% c(371, 201, 221, 222, 224, 225, 261, 922))

# how many unique PLT_CN values have multiple rows for CONDID

CA_COND_filtered %>%
  dplyr::group_by(PLT_CN) %>%
  dplyr::summarise(n_condid = n_distinct(CONDID), .groups = "drop") %>%
  dplyr::filter(n_condid > 1) %>%
  nrow()

# 39 / 506 have multiple rows for CONDID per PLT_CN

# CA_COND has multiple rows per PLT_CN due to variation in CONDID
#to collapse CA_COND table to have 1 row per PLT_CN first check that CONDID is the only thing that differs across rows with the same PLT_CN, we just want to make sure DSTRBCD dont differ across PLT_CN in COND table

CA_COND_filtered %>%
  dplyr::group_by(PLT_CN) %>%
  dplyr::summarise(
    problem = dplyr::n_distinct(
      paste(DSTRBCD1, DSTRBCD2, DSTRBCD3,
            DSTRBYR1, DSTRBYR2, DSTRBYR3)
    ) > 1,
    .groups = "drop"
  ) %>%
  dplyr::summarise(any_problem = any(problem))


CA_COND_filtered %>%
  group_by(PLT_CN) %>%
  summarise(
    n_profiles = n_distinct(
      paste(DSTRBCD1, DSTRBCD2, DSTRBCD3,
            DSTRBYR1, DSTRBYR2, DSTRBYR3)
    ),
    .groups = "drop"
  ) %>%
  filter(n_profiles > 1) %>%
  nrow()

# 10 PLT_CNs have differing DSTRBCD between rows of the same PLT_CN in COND table

CA_duplicate_CONID<- CA_COND_filtered %>%
  group_by(PLT_CN) %>%
  filter(
    n_distinct(
      paste(DSTRBCD1, DSTRBCD2, DSTRBCD3,
            DSTRBYR1, DSTRBYR2, DSTRBYR3)
    ) > 1
  ) %>%
  arrange(PLT_CN) %>%
  ungroup()

#variation in COND rows per plot may be ecologically meaningful
# when multiple rows for a PLT_CN are present, disturbance only differs between 31-32

# lets simplify by taking the dominant condition using 
# CONDPROP_UNADJ. the dominant condition = the largest CONDPROP_UNADJ for a given PLT_CN. the proportion of the plot occupied by that condition (ranges from # 0–1, sums to ~1 across conditions in a plot)

CA_COND_dominant <- CA_COND_filtered %>%
  group_by(PLT_CN) %>%
  slice_max(CONDPROP_UNADJ, n = 1, with_ties = FALSE) %>%
  ungroup()

CA_COND_summary <- CA_COND_dominant %>%
  dplyr::select(
    PLT_CN,
    CONDPROP_UNADJ,
    FORTYPCD,
    DSTRBCD1, DSTRBYR1,
    DSTRBCD2, DSTRBYR2,
    DSTRBCD3, DSTRBYR3
  )

# attach condition summary to PLOT table (with resurveys) to get the condition history for that plot. the same history will be attached to each resurvey for the plot
CA_burned_PLT_CN <- CA_COND_summary %>%
  distinct(PLT_CN)

CA_PLOT_filtered <- CA_PLOT %>%
  semi_join(CA_burned_PLT_CN, by = c("CN" = "PLT_CN"))

CA_PLOT_by_COND <- CA_PLOT_filtered %>%
  left_join(CA_COND_summary, by = c("CN" = "PLT_CN"))
```

#pre and post-fire surveys
```{r}
# number of revisits

# we can link CN to PREV_PLT_CN to determine resurveys 
# however, PREV_PLT_CN is not recorded for 'periodic' inventories
CA_PLOT_pairs <- CA_PLOT_by_COND %>%
  filter(!is.na(PREV_PLT_CN)) %>%
  left_join(
    CA_PLOT_by_COND %>% dplyr::select(CN, INVYR),
    by = c("PREV_PLT_CN" = "CN"),
    suffix = c("_current", "_previous")
  )

# reshape fire year data to long format 
# each row becomes one fire event so we can determine pre and post fire surveys
CA_fires <- CA_PLOT_by_COND %>%
  tidyr::pivot_longer(
    cols = c(DSTRBYR1, DSTRBYR2, DSTRBYR3),
    names_to = "fire_slot",
    values_to = "fire_year"
  ) %>%
  filter(!is.na(fire_year))

# attach fire events to pre and post visit pairs
CA_fire_events <- CA_PLOT_pairs %>%
  left_join(CA_fires, by = "CN")

# filter for plots that have inventories before and after fire years 
# asking: is there a fire event between previous and current inventory
CA_fire_events <- CA_fire_events %>%
  mutate(
    pre_fire  = INVYR_previous < fire_year,
    post_fire = INVYR_current > fire_year
  ) %>%
  filter(pre_fire & post_fire)

## 22 plots with pre and post fire surveys


# the above code doesnt look at pre and post plots for ANY and ALL fires
# trying this instead 
CA_PLOT_pairs <- CA_PLOT_by_COND %>%
  filter(!is.na(PREV_PLT_CN)) %>%
  left_join(
    CA_PLOT_by_COND %>% dplyr::select(CN, INVYR),
    by = c("PREV_PLT_CN" = "CN"),
    suffix = c("_current", "_previous")
  )

CA_fires <- CA_PLOT_by_COND %>%
  tidyr::pivot_longer(
    cols = c(DSTRBYR1, DSTRBYR2, DSTRBYR3),
    values_to = "fire_year"
  ) %>%
  filter(!is.na(fire_year)) %>%
  distinct(CN, fire_year)

CA_fire_events <- CA_PLOT_pairs %>%
  left_join(CA_fires, by = "CN")

CA_fire_events <- CA_fire_events %>%
  mutate(
    in_interval = fire_year > INVYR_previous &
                  fire_year <= INVYR_current
  )

CA_fire_events %>%
  group_by(PREV_PLT_CN, CN) %>%
  summarise(any_fire = any(in_interval), .groups = "drop") %>%
  count(any_fire)

CA_fire_events %>%
  group_by(PREV_PLT_CN, CN) %>%
  summarise(
    any_fire = any(in_interval, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  count(any_fire)

# 27 plots that had a pre and post fire survey around ANY fire
# potentially more flexible looks for any combination of pre and post fire surveys around all fire years 

# if plot resurveys are being left out due to the 'periodic' versus 'annual' inventory distinction, lets check for resurveys a different way

# create a unique plot ID 
CA_PLOT_IDs <- CA_PLOT_by_COND %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )
CA_PLOT_IDs %>%
  dplyr::count(plot_id, name = "n_resurveys") %>%
  dplyr::mutate(group = ifelse(n_resurveys == 1, "n = 1", "n > 1")) %>%
  dplyr::count(group)

# n= 1 (401) ; n =2 (33)

# trying it 3 different ways there are between 22-33 plots with pre and post fire surveys

```
# CA
## fire only plots
```{r}
# create a unique PLOT ID in COND and in PLOT tables 
CA_COND_IDs <- CA_COND %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )
CA_PLOT_IDs <- CA_PLOT %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

# collapse information from multiple CONDIDs associated with each plot to avoid many-to-many mapping
# and then filter by fire and forest type  

#plot_id → multiple CONDIDs → each CONDID has up to 3 disturbance codes and years
# one row for a specific plot_id, CONID, and disturbance event (combining DSTRBCD and DSTRBYR)
CA_COND_long <- CA_COND_IDs %>%
  dplyr::select(plot_id, CONDID, FORTYPCD,
    DSTRBCD1, DSTRBCD2, DSTRBCD3,
    DSTRBYR1, DSTRBYR2, DSTRBYR3) %>%
  pivot_longer(
    cols = starts_with("DSTR"),
    names_to = c(".value", "event"),
    names_pattern = "(DSTRBCD|DSTRBYR)([1-3])"
  ) %>%
  filter(DSTRBCD %in% c(30, 31, 32)) %>%
  filter(FORTYPCD %in% c(371, 201, 221, 222, 224, 225, 261, 922)) %>%
  mutate(
    fire_year_severity = paste(DSTRBYR, DSTRBCD, sep = "_")
  )

# summarize fire history across rows for unique plot_id (one row per plot_id)
CA_fire_summary <- CA_COND_long %>%
  group_by(plot_id) %>%
  summarise(
    n_fire_events = n_distinct(DSTRBYR), # number of distinct fire years per plot
    
    # this just collapses all codes into a list but not necessarily relative to year
    fire_types = paste(sort(unique(DSTRBCD)), collapse = ", "), 
    fire_years = paste(sort(unique(DSTRBYR)), collapse = ", "),
    
    fire_year_severity =
    paste(unique(fire_year_severity), collapse = ", "),
    
    multiple_fires = n_distinct(DSTRBYR) > 1, # did fire occur more than once?
    multiple_severity = n_distinct(DSTRBCD) > 1, # are there multiple fire types?
    first_fire_year = min(DSTRBYR, na.rm = TRUE),
    last_fire_year = max(DSTRBYR, na.rm = TRUE),
    .groups = "drop"
  )

# how many plots burned multiple times?
CA_fire_summary %>% summarise(n_plots_multiple_fires = sum(multiple_fires))
# 34 plots

# how many plots that burned multipe times have different severities?
CA_fire_summary %>% filter(multiple_fires) %>% summarise(n_plots_multiple_fire_and_severity = sum(multiple_severity))
# 10 plots

# how many plots that burned once have multiple severities within a plot?
CA_fire_summary %>%
  filter(!multiple_fires) %>%
  summarise(
    n_plots_single_fire_multiple_severity =
      sum(multiple_severity))
# 9 plots

# pre_and_post, pre_only, and post_only
CA_burned_PLOT_IDs <- CA_fire_summary %>% 
  distinct(plot_id)

# merge COND and PLOT by this unique ID 
CA_PLOT_by_COND <- CA_PLOT_IDs %>%
  semi_join(CA_burned_PLOT_IDs, by = "plot_id") %>%  #keep only plot_ids with at least one fire 
  left_join(CA_fire_summary, by = "plot_id")  #add fire summary to each plot, should be one-to-many mapping

#for each unique plot, count how many unique inventory years (INVYR) it has
CA_PLOT_by_COND %>%
  group_by(plot_id) %>%
  summarise(
    n_revisits = n_distinct(INVYR),
    .groups = "drop"
  ) %>%
  count(n_revisits)

# n=2 (288)
# n=3 (146)

# how many plots have pre and post fire surveys
CA_fire_events <- CA_COND_long %>%
  dplyr::select(plot_id, DSTRBYR) %>%
  distinct() %>%
  rename(fire_year = DSTRBYR)

# does each unique plot_id have a pre and post fire survey around any/all fire years?
CA_prepost_fire_summary <- CA_PLOT_by_COND %>%
  distinct(plot_id, INVYR) %>%
  inner_join(
    CA_fire_events %>% distinct(plot_id, fire_year),
    by = "plot_id"
  ) %>%
  mutate(
    relationship = case_when(
      INVYR < fire_year ~ "pre",
      INVYR > fire_year ~ "post",
      TRUE ~ "same_year"
    )
  ) %>%
  group_by(plot_id, fire_year) %>%
  summarise(
    has_pre  = any(relationship == "pre"),
    has_post = any(relationship == "post"),
    n_pre_surveys  = sum(relationship == "pre"),
    n_post_surveys = sum(relationship == "post"),
    .groups = "drop"
  ) %>%
  mutate(
      status = case_when(
      has_pre & has_post ~ "pre_and_post",
      has_pre ~ "pre_only",
      has_post ~ "post_only",
      TRUE ~ "no_clear_separation"
    )
  )

CA_prepost_fire_summary %>%
  count(status)

# post only = 106
# pre only = 49
#pre and post = 313

CA_prepost_fire_summary %>%
  distinct(plot_id, status) %>%
  count(status)
# post_only = 106
# pre_only = 49
# pre_and_post = 304

# some plots are reused for surveys: the total number of unique plots is 393

# how many plots are there in CA
CA_PLOT %>%
  distinct(LAT, LON) %>%
  nrow()
# 22362

length(unique(CA_COND_IDs$plot_id)) # 22589
length(unique(CA_PLOT_IDs$plot_id)) # 22589

```

## tree
```{r}
#Tree number. A number used to uniquely identify a tree on a subplot. Tree numbers can be used to track trees when PLOT.DESIGNCD or DESIGNCD is the same between inventories

# before looking into tree level data, just confirm that the design codes are the same across plots in our CA_PLOT_by_COND table that has our burned plots
n_distinct(CA_PLOT_by_COND$DESIGNCD) # one design code across the filtered plots
unique(CA_PLOT_by_COND$DESIGNCD) # that design code is 501

# all plots have the same design code so you can proceed with linking trees across inventories

# do we have surviving trees in post-fire surveys?
# create unique plot ids for CA_TREE
CA_TREE_IDs <- CA_TREE %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

#filter CA_TREE to only keep the unique plot ids from CA_PLOT_by_COND (our fire only plots in our focal forest types)
CA_TREE_filtered <- CA_TREE_IDs %>%
  filter(plot_id %in% CA_PLOT_by_COND$plot_id)

# to locate a specific tree ID, use the following variables 
# STATECD, INVYR, UNITCD, COUNTYCD, PLOT, SUBP, TREE
# TREE is unique to SUBP not plot_id
# we can look for unique combinations of SUBP and TREE within a plot_id (STATECD, UNITCD, COUNTYCD, PLOT)

# to count how many times a tree was surveyed
# group by plot_id, SUBP, TREE, and then count the number of INVYR
CA_TREE_resurveys <- CA_TREE_filtered %>% 
  group_by(plot_id, SUBP, TREE) %>% 
  summarise(n_surveys = n_distinct(INVYR),
            survey_years = paste(sort(unique(INVYR)), collapse = ", "),
            .groups = "drop")

# how many times have trees been resurveyed
CA_TREE_resurveys %>%
  count(n_surveys)

#n=1 (1777)
#n=2 (12611)
#n=3 (4818)

# is TREE our unique identifier of TREE ?
CA_TREE_filtered %>%
  group_by(TREE) %>%
  summarise(
    n_plots = n_distinct(plot_id),
    .groups = "drop"
  ) %>%
  filter(n_plots > 1) %>%
  summarise(
    n_trees_multiple_plots = n()
  )
# 299 trees are found in multiple plots 

CA_TREE_filtered %>%
  summarise(
    n_unique_trees = n_distinct(plot_id, SUBP, TREE)
  )
# number of unique trees 19206

# Tree number notes:
# TREE uniquely identifies a tree within a subplot (SUBP),
# so full tree identity is: plot_id + SUBP + TREE

# unique tree identity is plot_id + SUBP + TREE
# make a unique tree identifier too? or use plot_id + SUBP + TREE everytime
#tree_id = paste(plot_id, SUBP, TREE, sep = "_") 

# lets look at trees in the plots with both pre and post fire surveys
# pull the plot_ids for status = pre_and_post from CA_prepost_fire_summary and the fire year the post-fire plot is associated with

CA_TREE_prepost_fire <- CA_TREE_filtered %>%
  semi_join(
    CA_prepost_fire_summary %>% filter(status == "pre_and_post"),
    by = "plot_id"
  ) %>%
  left_join(
    CA_prepost_fire_summary %>%
      filter(status == "pre_and_post") %>%
      dplyr::select(plot_id, fire_year, status),
    by = "plot_id"
  )

CA_prepost_fire_summary %>%
  filter(status == "pre_and_post") %>%
  summarise(n_plots = n_distinct(plot_id)) #304
length(unique(CA_TREE_prepost_fire$plot_id)) #304


# both CA_fire_summary and CA_TREE_prepost_fire have 304 unique plot_id with pre and post fire inventories

# now we want to look at just post-fire plots (where INVYR > fire_year)
CA_TREE_post_fire_plots <- CA_TREE_prepost_fire %>%
  filter(INVYR > fire_year)

# number of trees per plot (post-fire plots)
CA_TREE_per_plot_post <- CA_TREE_post_fire_plots %>%
  filter(STATUSCD == 1) %>%
  mutate(which = "post") %>%
  group_by(plot_id, INVYR, fire_year, status, which) %>%
  summarise(
    n_live_tree = n_distinct(SUBP, TREE),
    .groups = "drop"
  )

# number of species with STATUSCD = 1 in post-fire plots
CA_TREE_post_fire_plot_summary <- CA_TREE_post_fire_plots %>%
  filter(STATUSCD == 1) %>%
  group_by(plot_id, INVYR, fire_year) %>%
  summarise(
    n_live_species = n_distinct(SPCD),
    n_live_trees = n_distinct(SUBP, TREE),
    .groups = "drop"
  )

# number of individuals (live) per species per plot
CA_TREE_post_fire_species_summary <- CA_TREE_post_fire_plots %>%
  filter(STATUSCD == 1) %>%
  group_by(plot_id, INVYR, fire_year, SPCD) %>%
  summarise(
    n_live_trees = n_distinct(SUBP, TREE),
    .groups = "drop"
  ) %>%
  left_join(
    REF_SPECIES %>%
      dplyr::select(SPCD, GENUS, SPECIES),
    by = "SPCD"
  ) %>%
  mutate(species = paste(GENUS, SPECIES, sep = "_"))

### visualize plot level summaries

# number of live species per plot (post-fire plots)
CA_TREE_post_fire_species_summary %>%
  group_by(plot_id) %>%
  summarise(n_species = n_distinct(SPCD)) %>%
  ggplot(aes(x = n_species)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  theme_bw() +
  labs(x = "Number of live species per plot", y = "Number of plots")

# abundance of each species per plot (top 20 most abundant species)
ggplot(
  CA_TREE_post_fire_species_summary %>%
    group_by(species, plot_id) %>%
    summarise(n_live_trees = sum(n_live_trees), .groups = "drop") %>%
    group_by(species) %>%
    summarise(mean_per_plot = mean(n_live_trees), .groups = "drop") %>%
    slice_max(mean_per_plot, n = 20),
  aes(x = reorder(species, mean_per_plot),
      y = mean_per_plot)
) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    x = "species",
    y = "average # live trees per plot",
    title = "average abundance of species per plot"
  ) +
  theme_bw()

# boxplots of tree abundance per species per plot
CA_TREE_post_fire_plots %>%
  filter(STATUSCD == 1) %>%
  group_by(plot_id, SPCD) %>%
  summarise(
    n_trees = n_distinct(SUBP, TREE),
    .groups = "drop"
  ) %>%
  left_join(
    REF_SPECIES %>%
      dplyr::select(SPCD, GENUS, SPECIES),
    by = "SPCD"
  ) %>%
  mutate(species = paste(GENUS, SPECIES, sep = "_")) %>%
  ggplot(aes(x = species, y = n_trees)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6) +
  coord_flip() +
  labs(
    x = "species",
    y = "trees per plot",
    title = "tree abundance per species per plot"
  ) +
  scale_y_continuous(
    breaks = function(x) seq(0, max(x, na.rm = TRUE), by = 5)
  ) +
  theme_bw()


## tree transitions and mortality rates

# closest PRE-fire observation for each TREE
CA_TREE_pre_fire_trees <- CA_TREE_prepost_fire %>%
  filter(INVYR < fire_year) %>%
  group_by(plot_id, SUBP, TREE, fire_year) %>%
  arrange(desc(INVYR), .by_group = TRUE) %>%
  slice(1) %>%
  ungroup() %>%
  rename(
    pre_INVYR = INVYR,
    pre_STATUSCD = STATUSCD
  )

# closest POST-fire observation for each TREE
CA_TREE_post_fire_trees <- CA_TREE_prepost_fire %>%
  filter(INVYR > fire_year) %>%
  group_by(plot_id, SUBP, TREE, fire_year) %>%
  arrange(INVYR, .by_group = TRUE) %>%
  slice(1) %>%
  ungroup() %>%
  rename(
    post_INVYR = INVYR,
    post_STATUSCD = STATUSCD
  )

# join pre and post observations
CA_TREE_transitions <- CA_TREE_pre_fire_trees %>%
  inner_join(
    CA_TREE_post_fire_trees,
    by = c("plot_id", "SUBP", "TREE", "fire_year"),
    suffix = c("_pre", "_post")
  )

# mortality indicators
CA_TREE_transitions <- CA_TREE_transitions %>%
  mutate(
    alive_pre = pre_STATUSCD == 1,
    dead_post = pre_STATUSCD == 1 & post_STATUSCD == 2
  )

# plot-level mortality
CA_TREE_mortality <- CA_TREE_transitions %>%
  group_by(plot_id, fire_year) %>%
  summarise(
    n_alive_pre = sum(alive_pre, na.rm = TRUE),
    n_dead_post = sum(dead_post, na.rm = TRUE),
    mortality_rate = n_dead_post / n_alive_pre,
    .groups = "drop"
  )

summary(CA_TREE_mortality$mortality_rate)
hist(CA_TREE_mortality$mortality_rate)

# check for any new trees in post fire surveys

# trees appearing post-fire but absent pre-fire (ingrowth)
CA_TREE_new_trees <- CA_TREE_post_fire_trees %>%
  anti_join(
    CA_TREE_transitions %>%
      distinct(plot_id, SUBP, fire_year, TREE),
    by = c("plot_id", "SUBP", "fire_year", "TREE")
  )


CA_TREE_new_trees <- CA_TREE_post_fire_trees %>%
  filter(post_STATUSCD == 1) %>%
  anti_join(
    CA_TREE_transitions %>%
      distinct(plot_id, SUBP, fire_year, TREE),
    by = c("plot_id", "SUBP", "fire_year", "TREE")
  )

# recruits per plot
CA_TREE_new_recruits_per_plot <- CA_TREE_new_trees %>%
  group_by(plot_id, fire_year, post_INVYR) %>%
  summarise(
    n_new_recruits = n_distinct(SUBP, TREE),
    .groups = "drop"
  )

summary(CA_TREE_new_recruits_per_plot$n_new_recruits)


CA_TREE_recruitment <- CA_TREE_new_recruits_per_plot %>%
  left_join(
    CA_TREE_per_plot_post %>%
      dplyr::select(
        plot_id,
        fire_year,
        post_INVYR = INVYR,
        n_live_tree
      ),
    by = c("plot_id", "fire_year", "post_INVYR")
  ) %>%
  mutate(
    recruit_prop = n_new_recruits / n_live_tree
  )

CA_TREE_recruitment %>%
  filter(recruit_prop == 1)

# 13 plots have only new trees (live)
summary(CA_TREE_recruitment$recruit_prop)

```

## seedling
```{r}
# is there recruitment in post-fire surveys?

# seedling counts are under TREECOUNT in CA_SEEDLING
# TREECOUNT is for  (plot × subplot/microplot × species × inventory)
# in the CA_SEEDLING table, SUBP = microplot since seedlings are inventoried at the microplot within a subplot

# create unique plot ids for CA_SEEDLING
CA_SEEDLING_IDs <- CA_SEEDLING %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

#filter CA_SEEDLING to only keep the unique plot ids from CA_PLOT_by_COND
CA_SEEDLING_filtered <- CA_SEEDLING_IDs %>%
  filter(plot_id %in% CA_PLOT_by_COND$plot_id)

# lets look at trees in the plots with both pre and post fire 
# pull the plot_ids for status = pre_and_post from CA_prepost_fire_summary

CA_SEEDLING_prepost_fire <- CA_SEEDLING_filtered %>%
  semi_join(CA_prepost_fire_summary %>% filter(status == "pre_and_post"),
    by = "plot_id") %>%
  left_join(CA_prepost_fire_summary %>%
      filter(status == "pre_and_post") %>%
      dplyr::select(plot_id, fire_year, status),
    by = "plot_id")

# filter to post-fire plots only
CA_SEEDLING_post_fire_plots <- CA_SEEDLING_prepost_fire %>%
  filter(INVYR > fire_year)


length(unique(CA_SEEDLING_post_fire_plots$plot_id)) # 237
length(unique(CA_TREE_post_fire_plots$plot_id)) # 304
# 237 unique plots (why is this different from the 304 for the TREE analysis ??)

## To qualify for counting, conifer seedlings must be at least 6 inches tall and hardwood seedlings must be at least 12 inches tall (which is maybe why that column is blank on the table).

# how many post fire plots have TREECOUNT = NA (no seedlings recorded)
CA_SEEDLING_post_fire_plots %>%
  filter(is.na(TREECOUNT)) %>%
  summarise(n_plot_ids = n_distinct(plot_id))
# 0 / 237 have no seedlings 

# number of seedlings per plot (aggregated across SUBP/microplot)
CA_SEEDLING_per_plot <- CA_SEEDLING_post_fire_plots %>%
  group_by(plot_id, INVYR, fire_year) %>%
  summarise(
    n_seedlings = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop"
  )

hist(CA_SEEDLING_per_plot$n_seedlings, breaks=30)
summary(CA_SEEDLING_per_plot$n_seedlings)

# number of seedlings per species per plot (aggregated across SUBP/microplot)
CA_SEEDLING_species_summary <- CA_SEEDLING_post_fire_plots %>%
  group_by(plot_id, INVYR, fire_year, SPCD) %>%
  summarise(n_seedlings = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop") %>%
  left_join(REF_SPECIES %>% dplyr::select(SPCD, GENUS, SPECIES),
    by = "SPCD") %>%
  mutate(species = paste(GENUS, SPECIES, sep = "_"))

CA_SEEDLING_species_summary %>%
  group_by(plot_id) %>%
  summarise(n_species = n_distinct(SPCD)) %>%
  ggplot(aes(x = n_species)) +
  geom_histogram(binwidth = 1, fill = "darkgreen", color = "white") +
  theme_bw() +
  labs(
    x = "Number of seedling species per plot",
    y = "Number of plots"
  )

# number of seedling species per plot 
CA_SEEDLING_plot_summary <- CA_SEEDLING_post_fire_plots %>%
  group_by(plot_id, INVYR, fire_year) %>%
  summarise(
    n_seedling_species = n_distinct(SPCD),
    n_seedlings = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop"
  )

# average number of seedlings per species per plot
ggplot(
  CA_SEEDLING_species_summary %>%
    group_by(species, plot_id) %>%
    summarise(n_seedlings = sum(n_seedlings), .groups = "drop") %>%
    group_by(species) %>%
    summarise(mean_per_plot = mean(n_seedlings), .groups = "drop") %>%
    slice_max(mean_per_plot, n = 20),
  aes(
    x = reorder(species, mean_per_plot),
    y = mean_per_plot
  )
) +
  geom_col(fill = "darkgreen") +
  coord_flip() +
  scale_y_continuous(
    breaks = function(x) seq(0, max(x, na.rm = TRUE), by = 5)
  ) +
  labs(
    x = "species",
    y = "average # seedlings per plot",
    title = "average seedling abundance per species per plot"
  ) +
  theme_bw()

# how many plots have 0 species?
CA_SEEDLING_post_fire_plots %>%
  group_by(plot_id) %>%
  summarise(n_species = n_distinct(SPCD[TREECOUNT > 0])) %>%
  filter(n_species == 0)
# no plots have zero species as seedlings

# boxplots of seedling abundances per species per plot
CA_SEEDLING_species_summary %>%
  group_by(plot_id, species) %>%
  summarise(n_seedlings = sum(n_seedlings), .groups = "drop") %>%
  
  mutate(n_seedlings_capped = pmin(n_seedlings, 50)) %>%
  
  ggplot(aes(x = species, y = n_seedlings_capped)) +
  geom_boxplot(fill = "darkgreen", alpha = 0.6) +
  coord_flip() +
  labs(
    x = "species",
    y = "seedlings per plot (capped at 50)",
    title = "seedling abundance per species per plot"
  ) +
  theme_bw()

```
## expanding to include post-only
```{r}

CA_TREE_IDs <- CA_TREE %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

#filter CA_TREE to only keep the unique plot ids from CA_PLOT_by_COND (our fire only plots in our focal forest types)
CA_TREE_filtered <- CA_TREE_IDs %>%
  filter(plot_id %in% CA_PLOT_by_COND$plot_id)

CA_TREE_prepost_fire <- CA_TREE_filtered %>%
  semi_join(
    CA_prepost_fire_summary %>%
      filter(status %in% c("pre_and_post", "post_only")),
    by = "plot_id"
  ) %>%
  left_join(
    CA_prepost_fire_summary %>%
      filter(status %in% c("pre_and_post", "post_only")) %>%
      dplyr::select(plot_id, fire_year, status),
    by = "plot_id"
  )


# now we want to look at just post-fire plots (where INVYR > fire_year)
CA_TREE_post_fire_plots <- CA_TREE_prepost_fire %>%
  filter(INVYR > fire_year)

# 393 post fire plots

# number of trees per plot (post-fire plots)
CA_TREE_per_plot_post <- CA_TREE_post_fire_plots %>%
  filter(STATUSCD == 1) %>%
  mutate(which = "post") %>%
  group_by(plot_id, INVYR, fire_year, status, which) %>%
  summarise(
    n_live_tree = n_distinct(SUBP, TREE),
    .groups = "drop"
  )


# number of species with STATUSCD = 1 in post-fire plots
CA_TREE_post_fire_plot_summary <- CA_TREE_post_fire_plots %>%
  filter(STATUSCD == 1) %>%
  group_by(plot_id, INVYR, fire_year) %>%
  summarise(
    n_live_species = n_distinct(SPCD),
    n_live_trees = n_distinct(SUBP, TREE),
    .groups = "drop"
  )

# number of individuals (live) per species per plot
CA_TREE_post_fire_species_summary <- CA_TREE_post_fire_plots %>%
  filter(STATUSCD == 1) %>%
  group_by(plot_id, INVYR, fire_year, SPCD) %>%
  summarise(
    n_live_trees = n_distinct(SUBP, TREE),
    .groups = "drop"
  ) %>%
  left_join(
    REF_SPECIES %>%
      dplyr::select(SPCD, GENUS, SPECIES),
    by = "SPCD"
  ) %>%
  mutate(species = paste(GENUS, SPECIES, sep = "_"))

# is there recruitment in post-fire surveys? (pre_and_post and post_only)

CA_SEEDLING_prepost_fire <- CA_SEEDLING_filtered %>%
  semi_join(CA_prepost_fire_summary %>% filter(status %in% c("pre_and_post", "post_only")),
    by = "plot_id") %>%
  left_join(CA_prepost_fire_summary %>%
      filter(status %in% c("pre_and_post", "post_only")) %>%
      dplyr::select(plot_id, fire_year, status),
    by = "plot_id")

# filter to post-fire plots only
CA_SEEDLING_post_fire_plots <- CA_SEEDLING_prepost_fire %>%
  filter(INVYR > fire_year)

length(unique(CA_SEEDLING_post_fire_plots$plot_id)) # 313
# 313 unique plots instead of the 234 in the pre_and_post analysis above

# how many post fire plots have TREECOUNT = NA (no seedlings recorded)
CA_SEEDLING_post_fire_plots %>%
  filter(is.na(TREECOUNT)) %>%
  summarise(n_plot_ids = n_distinct(plot_id))
# 16 / 237 have no seedlings 

# number of seedlings per plot (aggregated across SUBP/microplot)
CA_SEEDLING_per_plot <- CA_SEEDLING_post_fire_plots %>%
  group_by(plot_id, INVYR, fire_year) %>%
  summarise(
    n_seedlings = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop"
  )

hist(CA_SEEDLING_per_plot$n_seedlings, breaks=30)
summary(CA_SEEDLING_per_plot$n_seedlings)

# slightly less seedlings per plot when adding ~100 more plots that are post_only 
# median = 6, mean = 17

# median is a more useful metric for seedling counts that mean because seedlings counts are highly skewed and median values give us a better approximation of the most typical plot

# number of species of seedlings per plot 
CA_SEEDLING_plot_summary <- CA_SEEDLING_post_fire_plots %>%
  group_by(plot_id, INVYR, fire_year) %>%
  summarise(
    n_seedling_species = n_distinct(SPCD),
    n_seedlings = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop"
  )
summary(CA_SEEDLING_plot_summary$n_seedling_species)
# number of individuals per species per plot
CA_SEEDLING_species_summary <- CA_SEEDLING_post_fire_plots %>%
  group_by(plot_id, INVYR, fire_year, SPCD) %>%
  summarise(n_seedlings = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop") %>%
  left_join(REF_SPECIES %>% dplyr::select(SPCD, GENUS, SPECIES),
    by = "SPCD") %>%
  mutate(species = paste(GENUS, SPECIES, sep = "_"))

summary(CA_SEEDLING_species_summary$n_seedlings)
```

## time since fire
```{r}
# since replication is low we might just need to use a bunch of plots stratified across time since fire instead of pre and post fire surveys

CA_PLOT_time_since_fire <- CA_PLOT_by_COND %>%
  distinct(CN, INVYR, PREV_PLT_CN, .keep_all = TRUE) %>%
  tidyr::pivot_longer(
    cols = c(DSTRBYR1, DSTRBYR2, DSTRBYR3),
    values_to = "fire_year"
  ) %>%
  mutate(
    fire_year = ifelse(fire_year == 9999, NA, fire_year)
  ) %>%
  filter(!is.na(fire_year)) %>%
  mutate(
    time_since_fire = INVYR - fire_year,
    pre_fire  = time_since_fire < 0,
    post_fire = time_since_fire > 0
  )

hist(CA_PLOT_time_since_fire$time_since_fire)
```

## control plots

```{r}

# full forest types
c(201, 202, 221, 222, 224, 225, 226, 241, 261, 262, 263, 264, 265, 267, 268, 270, 271, 281, 301, 304, 305, 321, 341, 342, 361, 365, 371, 722, 911, 912, 922, 923, 933, 941, 961)

# bring in plot_ids for burned plots (pre_and_post, post_only)
CA_prepost_burned_plot_id <- CA_prepost_fire_summary %>%
  filter(status %in% c("pre_and_post", "post_only")) 

# filter for plots that have DSTRBCD == 0 in all columns (DSTRBCD1-3) and all rows
CA_COND_control_summary <- CA_COND_IDs %>%
  filter(FORTYPCD %in% c(371, 201, 221, 222, 224, 225, 261, 922)) %>%
  group_by(plot_id) %>%
  summarise(
    all_zero = all(DSTRBCD1 == 0 & DSTRBCD2 == 0 & DSTRBCD3 == 0),
    .groups = "drop"
  ) %>%
  filter(all_zero) %>%
  select(plot_id)

CA_COND_treatment <- CA_COND_IDs %>%
  filter(
    plot_id %in% CA_prepost_burned_plot_id$plot_id |
      plot_id %in% CA_COND_control_summary$plot_id
  ) %>%
  mutate(
    condition = if_else(
      plot_id %in% CA_prepost_burned_plot_id$plot_id,
      "burned",
      "control"
    ),
    fire_code = case_when(
      condition == "control" ~ "0",
      DSTRBCD1 %in% c(30, 31, 32) ~ as.character(DSTRBCD1),
      DSTRBCD2 %in% c(30, 31, 32) ~ as.character(DSTRBCD2),
      DSTRBCD3 %in% c(30, 31, 32) ~ as.character(DSTRBCD3),
      TRUE ~ NA_character_
    )
  )

# Confirm unique plot counts
CA_COND_treatment %>%
  group_by(condition) %>%
  summarise(n_unique_plots = n_distinct(plot_id))

# burned               393
# control             1601

# Pull coordinates for these plots
CA_PLOT_coordinates <- CA_PLOT_IDs %>%
  semi_join(CA_COND_treatment, by = "plot_id") %>%
  distinct(plot_id, LAT, LON)


# lets pull climate for control and burned plots 

# convert lat/long points in a spatial object
CA_PLOT_coordinates_vect <- vect(
  CA_PLOT_coordinates, geom = c("LON", "LAT"), crs = "EPSG:4326")

# extract climate for all lat/long coordinates (all unique plot IDs)

names(ppt_mean) <- "ppt_mean"
names(def_mean) <- "def_mean"
names(tmin_mean) <- "tmin_mean"
names(tmax_mean) <- "tmax_mean"
names(aet_mean) <- "aet_mean"
names(pet_mean) <- "pet_mean"
names(vpd_mean) <- "vpd_mean"

terra_climate_stack <- c(
  ppt_mean,
  def_mean,
  tmin_mean,
  tmax_mean,
  aet_mean,
  pet_mean,
  vpd_mean
)


terra_climate_values <- terra::extract(
  terra_climate_stack,
  CA_PLOT_coordinates_vect
) %>%
  bind_cols(
    CA_PLOT_coordinates %>% select(plot_id, LAT, LON)
  ) %>%
  select(plot_id, LAT, LON, everything(), -ID)

terra_climate_values <- terra_climate_values %>%
  left_join(
    CA_COND_treatment %>%
      select(plot_id, condition, fire_code),
    by = "plot_id"
  )

def_df <- as.data.frame(
  def_mean,
  xy = TRUE
)

ggplot(
  terra_climate_values,
  aes(
    x = LON,
    y = LAT,
    color = def_mean
  )
) +
  geom_point(
    size = 3
  ) +
  scale_color_viridis_c(
    name = "Climatic water deficit\n(mm)"
  ) +
  coord_equal() +
  theme_classic()

# join climate back to main data frame with all plots 
#CA_PLOT_climate <- CA_PLOT_treatment %>%
  #left_join(
   # terra_climate_values,
   # by = "plot_id")

```

## climate and distance matrices
```{r}
climate_variables <- c(
  "ppt_mean",
  "def_mean",
  "tmin_mean",
  "tmax_mean",
  "aet_mean",
  "pet_mean",
  "vpd_mean"
)

# how far apart are two plots, considering 7 climate variables at once? 
# we can't calculate raw distances from these variables, it would be meaningless and skewed 

# standardizing climate variables using z-scores
# how many standard deviations above or below the average is this value?
# z = x - mean / sd

climate_standardized <- terra_climate_values %>%
  mutate(across(all_of(climate_variables), ~ as.numeric(scale(.)), .names = "z_{.col}"))

z_vars <- paste0("z_", climate_variables)
# now, each climate value is the number of standard deviations from that variable's own average

CA_burned_climate  <- climate_standardized %>% filter(condition == "burned")
CA_control_climate <- climate_standardized %>% filter(condition == "control")

# if there are any plots with NAs in lat/lon it will throw NAs for everything downstream
CA_control_climate_cleaned <- CA_control_climate %>% 
  filter(if_all(all_of(c(z_vars, "LAT", "LON")), ~ !is.na(.)))

# rdist() (fields package) computes distances between two separate matrices — every row in matrix A vs. every row in matrix B. we don't want burned-vs-burned or control-vs-control distances, we only want burned-vs-control distances 
climate_distance_matrix <- rdist(
  as.matrix(CA_burned_climate[, z_vars]),
  as.matrix(CA_control_climate_cleaned[, z_vars])
)
# assign plot_ids to climate distance matrix
dimnames(climate_distance_matrix) <- list(CA_burned_climate$plot_id, CA_control_climate_cleaned$plot_id)

# generate geographic distance matrix
# distm() (geosphere package) computes a distance matrix between two sets of geographic coordinates, using distHaversine formula, the great-circle distance that accounts for Earth's curvature
geo_dist_mat <- distm(
  as.matrix(CA_burned_climate[, c("LON", "LAT")]),
  as.matrix(CA_control_climate_cleaned[, c("LON", "LAT")]),
  fun = distHaversine
) / 1000  # convert meters to km

dimnames(geo_dist_mat) <- list(CA_burned_climate$plot_id, CA_control_climate_cleaned$plot_id)

# now that we have both climate and geographic distance matrices, we want to combine them into one score by normalizing distance matrix values from 0-1
norm01 <- function(m) (m - min(m)) / (max(m) - min(m)) # could also be done with scales::rescale() 

clim_norm <- norm01(climate_distance_matrix)
geo_norm  <- norm01(geo_dist_mat)

# weight toward climate similarity, over geographic proximity. 60% climate, 40% geographic
w_clim <- 0.6   

# combined distance across all  standardized climate variables using Euclidean distance
combined_dist <- w_clim * norm01(climate_distance_matrix) +
  (1 - w_clim) * norm01(geo_dist_mat)

# pick the top 5 best matched control plots per single burned plot
n_match <- 5

match_list <- lapply(seq_len(nrow(combined_dist)), function(i) {
  ord <- order(combined_dist[i, ])
  top <- ord[seq_len(n_match)]
  data.frame(
    burned_id = CA_burned_climate$plot_id[i],
    control_id = CA_control_climate$plot_id[top],
    combined_distance = combined_dist[i, top],
    climate_distance = climate_distance_matrix[i, top],
    geo_distance_km = geo_dist_mat[i, top],
    rank = seq_len(n_match),
    row.names = NULL
  )
})

# 'matches' now has 5 rows per burned plot, ranked by closeness (in climate and geographic space)
matches <- bind_rows(match_list)
head(matches, 10)

# lets make sure the same control plot isnt being applied to all burned plots
# how many times did this specific control plot show up as a burned plot match?
matches %>% count(control_id, sort = TRUE) %>% head(10)
# 4 of the same control plots show up as a match for 10-13 burned plots
# most popular control plot was only used for about 3% of all burned plots

# lets check if those burned plots all have the same rankings for the popular control plot
matches %>% 
  filter(control_id == "6_2_93_54625") %>%
  select(burned_id, control_id, rank, combined_distance)

# filter control plot dataframe with the top 5 matched control plots per burned plot
matched_controls <- CA_control_climate_cleaned %>% 
  filter(plot_id %in% matches$control_id)

ggplot(matches, aes(x = geo_distance_km, y = climate_distance)) +
  geom_point(alpha = 0.4) +
  labs(
    x = "Geographic distance (km)",
    y = "Climate distance (standardized)",
    title = "Climate distance vs. geographic distance for matched pairs"
  ) +
  theme_minimal()


# Compare means for one variable, e.g. tmax_mean
mean(CA_burned_climate$tmax_mean)
mean(matched_controls$tmax_mean)          # should be closer to burned mean
mean(CA_control_climate_cleaned$tmax_mean) # than the full unmatched pool 

# Or for all 7 at once
sapply(climate_variables, function(v) {
  c(burned_mean = mean(CA_burned_climate[[v]]),
    matched_mean = mean(matched_controls[[v]]),
    full_pool_mean = mean(CA_control_climate_cleaned[[v]]))
})

```

## without replacement
```{r}

without_replacement <- function(dist_mat, clim_mat, geo_mat, n_match = 5) {
  
  # Build a long list of every possible (burned, control) pair with its distances
  triples <- expand.grid(
    burned_id  = rownames(dist_mat),
    control_id = colnames(dist_mat),
    stringsAsFactors = FALSE
  )
  triples$combined_distance <- as.vector(dist_mat)
  triples$climate_distance  <- as.vector(clim_mat)
  triples$geo_distance_km   <- as.vector(geo_mat)
  
  # Sort so the best (smallest distance) pairs come first
  triples <- triples[order(triples$combined_distance), ]
  
  needed   <- setNames(rep(n_match, nrow(dist_mat)), rownames(dist_mat))
  used_ctl <- setNames(rep(FALSE, ncol(dist_mat)), colnames(dist_mat))
  
  results <- vector("list", nrow(dist_mat) * n_match)
  counter <- 1
  
  for (k in seq_len(nrow(triples))) {
    b <- triples$burned_id[k]
    c <- triples$control_id[k]
    
    if (needed[b] > 0 && !used_ctl[c]) {
      results[[counter]] <- data.frame(
        burned_id         = b,
        control_id        = c,
        combined_distance = triples$combined_distance[k],
        climate_distance  = triples$climate_distance[k],
        geo_distance_km   = triples$geo_distance_km[k],
        rank              = n_match - needed[b] + 1,
        row.names         = NULL
      )
      counter <- counter + 1
      needed[b]   <- needed[b] - 1
      used_ctl[c] <- TRUE
    }
    
    if (all(needed == 0)) break
  }
  
  bind_rows(results)
}

matches_no_replacement <- without_replacement(
  dist_mat = combined_dist,
  clim_mat = climate_distance_matrix,
  geo_mat  = geo_dist_mat,
  n_match  = 3
)

matches_no_replacement %>% count(burned_id) %>% filter(n < 5)
```


## species overlap
```{r}
# now that we have our control and burned plot pairs, lets look into species overlap between each pair

CA_burned_and_control_plot_id <- unique(c(matches_no_replacement$burned_id, matches_no_replacement$control_id))

#1572 plots total
#393 burned
#1179 control

# lets filter CA_TREE and CA_SEEDLING for our burned and control plots
CA_TREE_prepost_fire

CA_TREE_IDs_filtered <- CA_TREE_IDs %>%
  filter(plot_id %in% CA_burned_and_control_plot_id)

CA_SEEDLING_IDs_filtered <- CA_SEEDLING_IDs %>%
  filter(plot_id %in% CA_burned_and_control_plot_id)

# add species names to species codes using REF_SPECIES

CA_TREE_species <- CA_TREE_IDs_filtered %>%
  left_join(
    REF_SPECIES %>%
      mutate(species = paste(GENUS, SPECIES, sep = "_")) %>%
      select(SPCD, species),
    by = "SPCD"
  ) %>%
  left_join(
    CA_COND_treatment %>% distinct(plot_id, condition),
    by = "plot_id"
  ) %>%
  left_join(
    CA_prepost_fire_summary %>% distinct(plot_id, fire_year),
    by = "plot_id"
  )

CA_TREE_species_filtered <- CA_TREE_species %>%
  filter(
    condition == "burned" | condition == "control"
  )
#Compute total abundance per species per plot 
CA_species_abundance_per_plot <- CA_TREE_species_filtered %>%
  group_by(plot_id, species) %>%
  summarise(
    abundance = n(),
    .groups = "drop"
  )

#Compute frequency (commonness) per species — count the number of distinct plots each species appears in, divide by total number of plots.

# abundance by species for all plots (less useful?) but frequency across plots is useful (number of times a species is present in a plot relative to the total number of plots)
species_summary_combined <- CA_TREE_species_filtered %>%
  group_by(species) %>%
  summarise(
    total_abundance = n(),
    n_plots = n_distinct(plot_id),
    .groups = "drop"
  ) %>%
  mutate(
    frequency = n_plots / n_distinct(CA_TREE_species_filtered$plot_id)
  ) %>%
  arrange(desc(total_abundance))

# total abundance and frequency by condition (also maybe not particularly helpful)
species_summary_by_condition <- CA_TREE_species_filtered %>%
  group_by(condition, species) %>%
  summarise(
    total_abundance = n(),
    n_plots = n_distinct(plot_id),
    .groups = "drop"
  ) %>%
  group_by(condition) %>%
  mutate(
    frequency = n_plots / n_distinct(CA_TREE_species_filtered$plot_id[CA_TREE_species_filtered$condition == condition])
  ) %>%
  ungroup() %>%
  arrange(condition, desc(total_abundance))


# Rank species on each metric separately, then combine the two ranks into one overall score (e.g., average rank, so a species that's #3 in abundance and #10 in frequency gets averaged into a combined rank).


#Sort by the combined rank and take the top 75.

top_75_species <- species_summary_combined %>%
  mutate(
    abundance_rank = rank(-total_abundance),
    frequency_rank = rank(-frequency),
    combined_rank = (abundance_rank + frequency_rank) / 2
  ) %>%
  arrange(combined_rank) %>%
  slice_head(n = 75)

#total number of species across both control and burned plots
CA_TREE_species_filtered %>%
  group_by(condition) %>%
  summarise(n_species = n_distinct(species), .groups = "drop")

CA_all_species <- CA_TREE_species_filtered %>%
  distinct(species) %>%
  pull(species)

length(CA_all_species)
# 67 total species




# to calculate dominance (the number of indiviudals of that species in that plot) and frequency (commonness)  (the number of plots the species occurs in / total number of plots)

CA_tree_count <- CA_TREE_species_filtered %>%
  group_by(plot_id) %>%
  summarise(total_trees_in_plot = n(), .groups = "drop")

CA_species_dominance <- CA_TREE_species_filtered %>%
  group_by(plot_id, species) %>%
  summarise(species_count = n(), .groups = "drop") %>%
  left_join(CA_tree_count, by = "plot_id") %>%
  mutate(relative_abundance = species_count / total_trees_in_plot)

CA_species_summary <- CA_species_dominance %>%
  group_by(species) %>%
  summarise(
    n_plots_present  = n_distinct(plot_id),
    avg_dominance    = mean(relative_abundance),
    median_dominance = median(relative_abundance),
    min_dominance    = min(relative_abundance),
    max_dominance    = max(relative_abundance),
    .groups = "drop"
  ) %>%
  mutate(frequency = n_plots_present / n_distinct(CA_TREE_species_filtered$plot_id)) %>%
  arrange(desc(frequency))

# lets check whether dominant and common species differ by condition (burned vs control)

CA_species_dominance_condition <- CA_species_dominance %>%
  left_join(
    CA_TREE_species_filtered %>% distinct(plot_id, condition),
    by = "plot_id"
  )

# now summarize by condition + species instead of species alone
CA_species_summary_by_condition <- CA_species_dominance_condition %>%
  group_by(condition, species) %>%
  summarise(
    n_plots_present  = n_distinct(plot_id),
    avg_dominance    = mean(relative_abundance),
    median_dominance = median(relative_abundance),
    min_dominance    = min(relative_abundance),
    max_dominance    = max(relative_abundance),
    .groups = "drop"
  ) %>%
  group_by(condition) %>%
  mutate(
    frequency = n_plots_present / n_distinct(CA_TREE_species_filtered$plot_id[CA_TREE_species_filtered$condition == condition])
  ) %>%
  ungroup() %>%
  arrange(condition, desc(frequency))

CA_species_condition_comparison <- CA_species_summary_by_condition %>%
  select(condition, species, frequency, median_dominance) %>%
  pivot_wider(
    names_from = condition,
    values_from = c(frequency, median_dominance)
  ) %>%
  mutate(
    frequency_diff = frequency_burned - frequency_control,
    dominance_diff = median_dominance_burned - median_dominance_control
  ) %>%
  arrange(desc(abs(dominance_diff)))

library(ggrepel)

CA_species_condition_comparison %>%
  filter(!is.na(dominance_diff), !is.na(frequency_diff)) %>%
  ggplot(aes(x = frequency_diff, y = dominance_diff, label = species)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray70") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray70") +
  geom_point(alpha = 0.6, size = 2) +
  geom_text_repel(size = 3, max.overlaps = 15) +
  labs(
    x = "Frequency difference (burned − control)",
    y = "Dominance difference (burned − control)",
    title = "Burned - control species dominance (median)"
  ) +
  theme_minimal()


#  is the number of species recruiting in control plots is greater than burned

CA_SEEDLING_condition_filtered <- CA_SEEDLING_IDs %>%
  filter(plot_id %in% CA_burned_and_control_plot_id) %>%
  left_join(
    CA_COND_treatment %>% distinct(plot_id, condition),
    by = "plot_id"
  )

CA_SEEDLING_species_summary_condition <- CA_SEEDLING_condition_filtered %>%
  group_by(plot_id, condition, SPCD) %>%
  summarise(n_seedlings = sum(TREECOUNT, na.rm = TRUE), .groups = "drop") %>%
  left_join(REF_SPECIES %>% dplyr::select(SPCD, GENUS, SPECIES), by = "SPCD") %>%
  mutate(species = paste(GENUS, SPECIES, sep = "_"))

CA_SEEDLING_plot_summary_condition <- CA_SEEDLING_species_summary_condition %>%
  group_by(plot_id, condition) %>%
  summarise(
    n_seedling_species = n_distinct(SPCD),
    n_seedlings = sum(n_seedlings, na.rm = TRUE),
    .groups = "drop"
  )

CA_SEEDLING_plot_summary_condition %>%
  ggplot(aes(x = n_seedling_species, fill = condition)) +
  geom_histogram(binwidth = 1, color = "white", position = "identity", alpha = 0.6) +
  facet_wrap(~condition, ncol = 1) +
  theme_bw() +
  labs(
    x = "Number of seedling species per plot",
    y = "Number of plots",
    title = "Seedling species richness: burned vs. control"
  )

CA_SEEDLING_plot_summary_condition %>%
  ggplot(aes(x = condition, y = n_seedling_species, fill = condition)) +
  geom_boxplot(alpha = 0.6) +
  theme_bw() +
  labs(
    x = NULL,
    y = "Number of seedling species per plot",
    title = "Seedling species richness: burned vs. control"
  )

CA_seedling_richness_summary_by_condition <- CA_SEEDLING_plot_summary_condition %>%
  group_by(condition) %>%
  summarise(
    mean_richness   = mean(n_seedling_species),
    median_richness = median(n_seedling_species),
    n_plots         = n(),
    .groups = "drop"
  )

CA_seedling_richness_summary_by_condition

CA_SEEDLING_species_summary_condition %>%
  group_by(species, condition, plot_id) %>%
  summarise(n_seedlings = sum(n_seedlings), .groups = "drop") %>%
  group_by(species, condition) %>%
  summarise(mean_per_plot = mean(n_seedlings), .groups = "drop") %>%
  group_by(condition) %>%
  slice_max(mean_per_plot, n = 20) %>%
  ungroup() %>%
  ggplot(aes(x = reorder_within(species, mean_per_plot, condition), y = mean_per_plot, fill = condition)) +
  geom_col() +
  coord_flip() +
  scale_x_reordered() +
  facet_wrap(~condition, scales = "free_y") +
  labs(
    x = "species",
    y = "average # seedlings per plot",
    title = "Average seedling abundance per species per plot: burned vs. control"
  ) +
  theme_bw()

CA_SEEDLING_species_summary_condition %>%
  group_by(plot_id, condition, species) %>%
  summarise(n_seedlings = sum(n_seedlings), .groups = "drop") %>%
  mutate(n_seedlings_capped = pmin(n_seedlings, 50)) %>%
  ggplot(aes(x = species, y = n_seedlings_capped, fill = condition)) +
  geom_boxplot(alpha = 0.6) +
  coord_flip() +
  facet_wrap(~condition, scales = "free_y") +
  labs(
    x = "species",
    y = "seedlings per plot (capped at 50)",
    title = "Seedling abundance per species per plot: burned vs. control"
  ) +
  theme_bw()

summary(CA_SEEDLING_plot_summary_condition$n_seedling_species[CA_SEEDLING_plot_summary_condition$condition == "burned"])

summary(CA_SEEDLING_plot_summary_condition$n_seedling_species[CA_SEEDLING_plot_summary_condition$condition == "control"])

CA_SEEDLING_plot_summary_condition %>%
  ggplot(aes(x = n_seedling_species, fill = condition)) +
  geom_histogram(binwidth = 1, color = "white", position = "identity", alpha = 0.6) +
  facet_wrap(~condition, ncol = 1) +
  theme_bw() +
  labs(
    x = "Number of seedling species per plot",
    y = "Number of plots",
    title = "Seedling species richness: burned vs. control"
  )

```

## whole plot network
```{r}

# filter CA_COND_IDs for forest types only, no disturbance filtering
CA_COND_forest_type <- CA_COND_IDs %>%
  filter(FORTYPCD %in% c(201, 202, 221, 222, 224, 225, 226, 241, 261, 262, 263, 264, 265, 267, 268, 270, 271, 281, 301, 304, 305, 321, 341, 342, 361, 365, 371, 722, 911, 912, 922, 923, 933, 941, 961))
           
(371, 201, 221, 222, 224, 225, 261, 922))



# get the unique plot_ids that qualify
CA_forest_type_plot_id <- CA_COND_forest_type %>%
  distinct(plot_id) %>%
  pull(plot_id)

length(CA_forest_type_plot_id)

# filter CA_PLOT_IDs for those plot_ids
CA_PLOT_forest_type <- CA_PLOT_IDs %>%
  filter(plot_id %in% CA_forest_type_plot_id)

# filter CA_TREE_IDs for those plot_ids
CA_TREE_forest_type <- CA_TREE_IDs %>%
  filter(plot_id %in% CA_forest_type_plot_id)

nrow(CA_PLOT_forest_type)
n_distinct(CA_TREE_forest_type$plot_id)

CA_TREE_forest_type_species <- CA_TREE_forest_type %>%
  left_join(
    REF_SPECIES %>%
      mutate(species = paste(GENUS, SPECIES, sep = "_")) %>%
      select(SPCD, species),
    by = "SPCD"
  )
sum(is.na(CA_TREE_forest_type_species$species))
n_distinct(CA_TREE_forest_type_species$species)

# how different are the species from this workflow and the previous control plot filtering by climate?

# species from your burned/control workflow (already built)
CA_all_species   # from earlier: distinct species across CA_TREE_species_filtered

# species from your new forest-type-only workflow
CA_forest_type_all_species <- CA_TREE_forest_type_species %>%
  distinct(species) %>%
  pull(species)

length(CA_forest_type_all_species)

# species found in both workflows
intersect(CA_all_species, CA_forest_type_all_species)

# species found only in the burned/control workflow
setdiff(CA_all_species, CA_forest_type_all_species)

# species found only in the forest-type-only workflow
setdiff(CA_forest_type_all_species, CA_all_species)


# make sure the species coming out of SEEDLING match TREE
# filter CA_SEEDLING_IDs for those plot_ids
CA_SEEDLING_forest_type <- CA_SEEDLING_IDs %>%
  filter(plot_id %in% CA_forest_type_plot_id)

CA_SEEDLING_forest_type_species <- CA_SEEDLING_forest_type %>%
  left_join(
    REF_SPECIES %>%
      mutate(species = paste(GENUS, SPECIES, sep = "_")) %>%
      select(SPCD, species),
    by = "SPCD"
  )

# species list from seedling data (forest-type-only workflow)
CA_forest_type_all_seedling_species <- CA_SEEDLING_forest_type_species %>%
  distinct(species) %>%
  pull(species)

length(CA_forest_type_all_species)
length(CA_forest_type_all_seedling_species)

# species found in both tree and seedling data
intersect(CA_forest_type_all_species, CA_forest_type_all_seedling_species)

# species found only in tree data (mature trees, no seedlings recorded)
setdiff(CA_forest_type_all_species, CA_forest_type_all_seedling_species)

# species found only in seedling data (recruiting, but no mature trees recorded)
setdiff(CA_forest_type_all_seedling_species, CA_forest_type_all_species)

# create a shared species list
CA_forest_type_all_species
CA_forest_type_all_seedling_species

CA_species <- data.frame(
  species = unique(c(
    CA_forest_type_all_species,
    CA_forest_type_all_seedling_species
  ))
)

```


# expand states
# OR
### fire only plots
```{r}
#create a unique PLOT ID in COND and in PLOT tables 
OR_COND_IDs <- OR_COND %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )
OR_PLOT_IDs <- OR_PLOT %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

# collapse information from multiple CONDIDs associated with each plot to avoid many-to-many mapping
# and then filter by fire and forest type  

#plot_id → multiple CONDIDs → each CONDID has up to 3 disturbance codes and years
# one row for a specific plot_id, CONID, and disturbance event (combining DSTRBCD and DSTRBYR)
OR_COND_long <- OR_COND_IDs %>%
  dplyr::select(plot_id, CONDID, FORTYPCD,
    DSTRBCD1, DSTRBCD2, DSTRBCD3,
    DSTRBYR1, DSTRBYR2, DSTRBYR3) %>%
  pivot_longer(
    cols = starts_with("DSTR"),
    names_to = c(".value", "event"),
    names_pattern = "(DSTRBCD|DSTRBYR)([1-3])"
  ) %>%
  filter(DSTRBCD %in% c(30, 31, 32)) %>%
  filter(FORTYPCD %in% c(371, 201, 221, 222, 224, 225, 261, 922)) %>%
  mutate(
    fire_year_severity = paste(DSTRBYR, DSTRBCD, sep = "_")
  )

# summarize fire history across rows for unique plot_id (one row per plot_id)
OR_fire_summary <- OR_COND_long %>%
  group_by(plot_id) %>%
  summarise(
    n_fire_events = n_distinct(DSTRBYR), # number of distinct fire years per plot
    
    # this just collapses all codes into a list but not necessarily relative to year
    fire_types = paste(sort(unique(DSTRBCD)), collapse = ", "), 
    fire_years = paste(sort(unique(DSTRBYR)), collapse = ", "),
    
    fire_year_severity =
    paste(unique(fire_year_severity), collapse = ", "),
    
    multiple_fires = n_distinct(DSTRBYR) > 1, # did fire occur more than once?
    multiple_severity = n_distinct(DSTRBCD) > 1, # are there multiple fire types?
    first_fire_year = min(DSTRBYR, na.rm = TRUE),
    last_fire_year = max(DSTRBYR, na.rm = TRUE),
    .groups = "drop"
  )

# how many plots burned multiple times?
OR_fire_summary %>% summarise(n_plots_multiple_fires = sum(multiple_fires))
# 20 plots

# how many plots that burned multipe times have different severities?
OR_fire_summary %>% filter(multiple_fires) %>% summarise(n_plots_multiple_fire_and_severity = sum(multiple_severity))
# 11 plots

# how many plots that burned once have multiple severities within a plot?
OR_fire_summary %>%
  filter(!multiple_fires) %>%
  summarise(
    n_plots_single_fire_multiple_severity =
      sum(multiple_severity))
# 9 plots

OR_burned_PLOT_IDs <- OR_fire_summary %>% 
  distinct(plot_id)

# merge COND and PLOT by this unique ID 
OR_PLOT_by_COND <- OR_PLOT_IDs %>%
  semi_join(OR_burned_PLOT_IDs, by = "plot_id") %>%  #keep only plot_ids with at least one fire 
  left_join(OR_fire_summary, by = "plot_id")  #add fire summary to each plot, should be one-to-many mapping

#for each unique plot, count how many unique inventory years (INVYR) it has
OR_PLOT_by_COND %>%
  group_by(plot_id) %>%
  summarise(
    n_revisits = n_distinct(INVYR),
    .groups = "drop"
  ) %>%
  count(n_revisits)

# n=1 (23)
# n=2 (485)
# n=3 (87)

# how many plots have pre and post fire surveys
OR_fire_events <- OR_COND_long %>%
  dplyr::select(plot_id, DSTRBYR) %>%
  distinct() %>%
  rename(fire_year = DSTRBYR)

# does each unique plot_id have a pre and post fire survey around any/all fire years?
OR_prepost_fire_summary <- OR_PLOT_by_COND %>%
  distinct(plot_id, INVYR) %>%
  inner_join(
    OR_fire_events %>% distinct(plot_id, fire_year),
    by = "plot_id"
  ) %>%
  mutate(
    relationship = case_when(
      INVYR < fire_year ~ "pre",
      INVYR > fire_year ~ "post",
      TRUE ~ "same_year"
    )
  ) %>%
  group_by(plot_id, fire_year) %>%
  summarise(
    has_pre  = any(relationship == "pre"),
    has_post = any(relationship == "post"),
    n_pre_surveys  = sum(relationship == "pre"),
    n_post_surveys = sum(relationship == "post"),
    .groups = "drop"
  ) %>%
  mutate(
      status = case_when(
      has_pre & has_post ~ "pre_and_post",
      has_pre ~ "pre_only",
      has_post ~ "post_only",
      TRUE ~ "no_clear_separation"
    )
  )

OR_prepost_fire_summary %>%
  count(status)
# no clear separation = 5 
# post only = 232
# pre only = 44
#pre and post = 334


# how many plots are there in CA
OR_PLOT %>%
  distinct(LAT, LON) %>%
  nrow()
# 22411

length(unique(OR_COND_IDs$plot_id)) # 22418
length(unique(OR_PLOT_IDs$plot_id)) # 22418


```

## tree
```{r}
#Tree number. A number used to uniquely identify a tree on a subplot. Tree numbers can be used to track trees when PLOT.DESIGNCD or DESIGNCD is the same between inventories

# before looking into tree level data, just confirm that the design codes are the same across plots in our CA_PLOT_by_COND table that has our burned plots
n_distinct(OR_PLOT_by_COND$DESIGNCD) # two design codes across the filtered plots
unique(OR_PLOT_by_COND$DESIGNCD) # 501, 502

# do we have surviving trees in post-fire surveys?
# create unique plot ids for CA_TREE
OR_TREE_IDs <- OR_TREE %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

#filter CA_TREE to only keep the unique plot ids from CA_PLOT_by_COND (our fire only plots in our focal forest types)
OR_TREE_filtered <- OR_TREE_IDs %>%
  filter(plot_id %in% OR_PLOT_by_COND$plot_id)

# to locate a specific tree ID, use the following variables 
# STATECD, INVYR, UNITCD, COUNTYCD, PLOT, SUBP, TREE
# TREE is unique to SUBP not plot_id
# we can look for unique combinations of SUBP and TREE within a plot_id (STATECD, UNITCD, COUNTYCD, PLOT)

# is TREE our unique identifier of TREE ?
OR_TREE_filtered %>%
  group_by(TREE) %>%
  summarise(
    n_plots = n_distinct(plot_id),
    .groups = "drop"
  ) %>%
  filter(n_plots > 1) %>%
  summarise(
    n_trees_multiple_plots = n()
  )
# 303 trees are found in multiple plots 

OR_TREE_filtered %>%
  summarise(
    n_unique_trees = n_distinct(plot_id, SUBP, TREE)
  )
# number of unique trees 20470

# Tree number notes:
# TREE uniquely identifies a tree within a subplot (SUBP),
# so full tree identity is: plot_id + SUBP + TREE

# unique tree identity is plot_id + SUBP + TREE
# make a unique tree identifier too? or use plot_id + SUBP + TREE everytime
#tree_id = paste(plot_id, SUBP, TREE, sep = "_") 

# lets look at trees in the plots with both pre and post fire surveys
# pull the plot_ids for status = pre_and_post from CA_prepost_fire_summary and the fire year the post-fire plot is associated with

OR_TREE_prepost_fire <- OR_TREE_filtered %>%
  semi_join(
    OR_prepost_fire_summary %>% filter(status == "pre_and_post"),
    by = "plot_id"
  ) %>%
  left_join(
    OR_prepost_fire_summary %>%
      filter(status == "pre_and_post") %>%
      dplyr::select(plot_id, fire_year, status),
    by = "plot_id"
  )

OR_prepost_fire_summary %>%
  filter(status == "pre_and_post") %>%
  summarise(n_plots = n_distinct(plot_id)) #330
length(unique(OR_TREE_prepost_fire$plot_id)) #329


# now we want to look at just post-fire plots (where INVYR > fire_year)
OR_TREE_post_fire_plots <- OR_TREE_prepost_fire %>%
  filter(INVYR > fire_year)

# number of trees per plot (post-fire plots)
OR_TREE_per_plot_post <- OR_TREE_post_fire_plots %>%
  filter(STATUSCD == 1) %>%
  mutate(which = "post") %>%
  group_by(plot_id, INVYR, fire_year, status, which) %>%
  summarise(
    n_live_tree = n_distinct(SUBP, TREE),
    .groups = "drop"
  )
summary(OR_TREE_per_plot_post$n_live_tree)

# number of species with STATUSCD = 1 in post-fire plots
OR_TREE_post_fire_plot_summary <- OR_TREE_post_fire_plots %>%
  filter(STATUSCD == 1) %>%
  group_by(plot_id, INVYR, fire_year) %>%
  summarise(
    n_live_species = n_distinct(SPCD),
    n_live_trees = n_distinct(SUBP, TREE),
    .groups = "drop"
  )
summary(OR_TREE_post_fire_plot_summary$n_live_species)

# number of individuals (live) per species per plot
OR_TREE_post_fire_species_summary <- OR_TREE_post_fire_plots %>%
  filter(STATUSCD == 1) %>%
  group_by(plot_id, INVYR, fire_year, SPCD) %>%
  summarise(
    n_live_trees = n_distinct(SUBP, TREE),
    .groups = "drop"
  ) %>%
  left_join(
    REF_SPECIES %>%
      dplyr::select(SPCD, GENUS, SPECIES),
    by = "SPCD"
  ) %>%
  mutate(species = paste(GENUS, SPECIES, sep = "_"))
summary(OR_TREE_post_fire_species_summary$n_live_trees)
## tree transitions and mortality rates

# closest PRE-fire observation for each TREE
OR_TREE_pre_fire_trees <- OR_TREE_prepost_fire %>%
  filter(INVYR < fire_year) %>%
  group_by(plot_id, SUBP, TREE, fire_year) %>%
  arrange(desc(INVYR), .by_group = TRUE) %>%
  slice(1) %>%
  ungroup() %>%
  rename(
    pre_INVYR = INVYR,
    pre_STATUSCD = STATUSCD
  )

# closest POST-fire observation for each TREE
OR_TREE_post_fire_trees <- OR_TREE_prepost_fire %>%
  filter(INVYR > fire_year) %>%
  group_by(plot_id, SUBP, TREE, fire_year) %>%
  arrange(INVYR, .by_group = TRUE) %>%
  slice(1) %>%
  ungroup() %>%
  rename(
    post_INVYR = INVYR,
    post_STATUSCD = STATUSCD
  )

# join pre and post observations
OR_TREE_transitions <- OR_TREE_pre_fire_trees %>%
  inner_join(
    OR_TREE_post_fire_trees,
    by = c("plot_id", "SUBP", "TREE", "fire_year"),
    suffix = c("_pre", "_post")
  )

# mortality indicators
OR_TREE_transitions <- OR_TREE_transitions %>%
  mutate(
    alive_pre = pre_STATUSCD == 1,
    dead_post = pre_STATUSCD == 1 & post_STATUSCD == 2
  )

# plot-level mortality
OR_TREE_mortality <- OR_TREE_transitions %>%
  group_by(plot_id, fire_year) %>%
  summarise(
    n_alive_pre = sum(alive_pre, na.rm = TRUE),
    n_dead_post = sum(dead_post, na.rm = TRUE),
    mortality_rate = n_dead_post / n_alive_pre,
    .groups = "drop"
  )

summary(OR_TREE_mortality$mortality_rate)
hist(OR_TREE_mortality$mortality_rate)

# check for any new trees in post fire surveys

# trees appearing post-fire but absent pre-fire
OR_TREE_new_trees <- OR_TREE_post_fire_trees %>%
  anti_join(
    OR_TREE_transitions %>%
      distinct(plot_id, SUBP, fire_year, TREE),
    by = c("plot_id", "SUBP", "fire_year", "TREE")
  )

OR_TREE_new_trees %>%
  summarise(
    total_new_recruits = n()
  )
# 1351 new post-fire recruits in CA_TREE

OR_TREE_new_trees %>%
  summarise(
    total_new_recruits = n_distinct(plot_id, SUBP, TREE, fire_year)
  )

# recruits per plot
OR_TREE_new_recruits_per_plot <- OR_TREE_new_trees %>%
  group_by(plot_id, fire_year, post_INVYR) %>%
  summarise(
    n_new_recruits = n_distinct(SUBP, TREE),
    .groups = "drop"
  )

summary(OR_TREE_new_recruits_per_plot$n_new_recruits)

# proportion of new recruits per plot
OR_TREE_recruitment <- OR_TREE_new_recruits_per_plot %>%
  left_join(
    OR_TREE_per_plot_post %>%
      dplyr::select(
        plot_id,
        fire_year,
        post_INVYR = INVYR,
        n_live_tree
      ),
    by = c("plot_id", "fire_year", "post_INVYR")
  ) %>%
  mutate(
    recruit_prop = n_new_recruits / n_live_tree
  )

OR_TREE_recruitment %>%
  filter(recruit_prop == 1)

# 10 plots have only new trees

summary(OR_TREE_recruitment$recruit_prop)
```

## seedling
```{r}
# is there recruitment in post-fire surveys?

# seedling counts are under TREECOUNT in CA_SEEDLING
# TREECOUNT is for  (plot × subplot/microplot × species × inventory)
# in the CA_SEEDLING table, SUBP = microplot since seedlings are inventoried at the microplot within a subplot

# create unique plot ids for CA_SEEDLING
OR_SEEDLING_IDs <- OR_SEEDLING %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

#filter CA_SEEDLING to only keep the unique plot ids from CA_PLOT_by_COND
OR_SEEDLING_filtered <- OR_SEEDLING_IDs %>%
  filter(plot_id %in% OR_PLOT_by_COND$plot_id)

# lets look at trees in the plots with both pre and post fire 
# pull the plot_ids for status = pre_and_post from CA_prepost_fire_summary

OR_SEEDLING_prepost_fire <- OR_SEEDLING_filtered %>%
  semi_join(OR_prepost_fire_summary %>% filter(status == "pre_and_post"),
    by = "plot_id") %>%
  left_join(OR_prepost_fire_summary %>%
      filter(status == "pre_and_post") %>%
      dplyr::select(plot_id, fire_year, status),
    by = "plot_id")

# filter to post-fire plots only
OR_SEEDLING_post_fire_plots <- OR_SEEDLING_prepost_fire %>%
  filter(INVYR > fire_year)


length(unique(OR_SEEDLING_post_fire_plots$plot_id)) # 203
length(unique(OR_TREE_post_fire_plots$plot_id)) # 329
# 203 unique plots (why is this different from the 329 for the TREE analysis ??)

## To qualify for counting, conifer seedlings must be at least 6 inches tall and hardwood seedlings must be at least 12 inches tall (which is maybe why that column is blank on the table).

# how many post fire plots have TREECOUNT = NA (no seedlings recorded)
OR_SEEDLING_post_fire_plots %>%
  filter(is.na(TREECOUNT)) %>%
  summarise(n_plot_ids = n_distinct(plot_id))
# 0 / 203 have no seedlings 

# number of seedlings per plot (aggregated across SUBP/microplot)
OR_SEEDLING_per_plot <- OR_SEEDLING_post_fire_plots %>%
  group_by(plot_id, INVYR, fire_year) %>%
  summarise(
    n_seedlings = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop"
  )

hist(OR_SEEDLING_per_plot$n_seedlings, breaks=30)
summary(OR_SEEDLING_per_plot$n_seedlings)

# number of seedlings per species per plot (aggregated across SUBP/microplot)
OR_SEEDLING_species_summary <- OR_SEEDLING_post_fire_plots %>%
  group_by(plot_id, INVYR, fire_year, SPCD) %>%
  summarise(n_seedlings = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop") %>%
  left_join(REF_SPECIES %>% dplyr::select(SPCD, GENUS, SPECIES),
    by = "SPCD") %>%
  mutate(species = paste(GENUS, SPECIES, sep = "_"))

OR_SEEDLING_species_summary %>%
  group_by(plot_id) %>%
  summarise(n_species = n_distinct(SPCD)) %>%
  ggplot(aes(x = n_species)) +
  geom_histogram(binwidth = 1, fill = "darkgreen", color = "white") +
  theme_bw() +
  labs(
    x = "Number of seedling species per plot",
    y = "Number of plots"
  )

# number of seedling species per plot 
OR_SEEDLING_plot_summary <- OR_SEEDLING_post_fire_plots %>%
  group_by(plot_id, INVYR, fire_year) %>%
  summarise(
    n_seedling_species = n_distinct(SPCD),
    n_seedlings = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop"
  )



```
## expanding to include post-only
```{r}

OR_TREE_IDs <- OR_TREE %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

#filter CA_TREE to only keep the unique plot ids from CA_PLOT_by_COND (our fire only plots in our focal forest types)
OR_TREE_filtered <- OR_TREE_IDs %>%
  filter(plot_id %in% OR_PLOT_by_COND$plot_id)

OR_TREE_prepost_fire <- OR_TREE_filtered %>%
  semi_join(
    OR_prepost_fire_summary %>%
      filter(status %in% c("pre_and_post", "post_only")),
    by = "plot_id"
  ) %>%
  left_join(
    OR_prepost_fire_summary %>%
      filter(status %in% c("pre_and_post", "post_only")) %>%
      dplyr::select(plot_id, fire_year, status),
    by = "plot_id"
  )


# now we want to look at just post-fire plots (where INVYR > fire_year)
OR_TREE_post_fire_plots <- OR_TREE_prepost_fire %>%
  filter(INVYR > fire_year)

# 551 post fire plots

# number of living trees per plot (post-fire plots)
OR_TREE_per_plot_post <- OR_TREE_post_fire_plots %>%
  filter(STATUSCD == 1) %>%
  mutate(which = "post") %>%
  group_by(plot_id, INVYR, fire_year, status, which) %>%
  summarise(
    n_live_tree = n_distinct(SUBP, TREE),
    .groups = "drop"
  )
summary(OR_TREE_per_plot_post$n_live_tree)

# number of species with STATUSCD = 1 in post-fire plots
OR_TREE_post_fire_plot_summary <- OR_TREE_post_fire_plots %>%
  filter(STATUSCD == 1) %>%
  group_by(plot_id, INVYR, fire_year) %>%
  summarise(
    n_live_species = n_distinct(SPCD),
    n_live_trees = n_distinct(SUBP, TREE),
    .groups = "drop"
  )
summary(OR_TREE_post_fire_plot_summary$n_live_species)

# number of individuals (live) per species per plot
OR_TREE_post_fire_species_summary <- OR_TREE_post_fire_plots %>%
  filter(STATUSCD == 1) %>%
  group_by(plot_id, INVYR, fire_year, SPCD) %>%
  summarise(
    n_live_trees = n_distinct(SUBP, TREE),
    .groups = "drop"
  ) %>%
  left_join(
    REF_SPECIES %>%
      dplyr::select(SPCD, GENUS, SPECIES),
    by = "SPCD"
  ) %>%
  mutate(species = paste(GENUS, SPECIES, sep = "_"))
summary(OR_TREE_post_fire_species_summary$n_live_trees)

# is there recruitment in post-fire surveys? (pre_and_post and post_only)

OR_SEEDLING_prepost_fire <- OR_SEEDLING_filtered %>%
  semi_join(OR_prepost_fire_summary %>% filter(status %in% c("pre_and_post", "post_only")),
    by = "plot_id") %>%
  left_join(OR_prepost_fire_summary %>%
      filter(status %in% c("pre_and_post", "post_only")) %>%
      dplyr::select(plot_id, fire_year, status),
    by = "plot_id")

# filter to post-fire plots only
OR_SEEDLING_post_fire_plots <- OR_SEEDLING_prepost_fire %>%
  filter(INVYR > fire_year)

length(unique(OR_SEEDLING_post_fire_plots$plot_id)) # 381
# 381 unique plots instead of the 203 in the pre_and_post analysis above

# how many post fire plots have TREECOUNT = NA (no seedlings recorded)
OR_SEEDLING_post_fire_plots %>%
  filter(is.na(TREECOUNT)) %>%
  summarise(n_plot_ids = n_distinct(plot_id))
# 23 / 237 have no seedlings 

# number of seedlings per plot (aggregated across SUBP/microplot)
OR_SEEDLING_per_plot <- OR_SEEDLING_post_fire_plots %>%
  group_by(plot_id, INVYR, fire_year) %>%
  summarise(
    n_seedlings = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop"
  )

hist(OR_SEEDLING_per_plot$n_seedlings, breaks=30)
summary(OR_SEEDLING_per_plot$n_seedlings)

# number of species of seedlings per plot 
OR_SEEDLING_plot_summary <- OR_SEEDLING_post_fire_plots %>%
  group_by(plot_id, INVYR, fire_year) %>%
  summarise(
    n_seedling_species = n_distinct(SPCD),
    n_seedlings = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop"
  )
summary(OR_SEEDLING_plot_summary$n_seedling_species)

# number of individuals per species per plot
OR_SEEDLING_species_summary <- OR_SEEDLING_post_fire_plots %>%
  group_by(plot_id, INVYR, fire_year, SPCD) %>%
  summarise(n_seedlings = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop") %>%
  left_join(REF_SPECIES %>% dplyr::select(SPCD, GENUS, SPECIES),
    by = "SPCD") %>%
  mutate(species = paste(GENUS, SPECIES, sep = "_"))

summary(OR_SEEDLING_species_summary$n_seedlings)
```

## control plots

```{r}
# bring in plot_ids for burned plots (pre_and_post, post_only)
OR_prepost_burned_plot_id <- OR_prepost_fire_summary %>%
  filter(status %in% c("pre_and_post", "post_only")) 

# filter for plots that have DSTRBCD == 0 in all columns (DSTRBCD1-3) and all rows
OR_COND_control_summary <- OR_COND_IDs %>%
  filter(FORTYPCD %in% c(371, 201, 221, 222, 224, 225, 261, 922)) %>%
  group_by(plot_id) %>%
  summarise(
    all_zero = all(DSTRBCD1 == 0 & DSTRBCD2 == 0 & DSTRBCD3 == 0),
    .groups = "drop"
  ) %>%
  filter(all_zero) %>%
  select(plot_id)

OR_COND_treatment <- OR_COND_IDs %>%
  filter(
    plot_id %in% OR_prepost_burned_plot_id$plot_id |
      plot_id %in% OR_COND_control_summary$plot_id
  ) %>%
  mutate(
    condition = if_else(
      plot_id %in% OR_prepost_burned_plot_id$plot_id,
      "burned",
      "control"
    ),
    fire_code = case_when(
      condition == "control" ~ "0",
      DSTRBCD1 %in% c(30, 31, 32) ~ as.character(DSTRBCD1),
      DSTRBCD2 %in% c(30, 31, 32) ~ as.character(DSTRBCD2),
      DSTRBCD3 %in% c(30, 31, 32) ~ as.character(DSTRBCD3),
      TRUE ~ NA_character_
    )
  )

# Confirm unique plot counts
OR_COND_treatment %>%
  group_by(condition) %>%
  summarise(n_unique_plots = n_distinct(plot_id))

# burned               553
# control             4813

# Pull coordinates for these plots
OR_PLOT_coordinates <- OR_PLOT_IDs %>%
  semi_join(OR_COND_treatment, by = "plot_id") %>%
  distinct(plot_id, LAT, LON)


# lets pull climate for control and burned plots 

# convert lat/long points in a spatial object
OR_PLOT_coordinates_vect <- vect(
  OR_PLOT_coordinates, geom = c("LON", "LAT"), crs = "EPSG:4326")

# extract climate for all lat/long coordinates (all unique plot IDs)



OR_terra_climate_values <- terra::extract(
  terra_climate_stack,
  OR_PLOT_coordinates_vect
) %>%
  bind_cols(
    OR_PLOT_coordinates %>% select(plot_id, LAT, LON)
  ) %>%
  select(plot_id, LAT, LON, everything(), -ID)

OR_terra_climate_values <- OR_terra_climate_values %>%
  left_join(
    OR_COND_treatment %>%
      select(plot_id, condition, fire_code),
    by = "plot_id"
  )

points(
  OR_PLOT_coordinates_vect,
  col = ifelse(OR_terra_climate_values$condition == "burned", "red", "blue"),
  pch = 19
)

def_df <- as.data.frame(
  def_mean,
  xy = TRUE
)

ggplot(
  OR_terra_climate_values,
  aes(
    x = LON,
    y = LAT,
    color = def_mean
  )
) +
  geom_point(
    size = 3
  ) +
  scale_color_viridis_c(
    name = "Climatic water deficit\n(mm)"
  ) +
  coord_equal() +
  theme_classic()

# join climate back to main data frame with all plots 
#CA_PLOT_climate <- CA_PLOT_treatment %>%
  #left_join(
   # terra_climate_values,
   # by = "plot_id")

```

## climate and distance matrices
```{r}
# how far apart are two plots, considering 7 climate variables at once? 
# we can't calculate raw distances from these variables, it would be meaningless and skewed 

# standardizing climate variables using z-scores
# how many standard deviations above or below the average is this value?
# z = x - mean / sd

OR_climate_standardized <- OR_terra_climate_values %>%
  mutate(across(all_of(climate_variables), ~ as.numeric(scale(.)), .names = "z_{.col}"))

z_vars <- paste0("z_", climate_variables)
# now, each climate value is the number of standard deviations from that variable's own average

OR_burned_climate  <- OR_climate_standardized %>% filter(condition == "burned")
OR_control_climate <- OR_climate_standardized %>% filter(condition == "control")

# if there are any plots with NAs in lat/lon it will throw NAs for everything downstream
OR_control_climate_cleaned <- OR_control_climate %>% 
  filter(if_all(all_of(c(z_vars, "LAT", "LON")), ~ !is.na(.)))

# rdist() (fields package) computes distances between two separate matrices — every row in matrix A vs. every row in matrix B. we don't want burned-vs-burned or control-vs-control distances, we only want burned-vs-control distances 
OR_climate_distance_matrix <- rdist(
  as.matrix(OR_burned_climate[, z_vars]),
  as.matrix(OR_control_climate_cleaned[, z_vars])
)
# assign plot_ids to climate distance matrix
dimnames(OR_climate_distance_matrix) <- list(OR_burned_climate$plot_id, OR_control_climate_cleaned$plot_id)

# generate geographic distance matrix
# distm() (geosphere package) computes a distance matrix between two sets of geographic coordinates, using distHaversine formula, the great-circle distance that accounts for Earth's curvature
OR_geo_dist_mat <- distm(
  as.matrix(OR_burned_climate[, c("LON", "LAT")]),
  as.matrix(OR_control_climate_cleaned[, c("LON", "LAT")]),
  fun = distHaversine
) / 1000  # convert meters to km

dimnames(OR_geo_dist_mat) <- list(OR_burned_climate$plot_id, OR_control_climate_cleaned$plot_id)

# now that we have both climate and geographic distance matrices, we want to combine them into one score by normalizing distance matrix values from 0-1
norm01 <- function(m) (m - min(m)) / (max(m) - min(m)) # could also be done with scales::rescale() 

OR_clim_norm <- norm01(OR_climate_distance_matrix)
OR_geo_norm  <- norm01(OR_geo_dist_mat)

# weight toward climate similarity, over geographic proximity. 60% climate, 40% geographic
w_clim <- 0.6   

# combined distance across all  standardized climate variables using Euclidean distance
OR_combined_dist <- w_clim * norm01(OR_climate_distance_matrix) +
  (1 - w_clim) * norm01(OR_geo_dist_mat)

# pick the top 5 best matched control plots per single burned plot
n_match <- 5

OR_match_list <- lapply(seq_len(nrow(OR_combined_dist)), function(i) {
  ord <- order(OR_combined_dist[i, ])
  top <- ord[seq_len(n_match)]
  data.frame(
    burned_id = OR_burned_climate$plot_id[i],
    control_id = OR_control_climate$plot_id[top],
    combined_distance = OR_combined_dist[i, top],
    climate_distance = OR_climate_distance_matrix[i, top],
    geo_distance_km = OR_geo_dist_mat[i, top],
    rank = seq_len(n_match),
    row.names = NULL
  )
})

# 'matches' now has 5 rows per burned plot, ranked by closeness (in climate and geographic space)
OR_matches <- bind_rows(OR_match_list)
head(OR_matches, 10)

# lets make sure the same control plot isnt being applied to all burned plots
# how many times did this specific control plot show up as a burned plot match?
OR_matches %>% count(control_id, sort = TRUE) %>% head(10)
# 4 of the same control plots show up as a match for 10-13 burned plots
# most popular control plot was only used for about 3% of all burned plots


# filter control plot dataframe with the top 5 matched control plots per burned plot
OR_matched_controls <- OR_control_climate_cleaned %>% 
  filter(plot_id %in% OR_matches$control_id)

ggplot(OR_matches, aes(x = geo_distance_km, y = climate_distance)) +
  geom_point(alpha = 0.4) +
  labs(
    x = "Geographic distance (km)",
    y = "Climate distance (standardized)",
    title = "Climate distance vs. geographic distance for matched pairs"
  ) +
  theme_minimal()

```

## without replacement
```{r}

without_replacement <- function(dist_mat, clim_mat, geo_mat, n_match = 5) {
  
  # Build a long list of every possible (burned, control) pair with its distances
  triples <- expand.grid(
    burned_id  = rownames(dist_mat),
    control_id = colnames(dist_mat),
    stringsAsFactors = FALSE
  )
  triples$combined_distance <- as.vector(dist_mat)
  triples$climate_distance  <- as.vector(clim_mat)
  triples$geo_distance_km   <- as.vector(geo_mat)
  
  # Sort so the best (smallest distance) pairs come first
  triples <- triples[order(triples$combined_distance), ]
  
  needed   <- setNames(rep(n_match, nrow(dist_mat)), rownames(dist_mat))
  used_ctl <- setNames(rep(FALSE, ncol(dist_mat)), colnames(dist_mat))
  
  results <- vector("list", nrow(dist_mat) * n_match)
  counter <- 1
  
  for (k in seq_len(nrow(triples))) {
    b <- triples$burned_id[k]
    c <- triples$control_id[k]
    
    if (needed[b] > 0 && !used_ctl[c]) {
      results[[counter]] <- data.frame(
        burned_id         = b,
        control_id        = c,
        combined_distance = triples$combined_distance[k],
        climate_distance  = triples$climate_distance[k],
        geo_distance_km   = triples$geo_distance_km[k],
        rank              = n_match - needed[b] + 1,
        row.names         = NULL
      )
      counter <- counter + 1
      needed[b]   <- needed[b] - 1
      used_ctl[c] <- TRUE
    }
    
    if (all(needed == 0)) break
  }
  
  bind_rows(results)
}

OR_matches_no_replacement <- without_replacement(
  dist_mat = OR_combined_dist,
  clim_mat = OR_climate_distance_matrix,
  geo_mat  = OR_geo_dist_mat,
  n_match  = 3
)

OR_matches_no_replacement %>% count(burned_id) %>% filter(n < 5)
```


## species overlap
```{r}
# now that we have our control and burned plot pairs, lets look into species overlap between each pair

OR_burned_and_control_plot_id <- unique(c(OR_matches_no_replacement$burned_id, OR_matches_no_replacement$control_id))

#2212 plots total
# burned
# control

# lets filter CA_TREE and CA_SEEDLING for our burned and control plots

OR_TREE_IDs_filtered <- OR_TREE_IDs %>%
  filter(plot_id %in% OR_burned_and_control_plot_id)


# add species names to species codes using REF_SPECIES

OR_TREE_species <- OR_TREE_IDs_filtered %>%
  left_join(
    REF_SPECIES %>%
      mutate(species = paste(GENUS, SPECIES, sep = "_")) %>%
      select(SPCD, species),
    by = "SPCD"
  ) %>%
  left_join(
    OR_COND_treatment %>% distinct(plot_id, condition),
    by = "plot_id"
  ) %>%
  left_join(
    OR_prepost_fire_summary %>% distinct(plot_id, fire_year),
    by = "plot_id"
  )

OR_TREE_species_filtered <- OR_TREE_species %>%
  filter(
    condition == "burned" | condition == "control"
  )

#Compute total abundance per species — count all tree rows per species, across every plot.
#Compute frequency (commonness) per species — count the number of distinct plots each species appears in, divide by total number of plots.

# for all plots
OR_species_summary_combined <- OR_TREE_species_filtered %>%
  group_by(species) %>%
  summarise(
    total_abundance = n(),
    n_plots = n_distinct(plot_id),
    .groups = "drop"
  ) %>%
  mutate(
    frequency = n_plots / n_distinct(OR_TREE_species_filtered$plot_id)
  ) %>%
  arrange(desc(total_abundance))

# by condition
OR_species_summary_by_condition <- OR_TREE_species_filtered %>%
  group_by(condition, species) %>%
  summarise(
    total_abundance = n(),
    n_plots = n_distinct(plot_id),
    .groups = "drop"
  ) %>%
  group_by(condition) %>%
  mutate(
    frequency = n_plots / n_distinct(OR_TREE_species_filtered$plot_id[OR_TREE_species_filtered$condition == condition])
  ) %>%
  ungroup() %>%
  arrange(condition, desc(total_abundance))


# Rank species on each metric separately, then combine the two ranks into one overall score (e.g., average rank, so a species that's #3 in abundance and #10 in frequency gets averaged into a combined rank).

#Sort by the combined rank and take the top 75.

OR_top_75_species <- OR_species_summary_combined %>%
  mutate(
    abundance_rank = rank(-total_abundance),
    frequency_rank = rank(-frequency),
    combined_rank = (abundance_rank + frequency_rank) / 2
  ) %>%
  arrange(combined_rank) %>%
  slice_head(n = 75)
# 54 species total

#total number of species across both control and burned plots
OR_TREE_species_filtered %>%
  group_by(condition) %>%
  summarise(n_species = n_distinct(species), .groups = "drop")

OR_all_species <- OR_TREE_species_filtered %>%
  distinct(species) %>%
  pull(species)

length(OR_all_species)
# 54 total species

# to calculate dominance (the number of indiviudals of that species in that plot) and frequency (commonness)  (the number of plots the species occurs in / total number of plots)

OR_tree_count <- OR_TREE_species_filtered %>%
  group_by(plot_id) %>%
  summarise(total_trees_in_plot = n(), .groups = "drop")

OR_species_dominance <- OR_TREE_species_filtered %>%
  group_by(plot_id, species) %>%
  summarise(species_count = n(), .groups = "drop") %>%
  left_join(OR_tree_count, by = "plot_id") %>%
  mutate(relative_abundance = species_count / total_trees_in_plot)

OR_species_summary <- OR_species_dominance %>%
  group_by(species) %>%
  summarise(
    n_plots_present  = n_distinct(plot_id),
    avg_dominance    = mean(relative_abundance),
    median_dominance = median(relative_abundance),
    min_dominance    = min(relative_abundance),
    max_dominance    = max(relative_abundance),
    .groups = "drop"
  ) %>%
  mutate(frequency = n_plots_present / n_distinct(OR_TREE_species_filtered$plot_id)) %>%
  arrange(desc(frequency))

# lets check whether dominant and common species differ by condition (burned vs control)

OR_species_dominance_condition <- OR_species_dominance %>%
  left_join(
    OR_TREE_species_filtered %>% distinct(plot_id, condition),
    by = "plot_id"
  )

# now summarize by condition + species instead of species alone
OR_species_summary_by_condition <- OR_species_dominance_condition %>%
  group_by(condition, species) %>%
  summarise(
    n_plots_present  = n_distinct(plot_id),
    avg_dominance    = mean(relative_abundance),
    median_dominance = median(relative_abundance),
    min_dominance    = min(relative_abundance),
    max_dominance    = max(relative_abundance),
    .groups = "drop"
  ) %>%
  group_by(condition) %>%
  mutate(
    frequency = n_plots_present / n_distinct(OR_TREE_species_filtered$plot_id[OR_TREE_species_filtered$condition == condition])
  ) %>%
  ungroup() %>%
  arrange(condition, desc(frequency))

OR_species_condition_comparison <- OR_species_summary_by_condition %>%
  select(condition, species, frequency, median_dominance) %>%
  pivot_wider(
    names_from = condition,
    values_from = c(frequency, median_dominance)
  ) %>%
  mutate(
    frequency_diff = frequency_burned - frequency_control,
    dominance_diff = median_dominance_burned - median_dominance_control
  ) %>%
  arrange(desc(abs(dominance_diff)))

library(ggrepel)

OR_species_condition_comparison %>%
  filter(!is.na(dominance_diff), !is.na(frequency_diff)) %>%
  ggplot(aes(x = frequency_diff, y = dominance_diff, label = species)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray70") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray70") +
  geom_point(alpha = 0.6, size = 2) +
  geom_text_repel(size = 3, max.overlaps = 15) +
  labs(
    x = "Frequency difference (burned − control)",
    y = "Dominance difference (burned − control)",
    title = "Burned - control species dominance (median)"
  ) +
  theme_minimal()

# is the number of species recruiting in control plots greater than burned (OR)

OR_SEEDLING_condition_filtered <- OR_SEEDLING_IDs %>%
  filter(plot_id %in% OR_burned_and_control_plot_id) %>%
  left_join(
    OR_COND_treatment %>% distinct(plot_id, condition),
    by = "plot_id"
  )

OR_SEEDLING_species_summary_condition <- OR_SEEDLING_condition_filtered %>%
  group_by(plot_id, condition, SPCD) %>%
  summarise(n_seedlings = sum(TREECOUNT, na.rm = TRUE), .groups = "drop") %>%
  left_join(REF_SPECIES %>% dplyr::select(SPCD, GENUS, SPECIES), by = "SPCD") %>%
  mutate(species = paste(GENUS, SPECIES, sep = "_"))

OR_SEEDLING_plot_summary_condition <- OR_SEEDLING_species_summary_condition %>%
  group_by(plot_id, condition) %>%
  summarise(
    n_seedling_species = n_distinct(SPCD),
    n_seedlings = sum(n_seedlings, na.rm = TRUE),
    .groups = "drop"
  )

OR_SEEDLING_plot_summary_condition %>%
  ggplot(aes(x = n_seedling_species, fill = condition)) +
  geom_histogram(binwidth = 1, color = "white", position = "identity", alpha = 0.6) +
  facet_wrap(~condition, ncol = 1) +
  theme_bw() +
  labs(
    x = "Number of seedling species per plot",
    y = "Number of plots",
    title = "Seedling species richness: burned vs. control (OR)"
  )

OR_SEEDLING_plot_summary_condition %>%
  ggplot(aes(x = condition, y = n_seedling_species, fill = condition)) +
  geom_boxplot(alpha = 0.6) +
  theme_bw() +
  labs(
    x = NULL,
    y = "Number of seedling species per plot",
    title = "Seedling species richness: burned vs. control (OR)"
  )

OR_seedling_richness_summary_by_condition <- OR_SEEDLING_plot_summary_condition %>%
  group_by(condition) %>%
  summarise(
    mean_richness   = mean(n_seedling_species),
    median_richness = median(n_seedling_species),
    n_plots         = n(),
    .groups = "drop"
  )

OR_seedling_richness_summary_by_condition

OR_SEEDLING_species_summary_condition %>%
  group_by(species, condition, plot_id) %>%
  summarise(n_seedlings = sum(n_seedlings), .groups = "drop") %>%
  group_by(species, condition) %>%
  summarise(mean_per_plot = mean(n_seedlings), .groups = "drop") %>%
  group_by(condition) %>%
  slice_max(mean_per_plot, n = 20) %>%
  ungroup() %>%
  ggplot(aes(x = reorder_within(species, mean_per_plot, condition), y = mean_per_plot, fill = condition)) +
  geom_col() +
  coord_flip() +
  scale_x_reordered() +
  facet_wrap(~condition, scales = "free_y") +
  labs(
    x = "species",
    y = "average # seedlings per plot",
    title = "Average seedling abundance per species per plot: burned vs. control (OR)"
  ) +
  theme_bw()

OR_SEEDLING_species_summary_condition %>%
  group_by(plot_id, condition, species) %>%
  summarise(n_seedlings = sum(n_seedlings), .groups = "drop") %>%
  mutate(n_seedlings_capped = pmin(n_seedlings, 50)) %>%
  ggplot(aes(x = species, y = n_seedlings_capped, fill = condition)) +
  geom_boxplot(alpha = 0.6) +
  coord_flip() +
  facet_wrap(~condition, scales = "free_y") +
  labs(
    x = "species",
    y = "seedlings per plot (capped at 50)",
    title = "Seedling abundance per species per plot: burned vs. control (OR)"
  ) +
  theme_bw()

summary(OR_SEEDLING_plot_summary_condition$n_seedling_species[OR_SEEDLING_plot_summary_condition$condition == "burned"])
summary(OR_SEEDLING_plot_summary_condition$n_seedling_species[OR_SEEDLING_plot_summary_condition$condition == "control"])

```

```{r}

# species found in both states
species_shared_CA_OR <- intersect(CA_all_species, OR_all_species)
length(species_shared_CA_OR)

# species found only in CA
species_only_CA <- setdiff(CA_all_species, OR_all_species)
length(species_only_CA)

# species found only in OR
species_only_OR <- setdiff(OR_all_species, CA_all_species)
length(species_only_OR)

# sanity check: total unique species across both states combined
length(union(CA_all_species, OR_all_species))
```

## whole plot network
```{r}

# filter CA_COND_IDs for forest types only, no disturbance filtering
OR_COND_forest_type <- OR_COND_IDs %>%
  filter(FORTYPCD %in% c(201, 202, 221, 222, 224, 225, 226, 241, 261, 262, 263, 264, 265, 267, 268, 270, 271, 281, 301, 304, 305, 321, 341, 342, 361, 365, 371, 722, 911, 912, 922, 923, 933, 941, 961))
         
(371, 201, 221, 222, 224, 225, 261, 922))

# get the unique plot_ids that qualify
OR_forest_type_plot_id <- OR_COND_forest_type %>%
  distinct(plot_id) %>%
  pull(plot_id)

length(OR_forest_type_plot_id)

# filter CA_PLOT_IDs for those plot_ids
OR_PLOT_forest_type <- OR_PLOT_IDs %>%
  filter(plot_id %in% OR_forest_type_plot_id)

# filter CA_TREE_IDs for those plot_ids
OR_TREE_forest_type <- OR_TREE_IDs %>%
  filter(plot_id %in% OR_forest_type_plot_id)

nrow(OR_PLOT_forest_type)
n_distinct(OR_TREE_forest_type$plot_id)

OR_TREE_forest_type_species <- OR_TREE_forest_type %>%
  left_join(
    REF_SPECIES %>%
      mutate(species = paste(GENUS, SPECIES, sep = "_")) %>%
      select(SPCD, species),
    by = "SPCD"
  )
sum(is.na(OR_TREE_forest_type_species$species))
n_distinct(OR_TREE_forest_type_species$species)

# how different are the species from this workflow and the previous control plot filtering by climate?

# species from your burned/control workflow (already built)
OR_all_species   # from earlier: distinct species across CA_TREE_species_filtered

# species from your new forest-type-only workflow
OR_forest_type_all_species <- OR_TREE_forest_type_species %>%
  distinct(species) %>%
  pull(species)

length(OR_forest_type_all_species)

# species found in both workflows
intersect(OR_all_species, OR_forest_type_all_species)

# species found only in the burned/control workflow
setdiff(OR_all_species, OR_forest_type_all_species)

# species found only in the forest-type-only workflow
setdiff(OR_forest_type_all_species, OR_all_species)


# make sure the species coming out of SEEDLING match TREE
# filter CA_SEEDLING_IDs for those plot_ids
OR_SEEDLING_forest_type <- OR_SEEDLING_IDs %>%
  filter(plot_id %in% OR_forest_type_plot_id)

OR_SEEDLING_forest_type_species <- OR_SEEDLING_forest_type %>%
  left_join(
    REF_SPECIES %>%
      mutate(species = paste(GENUS, SPECIES, sep = "_")) %>%
      select(SPCD, species),
    by = "SPCD"
  )

# species list from seedling data (forest-type-only workflow)
OR_forest_type_all_seedling_species <- OR_SEEDLING_forest_type_species %>%
  distinct(species) %>%
  pull(species)

length(OR_forest_type_all_species)
length(OR_forest_type_all_seedling_species)

# species found in both tree and seedling data
intersect(OR_forest_type_all_species, OR_forest_type_all_seedling_species)

# species found only in tree data (mature trees, no seedlings recorded)
setdiff(OR_forest_type_all_species, OR_forest_type_all_seedling_species)

# species found only in seedling data (recruiting, but no mature trees recorded)
setdiff(OR_forest_type_all_seedling_species, OR_forest_type_all_species)

# create a shared species list
OR_forest_type_all_species
OR_forest_type_all_seedling_species

OR_species <- data.frame(
  species = unique(c(
    OR_forest_type_all_species,
    OR_forest_type_all_seedling_species
  ))
)

# how much overlap is there between CA and OR all species for our forest types
setdiff(OR_species, CA_species) # 18 species in OR only 
setdiff(CA_species, OR_species) # 22 species in CA only 

# Number of shared species
length(intersect(CA_species$species, OR_species$species))

# Species only in Oregon
length(setdiff(OR_species$species, CA_species$species))   # 18

# Species only in California
length(setdiff(CA_species$species, OR_species$species))   # 22

# Total unique species across both states
length(union(CA_species$species, OR_species$species)) 
# 95 unique species across both states

# Or equivalently
length(unique(c(CA_species$species, OR_species$species)))

```


#WA
### fire only plots
```{r}
#create a unique PLOT ID in COND and in PLOT tables 
WA_COND_IDs <- WA_COND %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )
WA_PLOT_IDs <- WA_PLOT %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

# collapse information from multiple CONDIDs associated with each plot to avoid many-to-many mapping
# and then filter by fire and forest type  

#plot_id → multiple CONDIDs → each CONDID has up to 3 disturbance codes and years
# one row for a specific plot_id, CONID, and disturbance event (combining DSTRBCD and DSTRBYR)
WA_COND_long <- WA_COND_IDs %>%
  dplyr::select(plot_id, CONDID, FORTYPCD,
    DSTRBCD1, DSTRBCD2, DSTRBCD3,
    DSTRBYR1, DSTRBYR2, DSTRBYR3) %>%
  pivot_longer(
    cols = starts_with("DSTR"),
    names_to = c(".value", "event"),
    names_pattern = "(DSTRBCD|DSTRBYR)([1-3])"
  ) %>%
  filter(DSTRBCD %in% c(30, 31, 32)) %>%
  filter(FORTYPCD %in% c(371, 201, 221, 222, 224, 225, 261, 922)) %>%
  mutate(
    fire_year_severity = paste(DSTRBYR, DSTRBCD, sep = "_")
  )

# summarize fire history across rows for unique plot_id (one row per plot_id)
WA_fire_summary <- WA_COND_long %>%
  group_by(plot_id) %>%
  summarise(
    n_fire_events = n_distinct(DSTRBYR), # number of distinct fire years per plot
    
    # this just collapses all codes into a list but not necessarily relative to year
    fire_types = paste(sort(unique(DSTRBCD)), collapse = ", "), 
    fire_years = paste(sort(unique(DSTRBYR)), collapse = ", "),
    
    fire_year_severity =
    paste(unique(fire_year_severity), collapse = ", "),
    
    multiple_fires = n_distinct(DSTRBYR) > 1, # did fire occur more than once?
    multiple_severity = n_distinct(DSTRBCD) > 1, # are there multiple fire types?
    first_fire_year = min(DSTRBYR, na.rm = TRUE),
    last_fire_year = max(DSTRBYR, na.rm = TRUE),
    .groups = "drop"
  )

# how many plots burned multiple times?
WA_fire_summary %>% summarise(n_plots_multiple_fires = sum(multiple_fires))
# 4 plots

# how many plots that burned multipe times have different severities?
WA_fire_summary %>% filter(multiple_fires) %>% summarise(n_plots_multiple_fire_and_severity = sum(multiple_severity))
# 0 plots

# how many plots that burned once have multiple severities within a plot?
WA_fire_summary %>%
  filter(!multiple_fires) %>%
  summarise(
    n_plots_single_fire_multiple_severity =
      sum(multiple_severity))
# 5 plots

WA_burned_PLOT_IDs <- WA_fire_summary %>% 
  distinct(plot_id)

# merge COND and PLOT by this unique ID 
WA_PLOT_by_COND <- WA_PLOT_IDs %>%
  semi_join(WA_burned_PLOT_IDs, by = "plot_id") %>%  #keep only plot_ids with at least one fire 
  left_join(WA_fire_summary, by = "plot_id")  #add fire summary to each plot, should be one-to-many mapping

#for each unique plot, count how many unique inventory years (INVYR) it has
WA_PLOT_by_COND %>%
  group_by(plot_id) %>%
  summarise(
    n_revisits = n_distinct(INVYR),
    .groups = "drop"
  ) %>%
  count(n_revisits)

# n=2 (241)


# how many plots have pre and post fire surveys
WA_fire_events <- WA_COND_long %>%
  dplyr::select(plot_id, DSTRBYR) %>%
  distinct() %>%
  rename(fire_year = DSTRBYR)

# does each unique plot_id have a pre and post fire survey around any/all fire years?
WA_prepost_fire_summary <- WA_PLOT_by_COND %>%
  distinct(plot_id, INVYR) %>%
  inner_join(
    WA_fire_events %>% distinct(plot_id, fire_year),
    by = "plot_id"
  ) %>%
  mutate(
    relationship = case_when(
      INVYR < fire_year ~ "pre",
      INVYR > fire_year ~ "post",
      TRUE ~ "same_year"
    )
  ) %>%
  group_by(plot_id, fire_year) %>%
  summarise(
    has_pre  = any(relationship == "pre"),
    has_post = any(relationship == "post"),
    n_pre_surveys  = sum(relationship == "pre"),
    n_post_surveys = sum(relationship == "post"),
    .groups = "drop"
  ) %>%
  mutate(
      status = case_when(
      has_pre & has_post ~ "pre_and_post",
      has_pre ~ "pre_only",
      has_post ~ "post_only",
      TRUE ~ "no_clear_separation"
    )
  )

WA_prepost_fire_summary %>%
  count(status)

# post only = 79
# pre only = 11
#pre and post = 155


# how many plots are there in CA
WA_PLOT %>%
  distinct(LAT, LON) %>%
  nrow()
# 11778

length(unique(WA_COND_IDs$plot_id)) # 11988
length(unique(WA_PLOT_IDs$plot_id)) # 11988


```
## tree
```{r}
#Tree number. A number used to uniquely identify a tree on a subplot. Tree numbers can be used to track trees when PLOT.DESIGNCD or DESIGNCD is the same between inventories

# before looking into tree level data, just confirm that the design codes are the same across plots in our CA_PLOT_by_COND table that has our burned plots
n_distinct(WA_PLOT_by_COND$DESIGNCD) # two design codes across the filtered plots
unique(WA_PLOT_by_COND$DESIGNCD) # 501, 502

# do we have surviving trees in post-fire surveys?
# create unique plot ids for CA_TREE
WA_TREE_IDs <- WA_TREE %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

#filter CA_TREE to only keep the unique plot ids from CA_PLOT_by_COND (our fire only plots in our focal forest types)
WA_TREE_filtered <- WA_TREE_IDs %>%
  filter(plot_id %in% WA_PLOT_by_COND$plot_id)

# to locate a specific tree ID, use the following variables 
# STATECD, INVYR, UNITCD, COUNTYCD, PLOT, SUBP, TREE
# TREE is unique to SUBP not plot_id
# we can look for unique combinations of SUBP and TREE within a plot_id (STATECD, UNITCD, COUNTYCD, PLOT)

# is TREE our unique identifier of TREE ?
WA_TREE_filtered %>%
  group_by(TREE) %>%
  summarise(
    n_plots = n_distinct(plot_id),
    .groups = "drop"
  ) %>%
  filter(n_plots > 1) %>%
  summarise(
    n_trees_multiple_plots = n()
  )
# 295 trees are found in multiple plots 

WA_TREE_filtered %>%
  summarise(
    n_unique_trees = n_distinct(plot_id, SUBP, TREE)
  )
# number of unique trees 7215

# Tree number notes:
# TREE uniquely identifies a tree within a subplot (SUBP),
# so full tree identity is: plot_id + SUBP + TREE

# unique tree identity is plot_id + SUBP + TREE
# make a unique tree identifier too? or use plot_id + SUBP + TREE everytime
#tree_id = paste(plot_id, SUBP, TREE, sep = "_") 

# lets look at trees in the plots with both pre and post fire surveys
# pull the plot_ids for status = pre_and_post from CA_prepost_fire_summary and the fire year the post-fire plot is associated with

WA_TREE_prepost_fire <- WA_TREE_filtered %>%
  semi_join(
    WA_prepost_fire_summary %>% filter(status == "pre_and_post"),
    by = "plot_id"
  ) %>%
  left_join(
    WA_prepost_fire_summary %>%
      filter(status == "pre_and_post") %>%
      dplyr::select(plot_id, fire_year, status),
    by = "plot_id"
  )

WA_prepost_fire_summary %>%
  filter(status == "pre_and_post") %>%
  summarise(n_plots = n_distinct(plot_id)) #155
length(unique(WA_TREE_prepost_fire$plot_id)) #155


# now we want to look at just post-fire plots (where INVYR > fire_year)
WA_TREE_post_fire_plots <- WA_TREE_prepost_fire %>%
  filter(INVYR > fire_year)

# number of trees per plot (post-fire plots)
WA_TREE_per_plot_post <- WA_TREE_post_fire_plots %>%
  filter(STATUSCD == 1) %>%
  mutate(which = "post") %>%
  group_by(plot_id, INVYR, fire_year, status, which) %>%
  summarise(
    n_live_tree = n_distinct(SUBP, TREE),
    .groups = "drop"
  )
summary(WA_TREE_per_plot_post$n_live_tree)

# number of species with STATUSCD = 1 in post-fire plots
WA_TREE_post_fire_plot_summary <- WA_TREE_post_fire_plots %>%
  filter(STATUSCD == 1) %>%
  group_by(plot_id, INVYR, fire_year) %>%
  summarise(
    n_live_species = n_distinct(SPCD),
    n_live_trees = n_distinct(SUBP, TREE),
    .groups = "drop"
  )
summary(WA_TREE_post_fire_plot_summary$n_live_species)

# number of individuals (live) per species per plot
WA_TREE_post_fire_species_summary <- WA_TREE_post_fire_plots %>%
  filter(STATUSCD == 1) %>%
  group_by(plot_id, INVYR, fire_year, SPCD) %>%
  summarise(
    n_live_trees = n_distinct(SUBP, TREE),
    .groups = "drop"
  ) %>%
  left_join(
    REF_SPECIES %>%
      dplyr::select(SPCD, GENUS, SPECIES),
    by = "SPCD"
  ) %>%
  mutate(species = paste(GENUS, SPECIES, sep = "_"))
summary(WA_TREE_post_fire_species_summary$n_live_trees)

## tree transitions and mortality rates

# closest PRE-fire observation for each TREE
WA_TREE_pre_fire_trees <- WA_TREE_prepost_fire %>%
  filter(INVYR < fire_year) %>%
  group_by(plot_id, SUBP, TREE, fire_year) %>%
  arrange(desc(INVYR), .by_group = TRUE) %>%
  slice(1) %>%
  ungroup() %>%
  rename(
    pre_INVYR = INVYR,
    pre_STATUSCD = STATUSCD
  )

# closest POST-fire observation for each TREE
WA_TREE_post_fire_trees <- WA_TREE_prepost_fire %>%
  filter(INVYR > fire_year) %>%
  group_by(plot_id, SUBP, TREE, fire_year) %>%
  arrange(INVYR, .by_group = TRUE) %>%
  slice(1) %>%
  ungroup() %>%
  rename(
    post_INVYR = INVYR,
    post_STATUSCD = STATUSCD
  )

# join pre and post observations
WA_TREE_transitions <- WA_TREE_pre_fire_trees %>%
  inner_join(
    WA_TREE_post_fire_trees,
    by = c("plot_id", "SUBP", "TREE", "fire_year"),
    suffix = c("_pre", "_post")
  )

# mortality indicators
WA_TREE_transitions <- WA_TREE_transitions %>%
  mutate(
    alive_pre = pre_STATUSCD == 1,
    dead_post = pre_STATUSCD == 1 & post_STATUSCD == 2
  )

# plot-level mortality
WA_TREE_mortality <- WA_TREE_transitions %>%
  group_by(plot_id, fire_year) %>%
  summarise(
    n_alive_pre = sum(alive_pre, na.rm = TRUE),
    n_dead_post = sum(dead_post, na.rm = TRUE),
    mortality_rate = n_dead_post / n_alive_pre,
    .groups = "drop"
  )

summary(WA_TREE_mortality$mortality_rate)
hist(WA_TREE_mortality$mortality_rate)

# check for any new trees in post fire surveys

# trees appearing post-fire but absent pre-fire
WA_TREE_new_trees <- WA_TREE_post_fire_trees %>%
  anti_join(
    WA_TREE_transitions %>%
      distinct(plot_id, SUBP, fire_year, TREE),
    by = c("plot_id", "SUBP", "fire_year", "TREE")
  )

WA_TREE_new_trees %>%
  summarise(
    total_new_recruits = n()
  )
# 600 new post-fire recruits in WA_TREE

WA_TREE_new_trees %>%
  summarise(
    total_new_recruits = n_distinct(plot_id, SUBP, TREE, fire_year)
  )

# recruits per plot
WA_TREE_new_recruits_per_plot <- WA_TREE_new_trees %>%
  group_by(plot_id, fire_year, post_INVYR) %>%
  summarise(
    n_new_recruits = n_distinct(SUBP, TREE),
    .groups = "drop"
  )

summary(WA_TREE_new_recruits_per_plot$n_new_recruits)

# proportion of new recruits per plot
WA_TREE_recruitment <- WA_TREE_new_recruits_per_plot %>%
  left_join(
    WA_TREE_per_plot_post %>%
      dplyr::select(
        plot_id,
        fire_year,
        post_INVYR = INVYR,
        n_live_tree
      ),
    by = c("plot_id", "fire_year", "post_INVYR")
  ) %>%
  mutate(
    recruit_prop = n_new_recruits / n_live_tree
  )

WA_TREE_recruitment %>%
  filter(recruit_prop == 1)

# 1 plot has only new trees

summary(WA_TREE_recruitment$recruit_prop)
```

## seedling
```{r}
# is there recruitment in post-fire surveys?

# seedling counts are under TREECOUNT in CA_SEEDLING
# TREECOUNT is for  (plot × subplot/microplot × species × inventory)
# in the CA_SEEDLING table, SUBP = microplot since seedlings are inventoried at the microplot within a subplot

# create unique plot ids for CA_SEEDLING
WA_SEEDLING_IDs <- WA_SEEDLING %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

#filter CA_SEEDLING to only keep the unique plot ids from CA_PLOT_by_COND
WA_SEEDLING_filtered <- WA_SEEDLING_IDs %>%
  filter(plot_id %in% WA_PLOT_by_COND$plot_id)

# lets look at trees in the plots with both pre and post fire 
# pull the plot_ids for status = pre_and_post from CA_prepost_fire_summary

WA_SEEDLING_prepost_fire <- WA_SEEDLING_filtered %>%
  semi_join(WA_prepost_fire_summary %>% filter(status == "pre_and_post"),
    by = "plot_id") %>%
  left_join(WA_prepost_fire_summary %>%
      filter(status == "pre_and_post") %>%
      dplyr::select(plot_id, fire_year, status),
    by = "plot_id")

# filter to post-fire plots only
WA_SEEDLING_post_fire_plots <- WA_SEEDLING_prepost_fire %>%
  filter(INVYR > fire_year)


length(unique(WA_SEEDLING_post_fire_plots$plot_id)) # 68
length(unique(WA_TREE_post_fire_plots$plot_id)) # 155
# 68 unique plots (why is this different from the 155 for the TREE analysis ??)

## To qualify for counting, conifer seedlings must be at least 6 inches tall and hardwood seedlings must be at least 12 inches tall (which is maybe why that column is blank on the table).

# how many post fire plots have TREECOUNT = NA (no seedlings recorded)
WA_SEEDLING_post_fire_plots %>%
  filter(is.na(TREECOUNT)) %>%
  summarise(n_plot_ids = n_distinct(plot_id))
# 0 / 203 have no seedlings 

# number of seedlings per plot (aggregated across SUBP/microplot)
WA_SEEDLING_per_plot <- WA_SEEDLING_post_fire_plots %>%
  group_by(plot_id, INVYR, fire_year) %>%
  summarise(
    n_seedlings = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop"
  )

hist(WA_SEEDLING_per_plot$n_seedlings, breaks=30)
summary(WA_SEEDLING_per_plot$n_seedlings)

# number of seedlings per species per plot (aggregated across SUBP/microplot)
WA_SEEDLING_species_summary <- WA_SEEDLING_post_fire_plots %>%
  group_by(plot_id, INVYR, fire_year, SPCD) %>%
  summarise(n_seedlings = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop") %>%
  left_join(REF_SPECIES %>% dplyr::select(SPCD, GENUS, SPECIES),
    by = "SPCD") %>%
  mutate(species = paste(GENUS, SPECIES, sep = "_"))

summary(WA_SEEDLING_species_summary$n_seedlings)


WA_SEEDLING_species_summary %>%
  group_by(plot_id) %>%
  summarise(n_species = n_distinct(SPCD)) %>%
  ggplot(aes(x = n_species)) +
  geom_histogram(binwidth = 1, fill = "darkgreen", color = "white") +
  theme_bw() +
  labs(
    x = "Number of seedling species per plot",
    y = "Number of plots"
  )

# number of seedling species per plot 
WA_SEEDLING_plot_summary <- WA_SEEDLING_post_fire_plots %>%
  group_by(plot_id, INVYR, fire_year) %>%
  summarise(
    n_seedling_species = n_distinct(SPCD),
    n_seedlings = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop"
  )

```

## expanding to include post-only
```{r}

WA_TREE_IDs <- WA_TREE %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

#filter CA_TREE to only keep the unique plot ids from CA_PLOT_by_COND (our fire only plots in our focal forest types)
WA_TREE_filtered <- WA_TREE_IDs %>%
  filter(plot_id %in% WA_PLOT_by_COND$plot_id)

WA_TREE_prepost_fire <- WA_TREE_filtered %>%
  semi_join(
    WA_prepost_fire_summary %>%
      filter(status %in% c("pre_and_post", "post_only")),
    by = "plot_id"
  ) %>%
  left_join(
    WA_prepost_fire_summary %>%
      filter(status %in% c("pre_and_post", "post_only")) %>%
      dplyr::select(plot_id, fire_year, status),
    by = "plot_id"
  )


# now we want to look at just post-fire plots (where INVYR > fire_year)
WA_TREE_post_fire_plots <- WA_TREE_prepost_fire %>%
  filter(INVYR > fire_year)

length(unique(WA_TREE_post_fire_plots$plot_id)) # 230 post fire plots

# number of living trees per plot (post-fire plots)
WA_TREE_per_plot_post <- WA_TREE_post_fire_plots %>%
  filter(STATUSCD == 1) %>%
  mutate(which = "post") %>%
  group_by(plot_id, INVYR, fire_year, status, which) %>%
  summarise(
    n_live_tree = n_distinct(SUBP, TREE),
    .groups = "drop"
  )
summary(WA_TREE_per_plot_post$n_live_tree)


# number of species with STATUSCD = 1 in post-fire plots
WA_TREE_post_fire_plot_summary <- WA_TREE_post_fire_plots %>%
  filter(STATUSCD == 1) %>%
  group_by(plot_id, INVYR, fire_year) %>%
  summarise(
    n_live_species = n_distinct(SPCD),
    n_live_trees = n_distinct(SUBP, TREE),
    .groups = "drop"
  )
summary(WA_TREE_post_fire_plot_summary$n_live_species)

# number of individuals (live) per species per plot
WA_TREE_post_fire_species_summary <- WA_TREE_post_fire_plots %>%
  filter(STATUSCD == 1) %>%
  group_by(plot_id, INVYR, fire_year, SPCD) %>%
  summarise(
    n_live_trees = n_distinct(SUBP, TREE),
    .groups = "drop"
  ) %>%
  left_join(
    REF_SPECIES %>%
      dplyr::select(SPCD, GENUS, SPECIES),
    by = "SPCD"
  ) %>%
  mutate(species = paste(GENUS, SPECIES, sep = "_"))

summary(WA_TREE_post_fire_species_summary$n_live_trees)

# is there recruitment in post-fire surveys? (pre_and_post and post_only)

WA_SEEDLING_prepost_fire <- WA_SEEDLING_filtered %>%
  semi_join(WA_prepost_fire_summary %>% filter(status %in% c("pre_and_post", "post_only")),
    by = "plot_id") %>%
  left_join(WA_prepost_fire_summary %>%
      filter(status %in% c("pre_and_post", "post_only")) %>%
      dplyr::select(plot_id, fire_year, status),
    by = "plot_id")

# filter to post-fire plots only
WA_SEEDLING_post_fire_plots <- WA_SEEDLING_prepost_fire %>%
  filter(INVYR > fire_year)

length(unique(WA_SEEDLING_post_fire_plots$plot_id)) # 126
# 126 unique plots instead of the 68 in the pre_and_post analysis above

# how many post fire plots have TREECOUNT = NA (no seedlings recorded)
WA_SEEDLING_post_fire_plots %>%
  filter(is.na(TREECOUNT)) %>%
  summarise(n_plot_ids = n_distinct(plot_id))
# 4 / 126 have no seedlings 

# number of seedlings per plot (aggregated across SUBP/microplot)
WA_SEEDLING_per_plot <- WA_SEEDLING_post_fire_plots %>%
  group_by(plot_id, INVYR, fire_year) %>%
  summarise(
    n_seedlings = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop"
  )

hist(WA_SEEDLING_per_plot$n_seedlings, breaks=30)
summary(WA_SEEDLING_per_plot$n_seedlings)

# number of species of seedlings per plot 
WA_SEEDLING_plot_summary <- WA_SEEDLING_post_fire_plots %>%
  group_by(plot_id, INVYR, fire_year) %>%
  summarise(
    n_seedling_species = n_distinct(SPCD),
    n_seedlings = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop"
  )

# number of individuals per species per plot
WA_SEEDLING_species_summary <- WA_SEEDLING_post_fire_plots %>%
  group_by(plot_id, INVYR, fire_year, SPCD) %>%
  summarise(n_seedlings = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop") %>%
  left_join(REF_SPECIES %>% dplyr::select(SPCD, GENUS, SPECIES),
    by = "SPCD") %>%
  mutate(species = paste(GENUS, SPECIES, sep = "_"))
```

## control plots
```{r}
# bring in plot_ids for burned plots (pre_and_post, post_only)
WA_prepost_burned_plot_id <- WA_prepost_fire_summary %>%
  filter(status %in% c("pre_and_post", "post_only")) 

# filter for plots that have DSTRBCD == 0 in all columns (DSTRBCD1-3) and all rows
WA_COND_control_summary <- WA_COND_IDs %>%
  filter(FORTYPCD %in% c(371, 201, 221, 222, 224, 225, 261, 922)) %>%
  group_by(plot_id) %>%
  summarise(
    all_zero = all(DSTRBCD1 == 0 & DSTRBCD2 == 0 & DSTRBCD3 == 0),
    .groups = "drop"
  ) %>%
  filter(all_zero) %>%
  select(plot_id)

WA_COND_treatment <- WA_COND_IDs %>%
  filter(
    plot_id %in% WA_prepost_burned_plot_id$plot_id |
      plot_id %in% WA_COND_control_summary$plot_id
  ) %>%
  mutate(
    condition = if_else(
      plot_id %in% WA_prepost_burned_plot_id$plot_id,
      "burned",
      "control"
    ),
    fire_code = case_when(
      condition == "control" ~ "0",
      DSTRBCD1 %in% c(30, 31, 32) ~ as.character(DSTRBCD1),
      DSTRBCD2 %in% c(30, 31, 32) ~ as.character(DSTRBCD2),
      DSTRBCD3 %in% c(30, 31, 32) ~ as.character(DSTRBCD3),
      TRUE ~ NA_character_
    )
  )

# Confirm unique plot counts
WA_COND_treatment %>%
  group_by(condition) %>%
  summarise(n_unique_plots = n_distinct(plot_id))

# Pull coordinates for these plots
WA_PLOT_coordinates <- WA_PLOT_IDs %>%
  semi_join(WA_COND_treatment, by = "plot_id") %>%
  distinct(plot_id, LAT, LON)

# convert lat/long points in a spatial object
WA_PLOT_coordinates_vect <- vect(
  WA_PLOT_coordinates, geom = c("LON", "LAT"), crs = "EPSG:4326")

# extract climate for all lat/long coordinates (all unique plot IDs)
WA_terra_climate_values <- terra::extract(
  terra_climate_stack,
  WA_PLOT_coordinates_vect
) %>%
  bind_cols(
    WA_PLOT_coordinates %>% select(plot_id, LAT, LON)
  ) %>%
  select(plot_id, LAT, LON, everything(), -ID)

WA_terra_climate_values <- WA_terra_climate_values %>%
  left_join(
    WA_COND_treatment %>%
      select(plot_id, condition, fire_code),
    by = "plot_id"
  )

points(
  WA_PLOT_coordinates_vect,
  col = ifelse(WA_terra_climate_values$condition == "burned", "red", "blue"),
  pch = 19
)

ggplot(
  WA_terra_climate_values,
  aes(
    x = LON,
    y = LAT,
    color = def_mean
  )
) +
  geom_point(
    size = 3
  ) +
  scale_color_viridis_c(
    name = "Climatic water deficit\n(mm)"
  ) +
  coord_equal() +
  theme_classic()
```

## climate and distance matrices
```{r}
WA_climate_standardized <- WA_terra_climate_values %>%
  mutate(across(all_of(climate_variables), ~ as.numeric(scale(.)), .names = "z_{.col}"))

WA_burned_climate  <- WA_climate_standardized %>% filter(condition == "burned")
WA_control_climate <- WA_climate_standardized %>% filter(condition == "control")

WA_control_climate_cleaned <- WA_control_climate %>% 
  filter(if_all(all_of(c(z_vars, "LAT", "LON")), ~ !is.na(.)))

WA_climate_distance_matrix <- rdist(
  as.matrix(WA_burned_climate[, z_vars]),
  as.matrix(WA_control_climate_cleaned[, z_vars])
)
dimnames(WA_climate_distance_matrix) <- list(WA_burned_climate$plot_id, WA_control_climate_cleaned$plot_id)

WA_geo_dist_mat <- distm(
  as.matrix(WA_burned_climate[, c("LON", "LAT")]),
  as.matrix(WA_control_climate_cleaned[, c("LON", "LAT")]),
  fun = distHaversine
) / 1000

dimnames(WA_geo_dist_mat) <- list(WA_burned_climate$plot_id, WA_control_climate_cleaned$plot_id)

WA_clim_norm <- norm01(WA_climate_distance_matrix)
WA_geo_norm  <- norm01(WA_geo_dist_mat)

WA_combined_dist <- w_clim * norm01(WA_climate_distance_matrix) +
  (1 - w_clim) * norm01(WA_geo_dist_mat)

n_match <- 5

WA_match_list <- lapply(seq_len(nrow(WA_combined_dist)), function(i) {
  ord <- order(WA_combined_dist[i, ])
  top <- ord[seq_len(n_match)]
  data.frame(
    burned_id = WA_burned_climate$plot_id[i],
    control_id = WA_control_climate$plot_id[top],
    combined_distance = WA_combined_dist[i, top],
    climate_distance = WA_climate_distance_matrix[i, top],
    geo_distance_km = WA_geo_dist_mat[i, top],
    rank = seq_len(n_match),
    row.names = NULL
  )
})

WA_matches <- bind_rows(WA_match_list)
head(WA_matches, 10)

WA_matches %>% count(control_id, sort = TRUE) %>% head(10)

WA_matched_controls <- WA_control_climate_cleaned %>% 
  filter(plot_id %in% WA_matches$control_id)

ggplot(WA_matches, aes(x = geo_distance_km, y = climate_distance)) +
  geom_point(alpha = 0.4) +
  labs(
    x = "Geographic distance (km)",
    y = "Climate distance (standardized)",
    title = "Climate distance vs. geographic distance for matched pairs"
  ) +
  theme_minimal()
```

## without replacement
```{r}
WA_matches_no_replacement <- without_replacement(
  dist_mat = WA_combined_dist,
  clim_mat = WA_climate_distance_matrix,
  geo_mat  = WA_geo_dist_mat,
  n_match  = 3
)

WA_matches_no_replacement %>% count(burned_id) %>% filter(n < 3)
```

## species overlap
```{r}

WA_burned_and_control_plot_id <- unique(c(WA_matches_no_replacement$burned_id, WA_matches_no_replacement$control_id))

WA_TREE_IDs_filtered <- WA_TREE_IDs %>%
  filter(plot_id %in% WA_burned_and_control_plot_id)

WA_TREE_species <- WA_TREE_IDs_filtered %>%
  left_join(
    REF_SPECIES %>%
      mutate(species = paste(GENUS, SPECIES, sep = "_")) %>%
      select(SPCD, species),
    by = "SPCD"
  ) %>%
  left_join(
    WA_COND_treatment %>% distinct(plot_id, condition),
    by = "plot_id"
  ) %>%
  left_join(
    WA_prepost_fire_summary %>% distinct(plot_id, fire_year),
    by = "plot_id"
  )

WA_TREE_species_filtered <- WA_TREE_species %>%
  filter(
    condition == "burned" | condition == "control"
  )

WA_species_summary_combined <- WA_TREE_species_filtered %>%
  group_by(species) %>%
  summarise(
    total_abundance = n(),
    n_plots = n_distinct(plot_id),
    .groups = "drop"
  ) %>%
  mutate(
    frequency = n_plots / n_distinct(WA_TREE_species_filtered$plot_id)
  ) %>%
  arrange(desc(total_abundance))

WA_species_summary_by_condition <- WA_TREE_species_filtered %>%
  group_by(condition, species) %>%
  summarise(
    total_abundance = n(),
    n_plots = n_distinct(plot_id),
    .groups = "drop"
  ) %>%
  group_by(condition) %>%
  mutate(
    frequency = n_plots / n_distinct(WA_TREE_species_filtered$plot_id[WA_TREE_species_filtered$condition == condition])
  ) %>%
  ungroup() %>%
  arrange(condition, desc(total_abundance))

WA_top_75_species <- WA_species_summary_combined %>%
  mutate(
    abundance_rank = rank(-total_abundance),
    frequency_rank = rank(-frequency),
    combined_rank = (abundance_rank + frequency_rank) / 2
  ) %>%
  arrange(combined_rank) %>%
  slice_head(n = 75)

WA_TREE_species_filtered %>%
  group_by(condition) %>%
  summarise(n_species = n_distinct(species), .groups = "drop")

WA_all_species <- WA_TREE_species_filtered %>%
  distinct(species) %>%
  pull(species)

length(WA_all_species) # 34 species

# to calculate dominance (the number of indiviudals of that species in that plot) and frequency (commonness)  (the number of plots the species occurs in / total number of plots)

WA_tree_count <- WA_TREE_species_filtered %>%
  group_by(plot_id) %>%
  summarise(total_trees_in_plot = n(), .groups = "drop")

WA_species_dominance <- WA_TREE_species_filtered %>%
  group_by(plot_id, species) %>%
  summarise(species_count = n(), .groups = "drop") %>%
  left_join(WA_tree_count, by = "plot_id") %>%
  mutate(relative_abundance = species_count / total_trees_in_plot)

WA_species_summary <- WA_species_dominance %>%
  group_by(species) %>%
  summarise(
    n_plots_present  = n_distinct(plot_id),
    avg_dominance    = mean(relative_abundance),
    median_dominance = median(relative_abundance),
    min_dominance    = min(relative_abundance),
    max_dominance    = max(relative_abundance),
    .groups = "drop"
  ) %>%
  mutate(frequency = n_plots_present / n_distinct(WA_TREE_species_filtered$plot_id)) %>%
  arrange(desc(frequency))

# is the number of species recruiting in control plots greater than burned (WA)
WA_SEEDLING_condition_filtered <- WA_SEEDLING_IDs %>%
  filter(plot_id %in% WA_burned_and_control_plot_id) %>%
  left_join(
    WA_COND_treatment %>% distinct(plot_id, condition),
    by = "plot_id"
  )

WA_SEEDLING_species_summary_condition <- WA_SEEDLING_condition_filtered %>%
  group_by(plot_id, condition, SPCD) %>%
  summarise(n_seedlings = sum(TREECOUNT, na.rm = TRUE), .groups = "drop") %>%
  left_join(REF_SPECIES %>% dplyr::select(SPCD, GENUS, SPECIES), by = "SPCD") %>%
  mutate(species = paste(GENUS, SPECIES, sep = "_"))

WA_SEEDLING_plot_summary_condition <- WA_SEEDLING_species_summary_condition %>%
  group_by(plot_id, condition) %>%
  summarise(
    n_seedling_species = n_distinct(SPCD),
    n_seedlings = sum(n_seedlings, na.rm = TRUE),
    .groups = "drop"
  )

WA_SEEDLING_plot_summary_condition %>%
  ggplot(aes(x = n_seedling_species, fill = condition)) +
  geom_histogram(binwidth = 1, color = "white", position = "identity", alpha = 0.6) +
  facet_wrap(~condition, ncol = 1) +
  theme_bw() +
  labs(
    x = "Number of seedling species per plot",
    y = "Number of plots",
    title = "Seedling species richness: burned vs. control (WA)"
  )

WA_SEEDLING_plot_summary_condition %>%
  ggplot(aes(x = condition, y = n_seedling_species, fill = condition)) +
  geom_boxplot(alpha = 0.6) +
  theme_bw() +
  labs(
    x = NULL,
    y = "Number of seedling species per plot",
    title = "Seedling species richness: burned vs. control (WA)"
  )

WA_seedling_richness_summary_by_condition <- WA_SEEDLING_plot_summary_condition %>%
  group_by(condition) %>%
  summarise(
    mean_richness   = mean(n_seedling_species),
    median_richness = median(n_seedling_species),
    n_plots         = n(),
    .groups = "drop"
  )

WA_seedling_richness_summary_by_condition

WA_SEEDLING_species_summary_condition %>%
  group_by(species, condition, plot_id) %>%
  summarise(n_seedlings = sum(n_seedlings), .groups = "drop") %>%
  group_by(species, condition) %>%
  summarise(mean_per_plot = mean(n_seedlings), .groups = "drop") %>%
  group_by(condition) %>%
  slice_max(mean_per_plot, n = 20) %>%
  ungroup() %>%
  ggplot(aes(x = reorder_within(species, mean_per_plot, condition), y = mean_per_plot, fill = condition)) +
  geom_col() +
  coord_flip() +
  scale_x_reordered() +
  facet_wrap(~condition, scales = "free_y") +
  labs(
    x = "species",
    y = "average # seedlings per plot",
    title = "Average seedling abundance per species per plot: burned vs. control (WA)"
  ) +
  theme_bw()

WA_SEEDLING_species_summary_condition %>%
  group_by(plot_id, condition, species) %>%
  summarise(n_seedlings = sum(n_seedlings), .groups = "drop") %>%
  mutate(n_seedlings_capped = pmin(n_seedlings, 50)) %>%
  ggplot(aes(x = species, y = n_seedlings_capped, fill = condition)) +
  geom_boxplot(alpha = 0.6) +
  coord_flip() +
  facet_wrap(~condition, scales = "free_y") +
  labs(
    x = "species",
    y = "seedlings per plot (capped at 50)",
    title = "Seedling abundance per species per plot: burned vs. control (WA)"
  ) +
  theme_bw()

summary(WA_SEEDLING_plot_summary_condition$n_seedling_species[WA_SEEDLING_plot_summary_condition$condition == "burned"])
summary(WA_SEEDLING_plot_summary_condition$n_seedling_species[WA_SEEDLING_plot_summary_condition$condition == "control"])

```

## whole plot network
```{r}

# filter CA_COND_IDs for forest types only, no disturbance filtering
WA_COND_forest_type <- WA_COND_IDs %>%
  filter(FORTYPCD %in% c(201, 202, 221, 222, 224, 225, 226, 241, 261, 262, 263, 264, 265, 267, 268, 270, 271, 281, 301, 304, 305, 321, 341, 342, 361, 365, 371, 722, 911, 912, 922, 923, 933, 941, 961))
         
         
(371, 201, 221, 222, 224, 225, 261, 922))

# get the unique plot_ids that qualify
WA_forest_type_plot_id <- WA_COND_forest_type %>%
  distinct(plot_id) %>%
  pull(plot_id)

length(WA_forest_type_plot_id)

# filter CA_PLOT_IDs for those plot_ids
WA_PLOT_forest_type <- WA_PLOT_IDs %>%
  filter(plot_id %in% WA_forest_type_plot_id)

# filter CA_TREE_IDs for those plot_ids
WA_TREE_forest_type <- WA_TREE_IDs %>%
  filter(plot_id %in% WA_forest_type_plot_id)

nrow(WA_PLOT_forest_type)
n_distinct(WA_TREE_forest_type$plot_id)

WA_TREE_forest_type_species <- WA_TREE_forest_type %>%
  left_join(
    REF_SPECIES %>%
      mutate(species = paste(GENUS, SPECIES, sep = "_")) %>%
      select(SPCD, species),
    by = "SPCD"
  )
sum(is.na(WA_TREE_forest_type_species$species))
n_distinct(WA_TREE_forest_type_species$species) # 49 species

# how different are the species from this workflow and the previous control plot filtering by climate?

# species from your burned/control workflow (already built)
WA_all_species   # from earlier: distinct species across CA_TREE_species_filtered

# species from your new forest-type-only workflow
WA_forest_type_all_species <- WA_TREE_forest_type_species %>%
  distinct(species) %>%
  pull(species)

length(WA_forest_type_all_species)

# species found in both workflows
intersect(WA_all_species, WA_forest_type_all_species)

# species found only in the burned/control workflow
setdiff(WA_all_species, WA_forest_type_all_species)

# species found only in the forest-type-only workflow
setdiff(WA_forest_type_all_species, WA_all_species) # 15 species more


# make sure the species coming out of SEEDLING match TREE
# filter CA_SEEDLING_IDs for those plot_ids
WA_SEEDLING_forest_type <- WA_SEEDLING_IDs %>%
  filter(plot_id %in% WA_forest_type_plot_id)

WA_SEEDLING_forest_type_species <- WA_SEEDLING_forest_type %>%
  left_join(
    REF_SPECIES %>%
      mutate(species = paste(GENUS, SPECIES, sep = "_")) %>%
      select(SPCD, species),
    by = "SPCD"
  )

# species list from seedling data (forest-type-only workflow)
WA_forest_type_all_seedling_species <- WA_SEEDLING_forest_type_species %>%
  distinct(species) %>%
  pull(species)

length(WA_forest_type_all_species)
length(WA_forest_type_all_seedling_species)

# species found in both tree and seedling data
intersect(WA_forest_type_all_species, WA_forest_type_all_seedling_species)

# species found only in tree data (mature trees, no seedlings recorded)
setdiff(WA_forest_type_all_species, WA_forest_type_all_seedling_species)

# species found only in seedling data (recruiting, but no mature trees recorded)
setdiff(WA_forest_type_all_seedling_species, WA_forest_type_all_species)

# create a shared species list
WA_forest_type_all_species
WA_forest_type_all_seedling_species

WA_species <- data.frame(
  species = unique(c(
    WA_forest_type_all_species,
    WA_forest_type_all_seedling_species
  ))
)

CA_OR_WA_forest_type_species_lists <- list(
  CA = CA_species,
  OR = OR_species,
  WA = WA_species
)

# species found in ALL THREE states
CA_OR_WA_forest_type_species <- Reduce(intersect, CA_OR_WA_forest_type_species_lists)
length(unique(CA_OR_WA_forest_type_species$species)) 
# 40 species shared in all 3 states

# total unique species across all three states combined
CA_OR_WA_forest_type_species_overlap <- Reduce(union, CA_OR_WA_forest_type_species_lists)
length(CA_OR_WA_forest_type_species_overlap$species) 
# 100 total unique species across all three states

write.csv(CA_OR_WA_forest_type_species_overlap, "updated_species_list_07182026.csv", row.names = FALSE)

# species found in exactly one state only (not shared with either other state)
length(setdiff(CA_species$species, union(OR_species$species, WA_species$species))) # 21 in CA only
length(setdiff(OR_species$species, union(CA_species$species, WA_species$species))) # 11 in OR only
length(setdiff(WA_species$species, union(CA_species$species, OR_species$species))) # 5 in WA only

# Shared by CA and OR, but NOT WA
setdiff(intersect(CA_species$species, OR_species$species), WA_species$species)
# 15 species

# Shared by CA and WA, but NOT OR
setdiff(intersect(CA_species$species, WA_species$species), OR_species$species)
# 1 species

# Shared by OR and WA, but NOT CA
setdiff(intersect(OR_species$species, WA_species$species), CA_species$species)
# 7 species 

# 100 unique species across all states, 40 shared and 37 unique

# Install and load the package if you haven't already
install.packages("ggvenn")
library(ggvenn)

# Combine your separate state lists into a single named list
venn_data <- list(
  California = CA_species$species,
  Oregon     = OR_species$species,
  Washington = WA_species$species
)

# Plot the diagram
ggvenn(venn_data, fill_color = c("#3A86FF", "#FF006E", "#FFBE0B"))

```


# CA, WA, OR species list

```{r}
CA_OR_WA_species_lists <- list(
  CA = CA_all_species,
  OR = OR_all_species,
  WA = WA_all_species
)

# species found in ALL THREE states
CA_OR_WA_species <- Reduce(intersect, CA_OR_WA_species_lists)
length(CA_OR_WA_species)

# total unique species across all three states combined
CA_OR_WA_species_overlap <- Reduce(union, CA_OR_WA_species_lists)
length(CA_OR_WA_species_overlap)

# species found in exactly one state only (not shared with either other state)
length(setdiff(CA_all_species, union(OR_all_species, WA_all_species)))
length(setdiff(OR_all_species, union(CA_all_species, WA_all_species)))
length(setdiff(WA_all_species, union(CA_all_species, OR_all_species)))

# 78 unique species across all states, 48 shared and 30 unique


# are burned plot species much different than control plot species
# CA
CA_species_burned  <- CA_TREE_species_filtered %>% filter(condition == "burned")  %>% distinct(species) %>% pull(species)
CA_species_control <- CA_TREE_species_filtered %>% filter(condition == "control") %>% distinct(species) %>% pull(species)

length(CA_species_burned) # 49 species
length(CA_species_control) # 60 species
length(intersect(CA_species_burned, CA_species_control)) # 44 species shared
length(setdiff(CA_species_burned, CA_species_control)) # 5 in burned only
length(setdiff(CA_species_control, CA_species_burned)) # 16 in control-only

# OR
OR_species_burned  <- OR_TREE_species_filtered %>% filter(condition == "burned")  %>% distinct(species) %>% pull(species)
OR_species_control <- OR_TREE_species_filtered %>% filter(condition == "control") %>% distinct(species) %>% pull(species)

length(OR_species_burned) # 44 species
length(OR_species_control) # 51 species
length(intersect(OR_species_burned, OR_species_control)) # 42 species shared
length(setdiff(OR_species_burned, OR_species_control)) # 2 in burned only
length(setdiff(OR_species_control, OR_species_burned)) # 9 in control only

# WA
WA_species_burned  <- WA_TREE_species_filtered %>% filter(condition == "burned")  %>% distinct(species) %>% pull(species)
WA_species_control <- WA_TREE_species_filtered %>% filter(condition == "control") %>% distinct(species) %>% pull(species)

length(WA_species_burned) # 25 species
length(WA_species_control) # 31 species
length(intersect(WA_species_burned, WA_species_control)) # 23 species shared
length(setdiff(WA_species_burned, WA_species_control)) # 2 in burned only
length(setdiff(WA_species_control, WA_species_burned)) # 8 in control only 
```


# species list
```{r}
# now lets aggregate a species list for each state
# STATE_TREE links to STAT_COND via the common identifier PLT_CN
# take the PLT_CN codes from forest_type_combined and filter STATE_TREE
CA_TREE_fire_only <- CA_TREE %>%
  filter(PLT_CN %in% CA_forest_type_combined$PLT_CN)

WA_TREE_fire_only <- WA_TREE %>%
  filter(PLT_CN %in% WA_forest_type_combined$PLT_CN)

OR_TREE_fire_only <- OR_TREE %>%
  filter(PLT_CN %in% OR_forest_type_combined$PLT_CN)

ID_TREE_fire_only <- ID_TREE %>%
  filter(PLT_CN %in% ID_forest_type_combined$PLT_CN)

MT_TREE_fire_only <- MT_TREE %>%
  filter(PLT_CN %in% MT_forest_type_combined$PLT_CN)

CO_TREE_fire_only <- CO_TREE %>%
  filter(PLT_CN %in% CO_forest_type_combined$PLT_CN)

NV_TREE_fire_only <- NV_TREE %>%
  filter(PLT_CN %in% NV_forest_type_combined$PLT_CN)

UT_TREE_fire_only <- UT_TREE %>%
  filter(PLT_CN %in% UT_forest_type_combined$PLT_CN)

WY_TREE_fire_only <- WY_TREE %>%
  filter(PLT_CN %in% WY_forest_type_combined$PLT_CN)

# attach SCIENTIFIC_NAME from REF_SPECIES to SPCD in CA_TREE_fire_only_spp_ID
CA_TREE_fire_only_spp_ID <- CA_TREE_fire_only %>%
  left_join(REF_SPECIES %>% dplyr::select(SPCD, SCIENTIFIC_NAME), by = "SPCD")

WA_TREE_fire_only_spp_ID <- WA_TREE_fire_only %>%
  left_join(REF_SPECIES %>% dplyr::select(SPCD, SCIENTIFIC_NAME), by = "SPCD")

OR_TREE_fire_only_spp_ID <- OR_TREE_fire_only %>%
  left_join(REF_SPECIES %>% dplyr::select(SPCD, SCIENTIFIC_NAME), by = "SPCD")

ID_TREE_fire_only_spp_ID <- ID_TREE_fire_only %>%
  left_join(REF_SPECIES %>% dplyr::select(SPCD, SCIENTIFIC_NAME), by = "SPCD")

MT_TREE_fire_only_spp_ID <- MT_TREE_fire_only %>%
  left_join(REF_SPECIES %>% dplyr::select(SPCD, SCIENTIFIC_NAME), by = "SPCD")

CO_TREE_fire_only_spp_ID <- CO_TREE_fire_only %>%
  left_join(REF_SPECIES %>% dplyr::select(SPCD, SCIENTIFIC_NAME), by = "SPCD")

NV_TREE_fire_only_spp_ID <- NV_TREE_fire_only %>%
  left_join(REF_SPECIES %>% dplyr::select(SPCD, SCIENTIFIC_NAME), by = "SPCD")

UT_TREE_fire_only_spp_ID <- UT_TREE_fire_only %>%
  left_join(REF_SPECIES %>% dplyr::select(SPCD, SCIENTIFIC_NAME), by = "SPCD")

WY_TREE_fire_only_spp_ID <- WY_TREE_fire_only %>%
  left_join(REF_SPECIES %>% dplyr::select(SPCD, SCIENTIFIC_NAME), by = "SPCD")

# create a list of tree species in fire only plots
CA_TREE_fire_only_spp_list <- unique(CA_TREE_fire_only_spp_ID$SCIENTIFIC_NAME)

WA_TREE_fire_only_spp_list <- unique(WA_TREE_fire_only_spp_ID$SCIENTIFIC_NAME)

OR_TREE_fire_only_spp_list <- unique(OR_TREE_fire_only_spp_ID$SCIENTIFIC_NAME)

ID_TREE_fire_only_spp_list <- unique(ID_TREE_fire_only_spp_ID$SCIENTIFIC_NAME)

MT_TREE_fire_only_spp_list <- unique(MT_TREE_fire_only_spp_ID$SCIENTIFIC_NAME)

CO_TREE_fire_only_spp_list <- unique(CO_TREE_fire_only_spp_ID$SCIENTIFIC_NAME)

NV_TREE_fire_only_spp_list <- unique(NV_TREE_fire_only_spp_ID$SCIENTIFIC_NAME)

UT_TREE_fire_only_spp_list <- unique(UT_TREE_fire_only_spp_ID$SCIENTIFIC_NAME)

WY_TREE_fire_only_spp_list <- unique(WY_TREE_fire_only_spp_ID$SCIENTIFIC_NAME)

# can we now take the species lists for each state and find the intersection?
#(i.e. we want a list of total unique species across states)

# merge all states species lists and keep only unique species
total_unique_species <- unique(c(
  CA_TREE_fire_only_spp_list,
  WA_TREE_fire_only_spp_list,
  OR_TREE_fire_only_spp_list,
  ID_TREE_fire_only_spp_list,
  MT_TREE_fire_only_spp_list,
  CO_TREE_fire_only_spp_list,
  NV_TREE_fire_only_spp_list,
  UT_TREE_fire_only_spp_list,
  WY_TREE_fire_only_spp_list
))

write.csv(total_unique_species, file = "fire-severity-unique-species-list.csv")
```


# plot locations/distribution
```{r}
# try plotting locations on a map of the US
# color code by burn severity to see the spatial distribution of high and low severity

#extract plot locations from CA_PLOT or 'CA' for the PLT_CN's on CA_fire_only

CA_burned_plot_coordinates <- CA_TREE_fire_only %>%
  left_join(
    CA %>% dplyr::select(CN, LAT, LON),
    by = c("PLT_CN" = "CN")
  )


# Convert to sf object using LAT and LON columns
CA_burned_sf <- st_as_sf(
  CA_burned_plot_coordinates,
  coords = c("LON", "LAT"),  # Order matters: (lon, lat)
  crs = 4326  # WGS84 coordinate system
)

# Get US states shapefile
usa_shapefile <- ne_states(country = "United States of America", returnclass = "sf")

# Filter for western states
western_states <- usa_shapefile %>%
  filter(name %in% c(
    "California", "Oregon", "Washington", "Nevada", "Idaho", "Utah",
    "Arizona", "New Mexico", "Montana", "Wyoming", "Colorado"
  ))


# Join disturbance code into your plots dataframe
CA_burned_plot_coordinates_DSTRBCD1 <- CA_burned_plot_coordinates %>%
  left_join(
    forest_type_combined %>% dplyr::select(PLT_CN, DSTRBCD1, DSTRBYR1),
    by = "PLT_CN"
  )

# Convert to sf with disturbance code included
CA_burned_sf <- st_as_sf(
  CA_burned_plot_coordinates_DSTRBCD1,
  coords = c("LON", "LAT"),
  crs = 4326
)

ggplot() +
  geom_sf(data = western_states, fill = "gray95", color = "black") +
  geom_sf(
    data = CA_burned_sf %>% filter(DSTRBCD1 %in% c(31, 32)),
    aes(color = factor(DSTRBCD1)),
    size = 1, alpha = 0.8
  ) +
  scale_color_manual(
    values = c("31" = "orange", "32" = "red"),
    labels = c("31" = "Disturbance 31", "32" = "Disturbance 32"),
    name = "Disturbance Code"
  ) +
  coord_sf(xlim = c(-125, -100), ylim = c(30, 45)) +
  theme_minimal() +
  labs(
    title = "Burned plots in low-moderate and high severity",
    x = "Longitude",
    y = "Latitude"
  )

# lets add other states now

# 1. Join coordinates from the state-level plot table
OR_burned_plot_coordinates <- OR_TREE_fire_only %>%
  left_join(
    OR %>% dplyr::select(CN, LAT, LON),
    by = c("PLT_CN" = "CN")
  )

# 2. Join in the disturbance code from forest_type_combined
OR_burned_plot_coordinates_DSTRBCD1 <- OR_burned_plot_coordinates %>%
  left_join(
    OR_forest_type_combined %>% dplyr::select(PLT_CN, DSTRBCD1, DSTRBYR1),
    by = "PLT_CN"
  )

# 3. Convert to sf object with disturbance info
OR_burned_sf <- st_as_sf(
  OR_burned_plot_coordinates_DSTRBCD1,
  coords = c("LON", "LAT"),
  crs = 4326
)


# 1. Join coordinates from the state-level plot table
WA_burned_plot_coordinates <- WA_TREE_fire_only %>%
  left_join(
    WA %>% dplyr::select(CN, LAT, LON),
    by = c("PLT_CN" = "CN")
  )

# 2. Join in the disturbance code from forest_type_combined
WA_burned_plot_coordinates_DSTRBCD1 <- WA_burned_plot_coordinates %>%
  left_join(
    WA_forest_type_combined %>% dplyr::select(PLT_CN, DSTRBCD1, DSTRBYR1),
    by = "PLT_CN"
  )

# 3. Convert to sf object with disturbance info
WA_burned_sf <- st_as_sf(
  WA_burned_plot_coordinates_DSTRBCD1,
  coords = c("LON", "LAT"),
  crs = 4326
)

# 1. Join coordinates from the state-level plot table
ID_burned_plot_coordinates <- ID_TREE_fire_only %>%
  left_join(
    ID %>% dplyr::select(CN, LAT, LON),
    by = c("PLT_CN" = "CN")
  )

# 2. Join in the disturbance code from forest_type_combined
ID_burned_plot_coordinates_DSTRBCD1 <- ID_burned_plot_coordinates %>%
  left_join(
    ID_forest_type_combined %>% dplyr::select(PLT_CN, DSTRBCD1),
    by = "PLT_CN"
  )

# 3. Convert to sf object with disturbance info
ID_burned_sf <- st_as_sf(
  ID_burned_plot_coordinates_DSTRBCD1,
  coords = c("LON", "LAT"),
  crs = 4326
)


# 1. Join coordinates from the state-level plot table
MT_burned_plot_coordinates <- MT_TREE_fire_only %>%
  left_join(
    MT %>% dplyr::select(CN, LAT, LON),
    by = c("PLT_CN" = "CN")
  )

# 2. Join in the disturbance code from forest_type_combined
MT_burned_plot_coordinates_DSTRBCD1 <- MT_burned_plot_coordinates %>%
  left_join(
    MT_forest_type_combined %>% dplyr::select(PLT_CN, DSTRBCD1),
    by = "PLT_CN"
  )

# 3. Convert to sf object with disturbance info
MT_burned_sf <- st_as_sf(
  MT_burned_plot_coordinates_DSTRBCD1,
  coords = c("LON", "LAT"),
  crs = 4326
)

# 1. Join coordinates from the state-level plot table
CO_burned_plot_coordinates <- CO_TREE_fire_only %>%
  left_join(
    CO %>% dplyr::select(CN, LAT, LON),
    by = c("PLT_CN" = "CN")
  )

# 2. Join in the disturbance code from forest_type_combined
CO_burned_plot_coordinates_DSTRBCD1 <- CO_burned_plot_coordinates %>%
  left_join(
    CO_forest_type_combined %>% dplyr::select(PLT_CN, DSTRBCD1),
    by = "PLT_CN"
  )

# 3. Convert to sf object with disturbance info
CO_burned_sf <- st_as_sf(
  CO_burned_plot_coordinates_DSTRBCD1,
  coords = c("LON", "LAT"),
  crs = 4326
)

# 1. Join coordinates from the state-level plot table
UT_burned_plot_coordinates <- UT_TREE_fire_only %>%
  left_join(
    UT %>% dplyr::select(CN, LAT, LON),
    by = c("PLT_CN" = "CN")
  )

# 2. Join in the disturbance code from forest_type_combined
UT_burned_plot_coordinates_DSTRBCD1 <- UT_burned_plot_coordinates %>%
  left_join(
    UT_forest_type_combined %>% dplyr::select(PLT_CN, DSTRBCD1),
    by = "PLT_CN"
  )

# 3. Convert to sf object with disturbance info
UT_burned_sf <- st_as_sf(
  UT_burned_plot_coordinates_DSTRBCD1,
  coords = c("LON", "LAT"),
  crs = 4326
)


# 1. Join coordinates from the state-level plot table
WY_burned_plot_coordinates <- WY_TREE_fire_only %>%
  left_join(
    WY %>% dplyr::select(CN, LAT, LON),
    by = c("PLT_CN" = "CN")
  )

# 2. Join in the disturbance code from forest_type_combined
WY_burned_plot_coordinates_DSTRBCD1 <- WY_burned_plot_coordinates %>%
  left_join(
    WY_forest_type_combined %>% dplyr::select(PLT_CN, DSTRBCD1),
    by = "PLT_CN"
  )

# 3. Convert to sf object with disturbance info
WY_burned_sf <- st_as_sf(
  WY_burned_plot_coordinates_DSTRBCD1,
  coords = c("LON", "LAT"),
  crs = 4326
)

###### this chunk for NV didnt work, see below instead
# 1. Join coordinates from the state-level plot table
NV_burned_plot_coordinates <- NV_TREE_fire_only %>%
  left_join(
    NV %>% select(CN, LAT, LON),
    by = c("PLT_CN" = "CN")
  )

# 2. Join in the disturbance code from forest_type_combined
NV_burned_plot_coordinates_DSTRBCD1 <- NV_burned_plot_coordinates %>%
  left_join(
    NV_forest_type_combined %>% select(PLT_CN, DSTRBCD1),
    by = "PLT_CN"
  )

# 3. Convert to sf object with disturbance info
NV_burned_sf <- st_as_sf(
  NV_burned_plot_coordinates_DSTRBCD1,
  coords = c("LON", "LAT"),
  crs = 4326
)

######
NV_fire_coords <- NV_fire_only %>%
  left_join(NV %>% dplyr::select(CN, LAT, LON), by = c("PLT_CN" = "CN"))

NV_fire_sf <- st_as_sf(NV_fire_coords, coords = c("LON", "LAT"), crs = 4326)



NV_burned_plot_coordinates <- NV_fire_only %>%
  left_join(NV %>% dplyr::select(CN, LAT, LON), by = c("PLT_CN" = "CN"))

# Step 3: Convert to sf object
NV_burned_sf <- st_as_sf(NV_burned_plot_coordinates, coords = c("LON", "LAT"), crs = 4326)

normalize_types <- function(x) {
  dplyr::mutate(
    x,
    dplyr::across(
      where(~ !inherits(.x, "sfc")),
      as.character
    )
  )
}



all_burned_sf <- dplyr::bind_rows(
  normalize_types(CA_burned_sf),
  normalize_types(OR_burned_sf),
  normalize_types(WA_burned_sf),
  normalize_types(ID_burned_sf),
  normalize_types(MT_burned_sf),
  normalize_types(CO_burned_sf),
  normalize_types(UT_burned_sf),
  normalize_types(NV_burned_sf),
  normalize_types(WY_burned_sf)
)




ggplot() +
  geom_sf(data = western_states, fill = "gray95", color = "black") +
  geom_sf(
    data = all_burned_sf %>% filter(DSTRBCD1 %in% c(31, 32)),
    aes(color = factor(DSTRBCD1)),
    size = 1, alpha = 0.8
  ) +
  scale_color_manual(
    values = c("31" = "orange", "32" = "red"),
    name = "Disturbance Code"
  ) +
  coord_sf(xlim = c(-125, -100), ylim = c(30, 50)) +
  theme_minimal() +
  labs(
    x = "Longitude",
    y = "Latitude"
  )
```
# plot level summary stats
```{r}
# lets visualize some summary stats about our burned and unburned plots

h1 <- ggplot(
  data = CA_burned_plot_coordinates_DSTRBCD1 %>%
    filter(DSTRBYR1 <= 2025) %>%
    group_by(PLT_CN) %>%
    slice_min(DSTRBYR1, with_ties = FALSE) %>%
    ungroup(),  # keep one row per unique plot
  aes(x = DSTRBYR1)
) +
  geom_histogram(binwidth = 1, boundary = 1990, color = "black", fill = "steelblue") +
  scale_x_continuous(breaks = seq(1990, 2025, by = 1)) +
  labs(
    title = "Histogram of Disturbance Years in CA (using first year fire was recorded)",
    x = "Disturbance Year",
    y = "Count (Unique PLT_CN)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

h1

h2 <- ggplot(
  data = OR_burned_plot_coordinates_DSTRBCD1 %>%
    filter(DSTRBYR1 >= 1990, DSTRBYR1 <= 2025) %>%
    group_by(PLT_CN) %>%
    slice_min(DSTRBYR1, with_ties = FALSE) %>%
    ungroup(),
  aes(x = DSTRBYR1)
) +
  geom_histogram(binwidth = 1, boundary = 1990, color = "black", fill = "steelblue") +
  scale_x_continuous(breaks = seq(1990, 2025, by = 1)) +
  labs(
    title = "Histogram of Disturbance Years in OR (using first year fire was recorded)",
    x = "Disturbance Year",
    y = "Count (Unique Plots)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))


h2



h3 <- ggplot(
  data = WA_burned_plot_coordinates_DSTRBCD1 %>%
    filter(DSTRBYR1 <= 2025) %>%
    group_by(PLT_CN) %>%
    slice_min(DSTRBYR1, with_ties = FALSE) %>%
    ungroup(),  # keep one row per unique plot
  aes(x = DSTRBYR1)
) +
  geom_histogram(binwidth = 1, boundary = 1990, color = "black", fill = "steelblue") +
  scale_x_continuous(breaks = seq(1990, 2025, by = 1)) +
  labs(
    title = "Histogram of Disturbance Years in WA (using first year fire was recorded)",
    x = "Disturbance Year",
    y = "Count (Unique Plots)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))



h3
grid.arrange (h1, h2, h3)

h4 <- ggplot(
  data = CA_burned_plot_coordinates_DSTRBCD1 %>%
    filter(!is.na(DSTRBYR1), DSTRBYR1 >= 1990, DSTRBYR1 <= 2025) %>%
    group_by(PLT_CN) %>%
    slice_min(DSTRBYR1, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(TimeSinceFire = 2025 - DSTRBYR1),
  aes(x = TimeSinceFire)
) +
  geom_histogram(binwidth = 1, boundary = 0, color = "black", fill = "darkorange") +
  scale_x_continuous(breaks = seq(0, 35, by = 1)) +
  labs(
    title = "CA",
    x = "Time Since Fire (Years)",
    y = "Count (Unique Plots)"
  ) +
  theme_minimal()
h4

h5 <- ggplot(
  data = OR_burned_plot_coordinates_DSTRBCD1 %>%
    filter(!is.na(DSTRBYR1), DSTRBYR1 >= 1990, DSTRBYR1 <= 2025) %>%
    group_by(PLT_CN) %>%
    slice_min(DSTRBYR1, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(TimeSinceFire = 2025 - DSTRBYR1),
  aes(x = TimeSinceFire)
) +
  geom_histogram(binwidth = 1, boundary = 0, color = "black", fill = "darkorange") +
  scale_x_continuous(breaks = seq(0, 35, by = 1)) +
  labs(
    title = "OR",
    x = "Time Since Fire (Years)",
    y = "Count (Unique Plots)"
  ) +
  theme_minimal()

h5

h6 <- ggplot(
  data = WA_burned_plot_coordinates_DSTRBCD1 %>%
    filter(!is.na(DSTRBYR1), DSTRBYR1 >= 1990, DSTRBYR1 <= 2025) %>%
    group_by(PLT_CN) %>%
    slice_min(DSTRBYR1, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(TimeSinceFire = 2025 - DSTRBYR1),
  aes(x = TimeSinceFire)
) +
  geom_histogram(binwidth = 1, boundary = 0, color = "black", fill = "darkorange") +
  scale_x_continuous(breaks = seq(0, 35, by = 1)) +
  labs(
    title = "WA",
    x = "Time Since Fire (Years)",
    y = "Count (Unique Plots)"
  ) +
  theme_minimal()

h6

grid.arrange (h4, h5, h6)


# 600+ plots in ground fire and 400+ plots in canopy fire in CA
ggplot(dplyr::filter(CA_cond, DSTRBCD1 %in% c(30, 31, 32)),
       aes(x = factor(DSTRBCD1))) +
  geom_bar() +
  labs(x = "Disturbance Code", y = "Count",
       title = "Fire-Related Disturbances (Codes 30–32)") +
  theme_minimal()
```

# climate extraction 
```{r}
```

# understory
```{r}
# lets look at the understory too...

# attach SCIENTIFIC_NAME from REF_SPECIES to CA_veg_understory_fire_only by matching VEG_SPCD
# to SPECIES_SYMBOL
CA_veg_understory_fire_only_spp_ID <- CA_veg_understory_fire_only %>%
  left_join(REF_SPECIES %>%
              select(SPECIES_SYMBOL, SCIENTIFIC_NAME),
            by = c("VEG_SPCD" = "SPECIES_SYMBOL")) ## this has the be redone, only gives trees

# take the PLT_CN codes from forest_type_combined and filter CA_veg_understory to those PLT_CN's
CA_veg_understory <- CA_P2VEG_SUBPLOT_SPP

CA_veg_understory_fire_only <- CA_veg_understory %>%
  filter(PLT_CN %in% forest_type_combined$PLT_CN)

# CA_veg_understory_fire_only uses VEG_SPCD instead of SPCD



# create a list of understory species in fire only plots



# how many resurveys do fire only mixed conifer plots in CA have?
#CA_PLOT (or just CA here) links to CA_COND via the common identifier PLT_CN


#If INVYR (plot visit year) < DSTRBYR1 → then this visit happened before the fire.

#If INVYR > DSTRBYR1 → then this visit was after the fire.

#If equal, the fire occurred in the same year as the plot visit

# if we want to link CA_PLOT (or CA) with CA_cond that has been filtered by disturbance codes
# and forest types then we want to lin forest_type_combined with CA_PLOT or CA but PLT_CN in
#CA_PLOT of CA PLT_CN is the same as CN

inventory_year <- CA_cond %>%
  filter(DSTRBCD1 %in% c(31,32) | DSTRBCD2 %in% c(31,32) | DSTRBCD3 %in% c(31,32)) %>%
  left_join(CA, by = "PLT_CN") %>%
  mutate(fire_year = coalesce(DSTRBYR1, DSTRBYR2, DSTRBYR3),
         pre_fire = INVYR < fire_year,
         post_fire  = INVYR > fire_year)


resurveyed_fire_only_plots <- CA_TREE_fire_only %>%
  group_by(PLT_CN) %>%
  summarise(n_years = n_distinct(INVYR))
# all plots only have 1 visit?

resurveyed_fire_only_plots <- CA_TREE_fire_only %>%
  group_by(PLT_CN) %>%
  summarise(INVYR_list = list(sort(unique(INVYR))))


```



