#load packages
```{r}
library(sf)
library(terra)
library(dplyr)
library(raster)
library(terra)
install.packages('tidyverse')
library(tidyverse)
library(here)
library(rnaturalearth)
library(rnaturalearthdata)
install.packages('rnaturalearthhires')
install.packages("remotes")
remotes::install_github("ropensci/rnaturalearthhires")
library(gridExtra)
```
#load data
```{r}
# load in FIA tree data for western North America (California, Oregon, Washington, Colorado, Utah, Nevada, Wyoming) to encompass the Sierra Nevada, Northern and Southern Coast Ranges, maybe also include New Mexico and Montana.

# Load points for fuzzed FIA plots
CA_PLOT <- read.csv("./data/FIA/CA_PLOT.csv")
CO_PLOT <- read.csv("./data/FIA/CO_PLOT.csv")
NV_PLOT <- read.csv("./data/FIA/NV_PLOT.csv")
OR_PLOT <- read.csv("./data/FIA/OR_PLOT.csv")
UT_PLOT <- read.csv("./data/FIA/UT_PLOT.csv")
WA_PLOT <- read.csv("./data/FIA/WA_PLOT.csv")
WY_PLOT <- read.csv("./data/FIA/WY_PLOT.csv")
ID_PLOT <- read.csv("./data/FIA/ID_PLOT.csv")
MT_PLOT <- read.csv("./data/FIA/MT_PLOT.csv")

# then load in condition code files for each state
CA_COND <- read.csv("./data/FIA/CA_COND.csv")
CO_COND <- read.csv("./data/FIA/CO_COND.csv")
NV_COND <- read.csv("./data/FIA/NV_COND.csv")
OR_COND <- read.csv("./data/FIA/OR_COND.csv")
UT_COND <- read.csv("./data/FIA/UT_COND.csv")
WA_COND <- read.csv("./data/FIA/WA_COND.csv")
WY_COND <- read.csv("./data/FIA/WY_COND.csv")
ID_COND <- read.csv("./data/FIA/ID_COND.csv")
MT_COND <- read.csv("./data/FIA/MT_COND.csv")

# then load in TREE level data for each state
CA_TREE <- read.csv("./data/FIA/CA_TREE.csv")
CO_TREE <- read.csv("./data/FIA/CO_TREE.csv")
NV_TREE <- read.csv("./data/FIA/NV_TREE.csv")
OR_TREE <- read.csv("./data/FIA/OR_TREE.csv")
UT_TREE <- read.csv("./data/FIA/UT_TREE.csv")
WA_TREE <- read.csv("./data/FIA/WA_TREE.csv")
WY_TREE <- read.csv("./data/FIA/WY_TREE.csv")
ID_TREE <- read.csv("./data/FIA/ID_TREE.csv")
MT_TREE <- read.csv("./data/FIA/MT_TREE.csv")

CA_SEEDLING <- read.csv("./data/FIA/CA_SEEDLING.csv")
CO_SEEDLING <- read.csv("./data/FIA/CO_SEEDLING.csv")
NV_SEEDLING <- read.csv("./data/FIA/NV_SEEDLING.csv")
OR_SEEDLING <- read.csv("./data/FIA/OR_SEEDLING.csv")
UT_SEEDLING <- read.csv("./data/FIA/UT_SEEDLING.csv")
WA_SEEDLING <- read.csv("./data/FIA/WA_SEEDLING.csv")
WY_SEEDLING <- read.csv("./data/FIA/WY_SEEDLING.csv")
ID_SEEDLING <- read.csv("./data/FIA/ID_SEEDLING.csv")
MT_SEEDLING <- read.csv("./data/FIA/MT_SEEDLING.csv")


# Load in reference species names
REF_SPECIES <- read.csv("./data/FIA/REF_SPECIES.csv")
```

# filter by disturbance and forest type
```{r}
# filter plots to DSTRBCD1 == 30, 31, or 32, this is selecting all plots with any fire damage [30], (from crown and ground fire, either prescribed or natural).
# [31] = ground fire, [32] = canopy fire
# and then filter CA burned plots to focal forest types including mixed conifer, douglas-fir, ponderosa pine, incense cedar, sugar pine, jefferyi, pine groups, and white fir, and CA black oak groups

CA_COND_filtered <- CA_COND %>%
  filter(if_any(
    c(DSTRBCD1, DSTRBCD2, DSTRBCD3),
    ~ .x %in% c(30, 31, 32)
  )) %>%
  filter(FORTYPCD %in% c(371, 201, 221, 222, 224, 225, 261, 922))

# how many unique PLT_CN values have multiple rows for CONDID

CA_COND_filtered %>%
  dplyr::group_by(PLT_CN) %>%
  dplyr::summarise(n_condid = n_distinct(CONDID), .groups = "drop") %>%
  dplyr::filter(n_condid > 1) %>%
  nrow()

# 39 / 506 have multiple rows for CONDID per PLT_CN

# CA_COND has multiple rows per PLT_CN due to variation in CONDID
#to collapse CA_COND table to have 1 row per PLT_CN first check that CONDID is the only thing that differs across rows with the same PLT_CN, we just want to make sure DSTRBCD dont differ across PLT_CN in COND table

CA_COND_filtered %>%
  dplyr::group_by(PLT_CN) %>%
  dplyr::summarise(
    problem = dplyr::n_distinct(
      paste(DSTRBCD1, DSTRBCD2, DSTRBCD3,
            DSTRBYR1, DSTRBYR2, DSTRBYR3)
    ) > 1,
    .groups = "drop"
  ) %>%
  dplyr::summarise(any_problem = any(problem))


CA_COND_filtered %>%
  group_by(PLT_CN) %>%
  summarise(
    n_profiles = n_distinct(
      paste(DSTRBCD1, DSTRBCD2, DSTRBCD3,
            DSTRBYR1, DSTRBYR2, DSTRBYR3)
    ),
    .groups = "drop"
  ) %>%
  filter(n_profiles > 1) %>%
  nrow()

# 10 PLT_CNs have differing DSTRBCD between rows of the same PLT_CN in COND table

CA_duplicate_CONID<- CA_COND_filtered %>%
  group_by(PLT_CN) %>%
  filter(
    n_distinct(
      paste(DSTRBCD1, DSTRBCD2, DSTRBCD3,
            DSTRBYR1, DSTRBYR2, DSTRBYR3)
    ) > 1
  ) %>%
  arrange(PLT_CN) %>%
  ungroup()

#variation in COND rows per plot may be ecologically meaningful
# when multiple rows for a PLT_CN are present, disturbance only differs between 31-32

# lets simplify by taking the dominant condition using 
# CONDPROP_UNADJ. the dominant condition = the largest CONDPROP_UNADJ for a given PLT_CN. the proportion of the plot occupied by that condition (ranges from # 0–1, sums to ~1 across conditions in a plot)

CA_COND_dominant <- CA_COND_filtered %>%
  group_by(PLT_CN) %>%
  slice_max(CONDPROP_UNADJ, n = 1, with_ties = FALSE) %>%
  ungroup()

CA_COND_summary <- CA_COND_dominant %>%
  dplyr::select(
    PLT_CN,
    CONDPROP_UNADJ,
    FORTYPCD,
    DSTRBCD1, DSTRBYR1,
    DSTRBCD2, DSTRBYR2,
    DSTRBCD3, DSTRBYR3
  )

# attach condition summary to PLOT table (with resurveys) to get the condition history for that plot. the same history will be attached to each resurvey for the plot
CA_burned_PLT_CN <- CA_COND_summary %>%
  distinct(PLT_CN)

CA_PLOT_filtered <- CA_PLOT %>%
  semi_join(CA_burned_PLT_CN, by = c("CN" = "PLT_CN"))

CA_PLOT_by_COND <- CA_PLOT_filtered %>%
  left_join(CA_COND_summary, by = c("CN" = "PLT_CN"))
```

#pre and post-fire surveys
```{r}
# number of revisits

# we can link CN to PREV_PLT_CN to determine resurveys 
# however, PREV_PLT_CN is not recorded for 'periodic' inventories
CA_PLOT_pairs <- CA_PLOT_by_COND %>%
  filter(!is.na(PREV_PLT_CN)) %>%
  left_join(
    CA_PLOT_by_COND %>% dplyr::select(CN, INVYR),
    by = c("PREV_PLT_CN" = "CN"),
    suffix = c("_current", "_previous")
  )

# reshape fire year data to long format 
# each row becomes one fire event so we can determine pre and post fire surveys
CA_fires <- CA_PLOT_by_COND %>%
  tidyr::pivot_longer(
    cols = c(DSTRBYR1, DSTRBYR2, DSTRBYR3),
    names_to = "fire_slot",
    values_to = "fire_year"
  ) %>%
  filter(!is.na(fire_year))

# attach fire events to pre and post visit pairs
CA_fire_events <- CA_PLOT_pairs %>%
  left_join(CA_fires, by = "CN")

# filter for plots that have inventories before and after fire years 
# asking: is there a fire event between previous and current inventory
CA_fire_events <- CA_fire_events %>%
  mutate(
    pre_fire  = INVYR_previous < fire_year,
    post_fire = INVYR_current > fire_year
  ) %>%
  filter(pre_fire & post_fire)

## 22 plots with pre and post fire surveys


# the above code doesnt look at pre and post plots for ANY and ALL fires
# trying this instead 
CA_PLOT_pairs <- CA_PLOT_by_COND %>%
  filter(!is.na(PREV_PLT_CN)) %>%
  left_join(
    CA_PLOT_by_COND %>% dplyr::select(CN, INVYR),
    by = c("PREV_PLT_CN" = "CN"),
    suffix = c("_current", "_previous")
  )

CA_fires <- CA_PLOT_by_COND %>%
  tidyr::pivot_longer(
    cols = c(DSTRBYR1, DSTRBYR2, DSTRBYR3),
    values_to = "fire_year"
  ) %>%
  filter(!is.na(fire_year)) %>%
  distinct(CN, fire_year)

CA_fire_events <- CA_PLOT_pairs %>%
  left_join(CA_fires, by = "CN")

CA_fire_events <- CA_fire_events %>%
  mutate(
    in_interval = fire_year > INVYR_previous &
                  fire_year <= INVYR_current
  )

CA_fire_events %>%
  group_by(PREV_PLT_CN, CN) %>%
  summarise(any_fire = any(in_interval), .groups = "drop") %>%
  count(any_fire)

CA_fire_events %>%
  group_by(PREV_PLT_CN, CN) %>%
  summarise(
    any_fire = any(in_interval, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  count(any_fire)

# 27 plots that had a pre and post fire survey around ANY fire
# potentially more flexible looks for any combination of pre and post fire surveys around all fire years 

# if plot resurveys are being left out due to the 'periodic' versus 'annual' inventory distinction, lets check for resurveys a different way

# create a unique plot ID 
CA_PLOT_IDs <- CA_PLOT_by_COND %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )
CA_PLOT_IDs %>%
  dplyr::count(plot_id, name = "n_resurveys") %>%
  dplyr::mutate(group = ifelse(n_resurveys == 1, "n = 1", "n > 1")) %>%
  dplyr::count(group)

# n= 1 (401) ; n =2 (33)

# trying it 3 different ways there are between 22-33 plots with pre and post fire surveys

```
# CA
## fire only plots
```{r}
# create a unique PLOT ID in COND and in PLOT tables 
CA_COND_IDs <- CA_COND %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )
CA_PLOT_IDs <- CA_PLOT %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

# collapse information from multiple CONDIDs associated with each plot to avoid many-to-many mapping
# and then filter by fire and forest type  

#plot_id → multiple CONDIDs → each CONDID has up to 3 disturbance codes and years
# one row for a specific plot_id, CONID, and disturbance event (combining DSTRBCD and DSTRBYR)
CA_COND_long <- CA_COND_IDs %>%
  dplyr::select(plot_id, CONDID, FORTYPCD,
    DSTRBCD1, DSTRBCD2, DSTRBCD3,
    DSTRBYR1, DSTRBYR2, DSTRBYR3) %>%
  pivot_longer(
    cols = starts_with("DSTR"),
    names_to = c(".value", "event"),
    names_pattern = "(DSTRBCD|DSTRBYR)([1-3])"
  ) %>%
  filter(DSTRBCD %in% c(30, 31, 32)) %>%
  filter(FORTYPCD %in% c(371, 201, 221, 222, 224, 225, 261, 922)) %>%
  mutate(
    fire_year_severity = paste(DSTRBYR, DSTRBCD, sep = "_")
  )

# summarize fire history across rows for unique plot_id (one row per plot_id)
CA_fire_summary <- CA_COND_long %>%
  group_by(plot_id) %>%
  summarise(
    n_fire_events = n_distinct(DSTRBYR), # number of distinct fire years per plot
    
    # this just collapses all codes into a list but not necessarily relative to year
    fire_types = paste(sort(unique(DSTRBCD)), collapse = ", "), 
    fire_years = paste(sort(unique(DSTRBYR)), collapse = ", "),
    
    fire_year_severity =
    paste(unique(fire_year_severity), collapse = ", "),
    
    multiple_fires = n_distinct(DSTRBYR) > 1, # did fire occur more than once?
    multiple_severity = n_distinct(DSTRBCD) > 1, # are there multiple fire types?
    first_fire_year = min(DSTRBYR, na.rm = TRUE),
    last_fire_year = max(DSTRBYR, na.rm = TRUE),
    .groups = "drop"
  )

# how many plots burned multiple times?
CA_fire_summary %>% summarise(n_plots_multiple_fires = sum(multiple_fires))
# 34 plots

# how many plots that burned multipe times have different severities?
CA_fire_summary %>% filter(multiple_fires) %>% summarise(n_plots_multiple_fire_and_severity = sum(multiple_severity))
# 10 plots

# how many plots that burned once have multiple severities within a plot?
CA_fire_summary %>%
  filter(!multiple_fires) %>%
  summarise(
    n_plots_single_fire_multiple_severity =
      sum(multiple_severity))
# 9 plots

CA_burned_PLOT_IDs <- CA_fire_summary %>% 
  distinct(plot_id)

# merge COND and PLOT by this unique ID 
CA_PLOT_by_COND <- CA_PLOT_IDs %>%
  semi_join(CA_burned_PLOT_IDs, by = "plot_id") %>%  #keep only plot_ids with at least one fire 
  left_join(CA_fire_summary, by = "plot_id")  #add fire summary to each plot, should be one-to-many mapping

#for each unique plot, count how many unique inventory years (INVYR) it has
CA_PLOT_by_COND %>%
  group_by(plot_id) %>%
  summarise(
    n_revisits = n_distinct(INVYR),
    .groups = "drop"
  ) %>%
  count(n_revisits)

# n=2 (288)
# n=3 (146)

# how many plots have pre and post fire surveys
CA_fire_events <- CA_COND_long %>%
  dplyr::select(plot_id, DSTRBYR) %>%
  distinct() %>%
  rename(fire_year = DSTRBYR)

# does each unique plot_id have a pre and post fire survey around any/all fire years?
CA_prepost_fire_summary <- CA_PLOT_by_COND %>%
  distinct(plot_id, INVYR) %>%
  inner_join(
    CA_fire_events %>% distinct(plot_id, fire_year),
    by = "plot_id"
  ) %>%
  mutate(
    relationship = case_when(
      INVYR < fire_year ~ "pre",
      INVYR > fire_year ~ "post",
      TRUE ~ "same_year"
    )
  ) %>%
  group_by(plot_id, fire_year) %>%
  summarise(
    has_pre  = any(relationship == "pre"),
    has_post = any(relationship == "post"),
    n_pre_surveys  = sum(relationship == "pre"),
    n_post_surveys = sum(relationship == "post"),
    .groups = "drop"
  ) %>%
  mutate(
      status = case_when(
      has_pre & has_post ~ "pre_and_post",
      has_pre ~ "pre_only",
      has_post ~ "post_only",
      TRUE ~ "no_clear_separation"
    )
  )

CA_prepost_fire_summary %>%
  count(status)

# post only = 106
# pre only = 49
#pre and post = 313

CA_prepost_fire_summary %>%
  distinct(plot_id, status) %>%
  count(status)

# post only = 106
# pre only = 49
# pre and post = 304


# how many plots are there in CA
CA_PLOT %>%
  distinct(LAT, LON) %>%
  nrow()
# 22362

length(unique(CA_COND_IDs$plot_id)) # 22589
length(unique(CA_PLOT_IDs$plot_id)) # 22589

```

## tree
```{r}
#Tree number. A number used to uniquely identify a tree on a subplot. Tree numbers can be used to track trees when PLOT.DESIGNCD or DESIGNCD is the same between inventories

# before looking into tree level data, just confirm that the design codes are the same across plots in our CA_PLOT_by_COND table that has our burned plots
n_distinct(CA_PLOT_by_COND$DESIGNCD) # one design code across the filtered plots
unique(CA_PLOT_by_COND$DESIGNCD) # that design code is 501

# all plots have the same design code so you can proceed with linking trees across inventories

# do we have surviving trees in post-fire surveys?
# create unique plot ids for CA_TREE
CA_TREE_IDs <- CA_TREE %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

#filter CA_TREE to only keep the unique plot ids from CA_PLOT_by_COND (our fire only plots in our focal forest types)
CA_TREE_filtered <- CA_TREE_IDs %>%
  filter(plot_id %in% CA_PLOT_by_COND$plot_id)

# to locate a specific tree ID, use the following variables 
# STATECD, INVYR, UNITCD, COUNTYCD, PLOT, SUBP, TREE
# TREE is unique to SUBP not plot_id
# we can look for unique combinations of SUBP and TREE within a plot_id (STATECD, UNITCD, COUNTYCD, PLOT)

# to count how many times a tree was surveyed
# group by plot_id, SUBP, TREE, and then count the number of INVYR
CA_TREE_resurveys <- CA_TREE_filtered %>% 
  group_by(plot_id, SUBP, TREE) %>% 
  summarise(n_surveys = n_distinct(INVYR),
            survey_years = paste(sort(unique(INVYR)), collapse = ", "),
            .groups = "drop")

# how many times have trees been resurveyed
CA_TREE_resurveys %>%
  count(n_surveys)

#n=1 (1777)
#n=2 (12611)
#n=3 (4818)

# is TREE our unique identifier of TREE ?
CA_TREE_filtered %>%
  group_by(TREE) %>%
  summarise(
    n_plots = n_distinct(plot_id),
    .groups = "drop"
  ) %>%
  filter(n_plots > 1) %>%
  summarise(
    n_trees_multiple_plots = n()
  )
# 299 trees are found in multiple plots 

CA_TREE_filtered %>%
  summarise(
    n_unique_trees = n_distinct(plot_id, SUBP, TREE)
  )
# number of unique trees 19206

# Tree number notes:
# TREE uniquely identifies a tree within a subplot (SUBP),
# so full tree identity is: plot_id + SUBP + TREE

# unique tree identity is plot_id + SUBP + TREE
# make a unique tree identifier too? or use plot_id + SUBP + TREE everytime
#tree_id = paste(plot_id, SUBP, TREE, sep = "_") 

# lets look at trees in the plots with both pre and post fire surveys
# pull the plot_ids for status = pre_and_post from CA_prepost_fire_summary and the fire year the post-fire plot is associated with

CA_TREE_prepost_fire <- CA_TREE_filtered %>%
  semi_join(
    CA_prepost_fire_summary %>% filter(status == "pre_and_post"),
    by = "plot_id"
  ) %>%
  left_join(
    CA_prepost_fire_summary %>%
      filter(status == "pre_and_post") %>%
      dplyr::select(plot_id, fire_year, status),
    by = "plot_id"
  )

CA_prepost_fire_summary %>%
  filter(status == "pre_and_post") %>%
  summarise(n_plots = n_distinct(plot_id)) #304
length(unique(CA_TREE_prepost_fire$plot_id)) #304


# both CA_fire_summary and CA_TREE_prepost_fire have 304 unique plot_id with pre and post fire inventories

# now we want to look at just post-fire plots (where INVYR > fire_year)
CA_TREE_post_fire_plots <- CA_TREE_prepost_fire %>%
  filter(INVYR > fire_year)

# number of trees per plot (post-fire plots)
CA_TREE_per_plot_post <- CA_TREE_post_fire_plots %>%
  mutate(which = "post") %>%
  distinct(plot_id, INVYR, fire_year, status, SUBP, TREE, which) %>%
  group_by(plot_id, INVYR, fire_year, status, which) %>%
  summarise(n_tree = n(), .groups = "drop")
# 9/304 post-fire plots have less than 10 trees, everything else has 10+ trees

# get the abundance of live and dead trees per species in each plot 
CA_TREE_post_fire_plots_tree_status <- CA_TREE_post_fire_plots %>%
  group_by(plot_id, INVYR, fire_year, SPCD, STATUSCD) %>%
  summarise(
    n_tree = n_distinct(SUBP, TREE),
    .groups = "drop"
  ) %>%
  left_join(
    REF_SPECIES %>%
      dplyr::select(SPCD, GENUS, SPECIES),
    by = "SPCD"
  )
# STATUSCD = 0 (not present), STATUSCD = 1 (live), STATUSCD = 2 (dead)

# number of species with STATUSCD = 1 in post-fire plots
CA_TREE_post_fire_plot_summary <- CA_TREE_post_fire_plots %>%
  filter(STATUSCD == 1) %>%
  group_by(plot_id, INVYR, fire_year) %>%
  summarise(
    n_live_species = n_distinct(SPCD),
    n_live_trees = n_distinct(SUBP, TREE),
    .groups = "drop"
  )

# number of individuals (live) per species per plot
CA_TREE_post_fire_species_summary <- CA_TREE_post_fire_plots %>%
  filter(STATUSCD == 1) %>%
  group_by(plot_id, INVYR, fire_year, SPCD) %>%
  summarise(
    n_live_trees = n_distinct(SUBP, TREE),
    .groups = "drop"
  ) %>%
  left_join(
    REF_SPECIES %>%
      dplyr::select(SPCD, GENUS, SPECIES),
    by = "SPCD"
  ) %>%
  mutate(species = paste(GENUS, SPECIES, sep = "_"))

### visualize plot level summaries

# number of live species per plot (post-fire plots)
CA_TREE_post_fire_species_summary %>%
  group_by(plot_id) %>%
  summarise(n_species = n_distinct(SPCD)) %>%
  ggplot(aes(x = n_species)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  theme_bw() +
  labs(x = "Number of live species per plot", y = "Number of plots")

# abundance of each species per plot (top 20 most abundant species)
ggplot(
  CA_TREE_post_fire_species_summary %>%
    group_by(species, plot_id) %>%
    summarise(n_live_trees = sum(n_live_trees), .groups = "drop") %>%
    group_by(species) %>%
    summarise(mean_per_plot = mean(n_live_trees), .groups = "drop") %>%
    slice_max(mean_per_plot, n = 20),
  aes(x = reorder(species, mean_per_plot),
      y = mean_per_plot)
) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    x = "species",
    y = "average # live trees per plot",
    title = "average abundance of species per plot"
  ) +
  theme_bw()

# boxplots of tree abundance per species per plot
CA_TREE_post_fire_plots %>%
  filter(STATUSCD == 1) %>%
  group_by(plot_id, SPCD) %>%
  summarise(
    n_trees = n_distinct(SUBP, TREE),
    .groups = "drop"
  ) %>%
  left_join(
    REF_SPECIES %>%
      dplyr::select(SPCD, GENUS, SPECIES),
    by = "SPCD"
  ) %>%
  mutate(species = paste(GENUS, SPECIES, sep = "_")) %>%
  ggplot(aes(x = species, y = n_trees)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6) +
  coord_flip() +
  labs(
    x = "species",
    y = "trees per plot",
    title = "tree abundance per species per plot"
  ) +
  scale_y_continuous(
    breaks = function(x) seq(0, max(x, na.rm = TRUE), by = 5)
  ) +
  theme_bw()


## tree transitions and mortality rates

# closest PRE-fire observation for each TREE
CA_TREE_pre_fire_trees <- CA_TREE_prepost_fire %>%
  filter(INVYR < fire_year) %>%
  group_by(plot_id, SUBP, TREE, fire_year) %>%
  arrange(desc(INVYR), .by_group = TRUE) %>%
  slice(1) %>%
  ungroup() %>%
  rename(
    pre_INVYR = INVYR,
    pre_STATUSCD = STATUSCD
  )

# closest POST-fire observation for each TREE
CA_TREE_post_fire_trees <- CA_TREE_prepost_fire %>%
  filter(INVYR > fire_year) %>%
  group_by(plot_id, SUBP, TREE, fire_year) %>%
  arrange(INVYR, .by_group = TRUE) %>%
  slice(1) %>%
  ungroup() %>%
  rename(
    post_INVYR = INVYR,
    post_STATUSCD = STATUSCD
  )

# join pre and post observations
CA_TREE_transitions <- CA_TREE_pre_fire_trees %>%
  inner_join(
    CA_TREE_post_fire_trees,
    by = c("plot_id", "SUBP", "TREE", "fire_year"),
    suffix = c("_pre", "_post")
  )

# mortality indicators
CA_TREE_transitions <- CA_TREE_transitions %>%
  mutate(
    alive_pre = pre_STATUSCD == 1,
    dead_post = pre_STATUSCD == 1 & post_STATUSCD == 2
  )

# plot-level mortality
CA_TREE_mortality <- CA_TREE_transitions %>%
  group_by(plot_id, fire_year) %>%
  summarise(
    n_alive_pre = sum(alive_pre, na.rm = TRUE),
    n_dead_post = sum(dead_post, na.rm = TRUE),
    mortality_rate = n_dead_post / n_alive_pre,
    .groups = "drop"
  )

summary(CA_TREE_mortality$mortality_rate)
hist(CA_TREE_mortality$mortality_rate)
```

## seedling
```{r}
# is there recruitment in post-fire surveys?

# seedling counts are under TREECOUNT in CA_SEEDLING
# TREECOUNT is for  (plot × subplot/microplot × species × inventory)
# in the CA_SEEDLING table, SUBP = microplot since seedlings are inventoried at the microplot within a subplot

# create unique plot ids for CA_SEEDLING
CA_SEEDLING_IDs <- CA_SEEDLING %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

#filter CA_SEEDLING to only keep the unique plot ids from CA_PLOT_by_COND
CA_SEEDLING_filtered <- CA_SEEDLING_IDs %>%
  filter(plot_id %in% CA_PLOT_by_COND$plot_id)

# lets look at trees in the plots with both pre and post fire 
# pull the plot_ids for status = pre_and_post from CA_prepost_fire_summary

CA_SEEDLING_prepost_fire <- CA_SEEDLING_filtered %>%
  semi_join(CA_prepost_fire_summary %>% filter(status == "pre_and_post"),
    by = "plot_id") %>%
  left_join(CA_prepost_fire_summary %>%
      filter(status == "pre_and_post") %>%
      dplyr::select(plot_id, fire_year, status),
    by = "plot_id")

# filter to post-fire plots only
CA_SEEDLING_post_fire_plots <- CA_SEEDLING_prepost_fire %>%
  filter(INVYR > fire_year)


length(unique(CA_SEEDLING_post_fire_plots$plot_id)) # 237
length(unique(CA_TREE_post_fire_plots$plot_id)) # 304
# 237 unique plots (why is this different from the 304 for the TREE analysis ??)

## To qualify for counting, conifer seedlings must be at least 6 inches tall and hardwood seedlings must be at least 12 inches tall (which is maybe why that column is blank on the table).

# how many post fire plots have TREECOUNT = NA (no seedlings recorded)
CA_SEEDLING_post_fire_plots %>%
  filter(is.na(TREECOUNT)) %>%
  summarise(n_plot_ids = n_distinct(plot_id))
# 0 / 237 have no seedlings 

# number of seedlings per plot (aggregated across SUBP/microplot)
CA_SEEDLING_per_plot <- CA_SEEDLING_post_fire_plots %>%
  group_by(plot_id, INVYR, fire_year) %>%
  summarise(
    n_seedlings = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop"
  )

hist(CA_SEEDLING_per_plot$n_seedlings, breaks=30)
summary(CA_SEEDLING_per_plot$n_seedlings)

# number of seedlings per species per plot (aggregated across SUBP/microplot)
CA_SEEDLING_species_summary <- CA_SEEDLING_post_fire_plots %>%
  group_by(plot_id, INVYR, fire_year, SPCD) %>%
  summarise(n_seedlings = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop") %>%
  left_join(REF_SPECIES %>% dplyr::select(SPCD, GENUS, SPECIES),
    by = "SPCD") %>%
  mutate(species = paste(GENUS, SPECIES, sep = "_"))

CA_SEEDLING_species_summary %>%
  group_by(plot_id) %>%
  summarise(n_species = n_distinct(SPCD)) %>%
  ggplot(aes(x = n_species)) +
  geom_histogram(binwidth = 1, fill = "darkgreen", color = "white") +
  theme_bw() +
  labs(
    x = "Number of seedling species per plot",
    y = "Number of plots"
  )

# number of seedling species per plot 
CA_SEEDLING_plot_summary <- CA_SEEDLING_post_fire_plots %>%
  group_by(plot_id, INVYR, fire_year) %>%
  summarise(
    n_seedling_species = n_distinct(SPCD),
    n_seedlings = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop"
  )

# average number of seedlings per species per plot
ggplot(
  CA_SEEDLING_species_summary %>%
    group_by(species, plot_id) %>%
    summarise(n_seedlings = sum(n_seedlings), .groups = "drop") %>%
    group_by(species) %>%
    summarise(mean_per_plot = mean(n_seedlings), .groups = "drop") %>%
    slice_max(mean_per_plot, n = 20),
  aes(
    x = reorder(species, mean_per_plot),
    y = mean_per_plot
  )
) +
  geom_col(fill = "darkgreen") +
  coord_flip() +
  scale_y_continuous(
    breaks = function(x) seq(0, max(x, na.rm = TRUE), by = 5)
  ) +
  labs(
    x = "species",
    y = "average # seedlings per plot",
    title = "average seedling abundance per species per plot"
  ) +
  theme_bw()

# how many plots have 0 species?
CA_SEEDLING_post_fire_plots %>%
  group_by(plot_id) %>%
  summarise(n_species = n_distinct(SPCD[TREECOUNT > 0])) %>%
  filter(n_species == 0)
# no plots have zero species as seedlings

# boxplots of seedling abundances per species per plot
CA_SEEDLING_species_summary %>%
  group_by(plot_id, species) %>%
  summarise(n_seedlings = sum(n_seedlings), .groups = "drop") %>%
  
  mutate(n_seedlings_capped = pmin(n_seedlings, 50)) %>%
  
  ggplot(aes(x = species, y = n_seedlings_capped)) +
  geom_boxplot(fill = "darkgreen", alpha = 0.6) +
  coord_flip() +
  labs(
    x = "species",
    y = "seedlings per plot (capped at 50)",
    title = "seedling abundance per species per plot"
  ) +
  theme_bw()

```
## seedling expanded
```{r}
# expand seedling analysis to look at trees in plots with both pre and post fire and just post-fire
# pull the plot_ids for status = pre_and_post, and post_only from CA_prepost_fire_summary

CA_SEEDLING_prepost_fire <- CA_SEEDLING_filtered %>%
  semi_join(CA_prepost_fire_summary %>% filter(status %in% c("pre_and_post", "post_only")),
    by = "plot_id") %>%
  left_join(CA_prepost_fire_summary %>%
      filter(status %in% c("pre_and_post", "post_only")) %>%
      dplyr::select(plot_id, fire_year, status),
    by = "plot_id")

# filter to post-fire plots only
CA_SEEDLING_post_fire_plots <- CA_SEEDLING_prepost_fire %>%
  filter(INVYR > fire_year)

length(unique(CA_SEEDLING_post_fire_plots$plot_id)) # 313
# 313 unique plots instead of the 234 in the pre_and_post analysis above

# how many post fire plots have TREECOUNT = NA (no seedlings recorded)
CA_SEEDLING_post_fire_plots %>%
  filter(is.na(TREECOUNT)) %>%
  summarise(n_plot_ids = n_distinct(plot_id))
# 16 / 237 have no seedlings 

# number of seedlings per plot (aggregated across SUBP/microplot)
CA_SEEDLING_per_plot <- CA_SEEDLING_post_fire_plots %>%
  group_by(plot_id, INVYR, fire_year) %>%
  summarise(
    n_seedlings = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop"
  )

hist(CA_SEEDLING_per_plot$n_seedlings, breaks=30)
summary(CA_SEEDLING_per_plot$n_seedlings)

# slightly less seedlings per plot when adding ~100 more plots that are post_only 
# median = 6, mean = 17

# median is a more useful metric for seedling counts that mean because seedlings counts are highly skewed and median values give us a better approximation of the most typical plot

# number of species of seedlings per plot 
CA_SEEDLING_plot_summary <- CA_SEEDLING_post_fire_plots %>%
  group_by(plot_id, INVYR, fire_year) %>%
  summarise(
    n_seedling_species = n_distinct(SPCD),
    n_seedlings = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop"
  )

# number of individuals per species per plot
CA_SEEDLING_species_summary <- CA_SEEDLING_post_fire_plots %>%
  group_by(plot_id, INVYR, fire_year, SPCD) %>%
  summarise(n_seedlings = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop") %>%
  left_join(REF_SPECIES %>% dplyr::select(SPCD, GENUS, SPECIES),
    by = "SPCD") %>%
  mutate(species = paste(GENUS, SPECIES, sep = "_"))
```

#post-fire mortality
```{r}

```

# time since fire
```{r}
# since replication is low we might just need to use a bunch of plots stratified across time since fire instead of pre and post fire surveys

CA_PLOT_time_since_fire <- CA_PLOT_by_COND %>%
  distinct(CN, INVYR, PREV_PLT_CN, .keep_all = TRUE) %>%
  tidyr::pivot_longer(
    cols = c(DSTRBYR1, DSTRBYR2, DSTRBYR3),
    values_to = "fire_year"
  ) %>%
  mutate(
    fire_year = ifelse(fire_year == 9999, NA, fire_year)
  ) %>%
  filter(!is.na(fire_year)) %>%
  mutate(
    time_since_fire = INVYR - fire_year,
    pre_fire  = time_since_fire < 0,
    post_fire = time_since_fire > 0
  )

hist(CA_PLOT_time_since_fire$time_since_fire)
```

# expand states
# OR
### fire only plots
```{r}
#create a unique PLOT ID in COND and in PLOT tables 
OR_COND_IDs <- OR_COND %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )
OR_PLOT_IDs <- OR_PLOT %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

# collapse information from multiple CONDIDs associated with each plot to avoid many-to-many mapping
# and then filter by fire and forest type  

#plot_id → multiple CONDIDs → each CONDID has up to 3 disturbance codes and years
# one row for a specific plot_id, CONID, and disturbance event (combining DSTRBCD and DSTRBYR)
OR_COND_long <- OR_COND_IDs %>%
  dplyr::select(plot_id, CONDID, FORTYPCD,
    DSTRBCD1, DSTRBCD2, DSTRBCD3,
    DSTRBYR1, DSTRBYR2, DSTRBYR3) %>%
  pivot_longer(
    cols = starts_with("DSTR"),
    names_to = c(".value", "event"),
    names_pattern = "(DSTRBCD|DSTRBYR)([1-3])"
  ) %>%
  filter(DSTRBCD %in% c(30, 31, 32)) %>%
  filter(FORTYPCD %in% c(371, 201, 221, 222, 224, 225, 261, 922)) %>%
  mutate(
    fire_year_severity = paste(DSTRBYR, DSTRBCD, sep = "_")
  )

# summarize fire history across rows for unique plot_id (one row per plot_id)
OR_fire_summary <- OR_COND_long %>%
  group_by(plot_id) %>%
  summarise(
    n_fire_events = n_distinct(DSTRBYR), # number of distinct fire years per plot
    
    # this just collapses all codes into a list but not necessarily relative to year
    fire_types = paste(sort(unique(DSTRBCD)), collapse = ", "), 
    fire_years = paste(sort(unique(DSTRBYR)), collapse = ", "),
    
    fire_year_severity =
    paste(unique(fire_year_severity), collapse = ", "),
    
    multiple_fires = n_distinct(DSTRBYR) > 1, # did fire occur more than once?
    multiple_severity = n_distinct(DSTRBCD) > 1, # are there multiple fire types?
    first_fire_year = min(DSTRBYR, na.rm = TRUE),
    last_fire_year = max(DSTRBYR, na.rm = TRUE),
    .groups = "drop"
  )

# how many plots burned multiple times?
OR_fire_summary %>% summarise(n_plots_multiple_fires = sum(multiple_fires))
# 20 plots

# how many plots that burned multipe times have different severities?
OR_fire_summary %>% filter(multiple_fires) %>% summarise(n_plots_multiple_fire_and_severity = sum(multiple_severity))
# 11 plots

# how many plots that burned once have multiple severities within a plot?
OR_fire_summary %>%
  filter(!multiple_fires) %>%
  summarise(
    n_plots_single_fire_multiple_severity =
      sum(multiple_severity))
# 9 plots

OR_burned_PLOT_IDs <- OR_fire_summary %>% 
  distinct(plot_id)

# merge COND and PLOT by this unique ID 
OR_PLOT_by_COND <- OR_PLOT_IDs %>%
  semi_join(OR_burned_PLOT_IDs, by = "plot_id") %>%  #keep only plot_ids with at least one fire 
  left_join(OR_fire_summary, by = "plot_id")  #add fire summary to each plot, should be one-to-many mapping

#for each unique plot, count how many unique inventory years (INVYR) it has
OR_PLOT_by_COND %>%
  group_by(plot_id) %>%
  summarise(
    n_revisits = n_distinct(INVYR),
    .groups = "drop"
  ) %>%
  count(n_revisits)

# n=1 (23)
# n=2 (485)
# n=3 (87)

# how many plots have pre and post fire surveys
OR_fire_events <- OR_COND_long %>%
  dplyr::select(plot_id, DSTRBYR) %>%
  distinct() %>%
  rename(fire_year = DSTRBYR)

# does each unique plot_id have a pre and post fire survey around any/all fire years?
OR_prepost_fire_summary <- OR_PLOT_by_COND %>%
  distinct(plot_id, INVYR) %>%
  inner_join(
    OR_fire_events %>% distinct(plot_id, fire_year),
    by = "plot_id"
  ) %>%
  mutate(
    relationship = case_when(
      INVYR < fire_year ~ "pre",
      INVYR > fire_year ~ "post",
      TRUE ~ "same_year"
    )
  ) %>%
  group_by(plot_id, fire_year) %>%
  summarise(
    has_pre  = any(relationship == "pre"),
    has_post = any(relationship == "post"),
    n_pre_surveys  = sum(relationship == "pre"),
    n_post_surveys = sum(relationship == "post"),
    .groups = "drop"
  ) %>%
  mutate(
      status = case_when(
      has_pre & has_post ~ "pre_and_post",
      has_pre ~ "pre_only",
      has_post ~ "post_only",
      TRUE ~ "no_clear_separation"
    )
  )

OR_prepost_fire_summary %>%
  count(status)
# no clear separation = 5 
# post only = 232
# pre only = 44
#pre and post = 334


# how many plots are there in CA
OR_PLOT %>%
  distinct(LAT, LON) %>%
  nrow()
# 22411

length(unique(OR_COND_IDs$plot_id)) # 22418
length(unique(OR_PLOT_IDs$plot_id)) # 22418


```
## tree
```{r}
#Tree number. A number used to uniquely identify a tree on a subplot. Tree numbers can be used to track trees when PLOT.DESIGNCD or DESIGNCD is the same between inventories

# before looking into tree level data, just confirm that the design codes are the same across plots in our CA_PLOT_by_COND table that has our burned plots
n_distinct(OR_PLOT_by_COND$DESIGNCD) # two design codes across the filtered plots
unique(OR_PLOT_by_COND$DESIGNCD) # that design code is 501, 502

# two different design codes, may be important for relating tree surveys

# do we have surviving trees in post-fire surveys?
# create unique plot ids for CA_TREE
OR_TREE_IDs <- OR_TREE %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

#filter CA_TREE to only keep the unique plot ids from CA_PLOT_by_COND (our fire only plots in our focal forest types)
OR_TREE_filtered <- OR_TREE_IDs %>%
  filter(plot_id %in% OR_PLOT_by_COND$plot_id)

# to locate a specific tree ID, use the following variables 
# STATECD, INVYR, UNITCD, COUNTYCD, PLOT, SUBP, TREE
# TREE is unique to SUBP not plot_id
# we can look for unique combinations of SUBP and TREE within a plot_id (STATECD, UNITCD, COUNTYCD, PLOT)

# to count how many times a tree was surveyed
# group by plot_id, SUBP, TREE, and then count the number of INVYR
OR_TREE_resurveys <- OR_TREE_filtered %>% 
  group_by(plot_id, SUBP, TREE) %>% 
  summarise(n_surveys = n_distinct(INVYR),
            survey_years = paste(sort(unique(INVYR)), collapse = ", "),
            .groups = "drop")

# how many times have trees been resurveyed
OR_TREE_resurveys %>%
  count(n_surveys)

#n=1 (3161)
#n=2 (14462)
#n=3 (2847)

# maybe make a unique tree identifier too?
#tree_id = paste(plot_id, SUBP, TREE, sep = "_") # also including INVYR?

# lets look at trees in the plots with both pre and post fire surveys
# pull the plot_ids for status = pre_and_post from CA_prepost_fire_summary and the fire year the post-fire plot is associated with

OR_TREE_prepost_fire <- OR_TREE_filtered %>%
  semi_join(
    OR_prepost_fire_summary %>% filter(status == "pre_and_post"),
    by = "plot_id"
  ) %>%
  left_join(
    OR_prepost_fire_summary %>%
      filter(status == "pre_and_post") %>%
      dplyr::select(plot_id, fire_year, status),
    by = "plot_id"
  )

OR_prepost_fire_summary %>%
  filter(status == "pre_and_post") %>%
  summarise(n_plots = n_distinct(plot_id)) #330
length(unique(OR_TREE_prepost_fire$plot_id)) #329

# both CA_fire_summary and CA_TREE_prepost_fire have 304 unique plot_id with pre and post fire inventories

# now we want to look at just post-fire plots (where INVYR > fire_year)
OR_TREE_post_fire_plots <- OR_TREE_prepost_fire %>%
  filter(INVYR > fire_year)

# number of trees per plot (post-fire plots)
OR_TREE_per_plot_post <- OR_TREE_post_fire_plots %>%
  mutate(which = "post") %>%
  distinct(plot_id, INVYR, fire_year, status, SUBP, TREE, which) %>%
  group_by(plot_id, INVYR, fire_year, status, which) %>%
  summarise(n_tree = n(), .groups = "drop")

summary(OR_TREE_per_plot_post$n_tree)

# get the abundance of live and dead trees per species in each plot 
OR_TREE_post_fire_plots_tree_status <- OR_TREE_post_fire_plots %>%
  group_by(plot_id, INVYR, fire_year, SPCD, STATUSCD) %>%
  summarise(n_tree = n(), .groups = "drop") %>%
  left_join(REF_SPECIES %>% dplyr::select(SPCD, GENUS, SPECIES), 
            by = "SPCD")

# STATUSCD = 0 (not present), STATUSCD = 1 (live), STATUSCD = 2 (dead)

# number of species with STATUSCD = 1 in post-fire plots
OR_TREE_post_fire_plot_summary <- OR_TREE_post_fire_plots %>%
  filter(STATUSCD == 1) %>%
  group_by(plot_id, INVYR, fire_year) %>%
  summarise(
    n_live_species = n_distinct(SPCD),
    n_live_trees = n(),
    .groups = "drop")

# number of individuals (live) per species per plot
OR_TREE_post_fire_species_summary <- OR_TREE_post_fire_plots %>%
  filter(STATUSCD == 1) %>%
  group_by(plot_id, INVYR, fire_year, SPCD) %>%
  summarise(
    n_live_trees = n(),
    .groups = "drop"
  ) %>%
  left_join(
    REF_SPECIES %>%
      dplyr::select(SPCD, GENUS, SPECIES),
    by = "SPCD"
  ) %>%
  mutate(species = paste(GENUS, SPECIES, sep = "_"))

### visualize plot level summaries

# number of live species per plot (post-fire plots)
OR_TREE_post_fire_species_summary %>%
  group_by(plot_id) %>%
  summarise(n_species = n_distinct(SPCD)) %>%
  ggplot(aes(x = n_species)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  theme_bw() +
  labs(x = "Number of live species per plot", y = "Number of plots")

# abundance of each species per plot (top 20 most abundant species)
ggplot(
  OR_TREE_post_fire_species_summary %>%
    group_by(species, plot_id) %>%
    summarise(n_live_trees = sum(n_live_trees), .groups = "drop") %>%
    group_by(species) %>%
    summarise(mean_per_plot = mean(n_live_trees), .groups = "drop") %>%
    slice_max(mean_per_plot, n = 20),
  aes(x = reorder(species, mean_per_plot),
      y = mean_per_plot)
) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    x = "species",
    y = "average # live trees per plot",
    title = "average abundance of species per plot"
  ) +
  theme_bw()

# boxplots of tree abundance per species per plot
OR_TREE_post_fire_plots %>%
  group_by(plot_id, SPCD) %>%
  summarise(n_trees = n(), .groups = "drop") %>%
  left_join(
    REF_SPECIES %>% dplyr:: select(SPCD, GENUS, SPECIES),
    by = "SPCD"
  ) %>%
  mutate(species = paste(GENUS, SPECIES, sep = "_")) %>%
  ggplot(aes(x = species, y = n_trees)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6) +
  coord_flip() +
  labs(
    x = "species",
    y = "trees per plot",
    title = "tree abundance per species per plot"
  ) +
  scale_y_continuous(
    breaks = function(x) seq(0, max(x, na.rm = TRUE), by = 5)
  ) +
  theme_bw()

```

## seedling
```{r}
# is there recruitment in post-fire surveys?

# seedling counts are under TREECOUNT in CA_SEEDLING
# TREECOUNT is for  (plot × subplot/microplot × species × inventory)
# in the CA_SEEDLING table, SUBP = microplot since seedlings are inventoried at the microplot within a subplot

# create unique plot ids for CA_SEEDLING
OR_SEEDLING_IDs <- OR_SEEDLING %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

#filter CA_SEEDLING to only keep the unique plot ids from CA_PLOT_by_COND
OR_SEEDLING_filtered <- OR_SEEDLING_IDs %>%
  filter(plot_id %in% OR_PLOT_by_COND$plot_id)

# lets look at trees in the plots with both pre and post fire 
# pull the plot_ids for status = pre_and_post from CA_prepost_fire_summary

OR_SEEDLING_prepost_fire <- OR_SEEDLING_filtered %>%
  semi_join(OR_prepost_fire_summary %>% filter(status == "pre_and_post"),
    by = "plot_id") %>%
  left_join(OR_prepost_fire_summary %>%
      filter(status == "pre_and_post") %>%
      dplyr::select(plot_id, fire_year, status),
    by = "plot_id")

# filter to post-fire plots only
OR_SEEDLING_post_fire_plots <- OR_SEEDLING_prepost_fire %>%
  filter(INVYR > fire_year)

length(unique(OR_SEEDLING_post_fire_plots$plot_id)) # 203
length(unique(OR_TREE_post_fire_plots$plot_id)) # 329
# 203 unique plots (why is this different from the 329 for the TREE analysis ??)

# number of seedlings per plot (aggregated across SUBP/microplot)
OR_SEEDLING_per_plot <- OR_SEEDLING_post_fire_plots %>%
  group_by(plot_id, INVYR, fire_year) %>%
  summarise(
    n_seedlings = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop"
  )

hist(OR_SEEDLING_per_plot$n_seedlings, breaks=30)
summary(OR_SEEDLING_per_plot$n_seedlings)

# number of seedlings per species per plot (aggregated across SUBP/microplot)
OR_SEEDLING_species_summary <- OR_SEEDLING_post_fire_plots %>%
  group_by(plot_id, INVYR, fire_year, SPCD) %>%
  summarise(n_seedlings = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop") %>%
  left_join(REF_SPECIES %>% dplyr::select(SPCD, GENUS, SPECIES),
    by = "SPCD") %>%
  mutate(species = paste(GENUS, SPECIES, sep = "_"))

OR_SEEDLING_species_summary %>%
  group_by(plot_id) %>%
  summarise(n_species = n_distinct(SPCD)) %>%
  ggplot(aes(x = n_species)) +
  geom_histogram(binwidth = 1, fill = "darkgreen", color = "white") +
  theme_bw() +
  labs(
    x = "Number of seedling species per plot",
    y = "Number of plots"
  )

# number of species of seedlings per plot 
OR_SEEDLING_plot_summary <- OR_SEEDLING_post_fire_plots %>%
  group_by(plot_id, INVYR, fire_year) %>%
  summarise(
    n_seedling_species = n_distinct(SPCD),
    n_seedlings = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop"
  )

# average number of seedlings per species per plot
ggplot(
  OR_SEEDLING_species_summary %>%
    group_by(species, plot_id) %>%
    summarise(n_seedlings = sum(n_seedlings), .groups = "drop") %>%
    group_by(species) %>%
    summarise(mean_per_plot = mean(n_seedlings), .groups = "drop") %>%
    slice_max(mean_per_plot, n = 20),
  aes(
    x = reorder(species, mean_per_plot),
    y = mean_per_plot
  )
) +
  geom_col(fill = "darkgreen") +
  coord_flip() +
  scale_y_continuous(
    breaks = function(x) seq(0, max(x, na.rm = TRUE), by = 5)
  ) +
  labs(
    x = "species",
    y = "average # seedlings per plot",
    title = "average seedling abundance per species per plot"
  ) +
  theme_bw()

# how many plots have 0 species?
OR_SEEDLING_post_fire_plots %>%
  group_by(plot_id) %>%
  summarise(n_species = n_distinct(SPCD[TREECOUNT > 0])) %>%
  filter(n_species == 0)
# no plots have zero species as seedlings

# boxplots of seedling abundances per species per plot
OR_SEEDLING_species_summary %>%
  group_by(plot_id, species) %>%
  summarise(n_seedlings = sum(n_seedlings), .groups = "drop") %>%
  
  mutate(n_seedlings_capped = pmin(n_seedlings, 50)) %>%
  
  ggplot(aes(x = species, y = n_seedlings_capped)) +
  geom_boxplot(fill = "darkgreen", alpha = 0.6) +
  coord_flip() +
  labs(
    x = "species",
    y = "seedlings per plot (capped at 50)",
    title = "seedling abundance per species per plot"
  ) +
  theme_bw()

```
## seedling expanded
```{r}
# expand seedling analysis to look at trees in plots with both pre and post fire and just post-fire
# pull the plot_ids for status = pre_and_post, and post_only from CA_prepost_fire_summary

OR_SEEDLING_prepost_fire <- OR_SEEDLING_filtered %>%
  semi_join(OR_prepost_fire_summary %>% filter(status %in% c("pre_and_post", "post_only")),
    by = "plot_id") %>%
  left_join(OR_prepost_fire_summary %>%
      filter(status %in% c("pre_and_post", "post_only")) %>%
      dplyr::select(plot_id, fire_year, status),
    by = "plot_id")

# filter to post-fire plots only
OR_SEEDLING_post_fire_plots <- OR_SEEDLING_prepost_fire %>%
  filter(INVYR > fire_year)

length(unique(OR_SEEDLING_post_fire_plots$plot_id)) # 381
# 381 unique plots instead of the 203 in the pre_and_post analysis above

# how many post fire plots have TREECOUNT = NA (no seedlings recorded)
CA_SEEDLING_post_fire_plots %>%
  filter(is.na(TREECOUNT)) %>%
  summarise(n_plot_ids = n_distinct(plot_id))
# 16 / 237 have no seedlings 

# number of seedlings per plot (aggregated across SUBP/microplot)
CA_SEEDLING_per_plot <- CA_SEEDLING_post_fire_plots %>%
  group_by(plot_id, INVYR, fire_year) %>%
  summarise(
    n_seedlings = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop"
  )

hist(CA_SEEDLING_per_plot$n_seedlings, breaks=30)
summary(CA_SEEDLING_per_plot$n_seedlings)

# slightly less seedlings per plot when adding ~100 more plots that are post_only 
# median = 6, mean = 17

# median is a more useful metric for seedling counts that mean because seedlings counts are highly skewed and median values give us a better approximation of the most typical plot

# number of species of seedlings per plot 
CA_SEEDLING_plot_summary <- CA_SEEDLING_post_fire_plots %>%
  group_by(plot_id, INVYR, fire_year) %>%
  summarise(
    n_seedling_species = n_distinct(SPCD),
    n_seedlings = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop"
  )

# number of individuals per species per plot
CA_SEEDLING_species_summary <- CA_SEEDLING_post_fire_plots %>%
  group_by(plot_id, INVYR, fire_year, SPCD) %>%
  summarise(n_seedlings = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop") %>%
  left_join(REF_SPECIES %>% dplyr::select(SPCD, GENUS, SPECIES),
    by = "SPCD") %>%
  mutate(species = paste(GENUS, SPECIES, sep = "_"))
```


#WA
### fire only plots
```{r}
#create a unique PLOT ID in COND and in PLOT tables 
WA_COND_IDs <- WA_COND %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )
WA_PLOT_IDs <- WA_PLOT %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

# collapse information from multiple CONDIDs associated with each plot to avoid many-to-many mapping
# and then filter by fire and forest type  

#plot_id → multiple CONDIDs → each CONDID has up to 3 disturbance codes and years
# one row for a specific plot_id, CONID, and disturbance event (combining DSTRBCD and DSTRBYR)
WA_COND_long <- WA_COND_IDs %>%
  dplyr::select(plot_id, CONDID, FORTYPCD,
    DSTRBCD1, DSTRBCD2, DSTRBCD3,
    DSTRBYR1, DSTRBYR2, DSTRBYR3) %>%
  pivot_longer(
    cols = starts_with("DSTR"),
    names_to = c(".value", "event"),
    names_pattern = "(DSTRBCD|DSTRBYR)([1-3])"
  ) %>%
  filter(DSTRBCD %in% c(30, 31, 32)) %>%
  filter(FORTYPCD %in% c(371, 201, 221, 222, 224, 225, 261, 922)) %>%
  mutate(
    fire_year_severity = paste(DSTRBYR, DSTRBCD, sep = "_")
  )

# summarize fire history across rows for unique plot_id (one row per plot_id)
WA_fire_summary <- WA_COND_long %>%
  group_by(plot_id) %>%
  summarise(
    n_fire_events = n_distinct(DSTRBYR), # number of distinct fire years per plot
    
    # this just collapses all codes into a list but not necessarily relative to year
    fire_types = paste(sort(unique(DSTRBCD)), collapse = ", "), 
    fire_years = paste(sort(unique(DSTRBYR)), collapse = ", "),
    
    fire_year_severity =
    paste(unique(fire_year_severity), collapse = ", "),
    
    multiple_fires = n_distinct(DSTRBYR) > 1, # did fire occur more than once?
    multiple_severity = n_distinct(DSTRBCD) > 1, # are there multiple fire types?
    first_fire_year = min(DSTRBYR, na.rm = TRUE),
    last_fire_year = max(DSTRBYR, na.rm = TRUE),
    .groups = "drop"
  )

# how many plots burned multiple times?
WA_fire_summary %>% summarise(n_plots_multiple_fires = sum(multiple_fires))
# 4 plots

# how many plots that burned multipe times have different severities?
WA_fire_summary %>% filter(multiple_fires) %>% summarise(n_plots_multiple_fire_and_severity = sum(multiple_severity))
# 0 plots

# how many plots that burned once have multiple severities within a plot?
WA_fire_summary %>%
  filter(!multiple_fires) %>%
  summarise(
    n_plots_single_fire_multiple_severity =
      sum(multiple_severity))
# 5 plots

WA_burned_PLOT_IDs <- WA_fire_summary %>% 
  distinct(plot_id)

# merge COND and PLOT by this unique ID 
WA_PLOT_by_COND <- WA_PLOT_IDs %>%
  semi_join(WA_burned_PLOT_IDs, by = "plot_id") %>%  #keep only plot_ids with at least one fire 
  left_join(WA_fire_summary, by = "plot_id")  #add fire summary to each plot, should be one-to-many mapping

#for each unique plot, count how many unique inventory years (INVYR) it has
WA_PLOT_by_COND %>%
  group_by(plot_id) %>%
  summarise(
    n_revisits = n_distinct(INVYR),
    .groups = "drop"
  ) %>%
  count(n_revisits)

# n=2 (241)


# how many plots have pre and post fire surveys
WA_fire_events <- WA_COND_long %>%
  dplyr::select(plot_id, DSTRBYR) %>%
  distinct() %>%
  rename(fire_year = DSTRBYR)

# does each unique plot_id have a pre and post fire survey around any/all fire years?
WA_prepost_fire_summary <- WA_PLOT_by_COND %>%
  distinct(plot_id, INVYR) %>%
  inner_join(
    WA_fire_events %>% distinct(plot_id, fire_year),
    by = "plot_id"
  ) %>%
  mutate(
    relationship = case_when(
      INVYR < fire_year ~ "pre",
      INVYR > fire_year ~ "post",
      TRUE ~ "same_year"
    )
  ) %>%
  group_by(plot_id, fire_year) %>%
  summarise(
    has_pre  = any(relationship == "pre"),
    has_post = any(relationship == "post"),
    n_pre_surveys  = sum(relationship == "pre"),
    n_post_surveys = sum(relationship == "post"),
    .groups = "drop"
  ) %>%
  mutate(
      status = case_when(
      has_pre & has_post ~ "pre_and_post",
      has_pre ~ "pre_only",
      has_post ~ "post_only",
      TRUE ~ "no_clear_separation"
    )
  )

WA_prepost_fire_summary %>%
  count(status)

# post only = 79
# pre only = 11
#pre and post = 155


# how many plots are there in CA
WA_PLOT %>%
  distinct(LAT, LON) %>%
  nrow()
# 11778

length(unique(WA_COND_IDs$plot_id)) # 11988
length(unique(WA_PLOT_IDs$plot_id)) # 11988


```
## tree
```{r}
#Tree number. A number used to uniquely identify a tree on a subplot. Tree numbers can be used to track trees when PLOT.DESIGNCD or DESIGNCD is the same between inventories

# before looking into tree level data, just confirm that the design codes are the same across plots in our CA_PLOT_by_COND table that has our burned plots
n_distinct(WA_PLOT_by_COND$DESIGNCD) # two design codes across the filtered plots
unique(WA_PLOT_by_COND$DESIGNCD) # that design code is 501, 502

# two different design codes, may be important for relating tree surveys

# do we have surviving trees in post-fire surveys?
# create unique plot ids for CA_TREE
WA_TREE_IDs <- WA_TREE %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

#filter CA_TREE to only keep the unique plot ids from CA_PLOT_by_COND (our fire only plots in our focal forest types)
WA_TREE_filtered <- WA_TREE_IDs %>%
  filter(plot_id %in% WA_PLOT_by_COND$plot_id)

# to locate a specific tree ID, use the following variables 
# STATECD, INVYR, UNITCD, COUNTYCD, PLOT, SUBP, TREE
# TREE is unique to SUBP not plot_id
# we can look for unique combinations of SUBP and TREE within a plot_id (STATECD, UNITCD, COUNTYCD, PLOT)

# to count how many times a tree was surveyed
# group by plot_id, SUBP, TREE, and then count the number of INVYR
WA_TREE_resurveys <- WA_TREE_filtered %>% 
  group_by(plot_id, SUBP, TREE) %>% 
  summarise(n_surveys = n_distinct(INVYR),
            survey_years = paste(sort(unique(INVYR)), collapse = ", "),
            .groups = "drop")

# how many times have trees been resurveyed
WA_TREE_resurveys %>%
  count(n_surveys)

#n=1 (836)
#n=2 (6024)
#n=3 (355)

# maybe make a unique tree identifier too?
#tree_id = paste(plot_id, SUBP, TREE, sep = "_") # also including INVYR?

# lets look at trees in the plots with both pre and post fire surveys
# pull the plot_ids for status = pre_and_post from CA_prepost_fire_summary and the fire year the post-fire plot is associated with

WA_TREE_prepost_fire <- WA_TREE_filtered %>%
  semi_join(
    WA_prepost_fire_summary %>% filter(status == "pre_and_post"),
    by = "plot_id"
  ) %>%
  left_join(
    WA_prepost_fire_summary %>%
      filter(status == "pre_and_post") %>%
      dplyr::select(plot_id, fire_year, status),
    by = "plot_id"
  )

WA_prepost_fire_summary %>%
  filter(status == "pre_and_post") %>%
  summarise(n_plots = n_distinct(plot_id)) #155
length(unique(WA_TREE_prepost_fire$plot_id)) #155

# both CA_fire_summary and CA_TREE_prepost_fire have 304 unique plot_id with pre and post fire inventories

# now we want to look at just post-fire plots (where INVYR > fire_year)
WA_TREE_post_fire_plots <- WA_TREE_prepost_fire %>%
  filter(INVYR > fire_year)

# number of trees per plot (post-fire plots)
WA_TREE_per_plot_post <- WA_TREE_post_fire_plots %>%
  mutate(which = "post") %>%
  distinct(plot_id, INVYR, fire_year, status, SUBP, TREE, which) %>%
  group_by(plot_id, INVYR, fire_year, status, which) %>%
  summarise(n_tree = n(), .groups = "drop")

summary(WA_TREE_per_plot_post$n_tree)

# get the abundance of live and dead trees per species in each plot 
WA_TREE_post_fire_plots_tree_status <- WA_TREE_post_fire_plots %>%
  group_by(plot_id, INVYR, fire_year, SPCD, STATUSCD) %>%
  summarise(n_tree = n(), .groups = "drop") %>%
  left_join(REF_SPECIES %>% dplyr::select(SPCD, GENUS, SPECIES), 
            by = "SPCD")

# STATUSCD = 0 (not present), STATUSCD = 1 (live), STATUSCD = 2 (dead)

# number of species with STATUSCD = 1 in post-fire plots
WA_TREE_post_fire_plot_summary <- WA_TREE_post_fire_plots %>%
  filter(STATUSCD == 1) %>%
  group_by(plot_id, INVYR, fire_year) %>%
  summarise(
    n_live_species = n_distinct(SPCD),
    n_live_trees = n(),
    .groups = "drop")

# number of individuals (live) per species per plot
WA_TREE_post_fire_species_summary <- WA_TREE_post_fire_plots %>%
  filter(STATUSCD == 1) %>%
  group_by(plot_id, INVYR, fire_year, SPCD) %>%
  summarise(
    n_live_trees = n(),
    .groups = "drop"
  ) %>%
  left_join(
    REF_SPECIES %>%
      dplyr::select(SPCD, GENUS, SPECIES),
    by = "SPCD"
  ) %>%
  mutate(species = paste(GENUS, SPECIES, sep = "_"))

### visualize plot level summaries

# number of live species per plot (post-fire plots)
WA_TREE_post_fire_species_summary %>%
  group_by(plot_id) %>%
  summarise(n_species = n_distinct(SPCD)) %>%
  ggplot(aes(x = n_species)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  theme_bw() +
  labs(x = "Number of live species per plot", y = "Number of plots")

# abundance of each species per plot (top 20 most abundant species)
ggplot(
  WA_TREE_post_fire_species_summary %>%
    group_by(species, plot_id) %>%
    summarise(n_live_trees = sum(n_live_trees), .groups = "drop") %>%
    group_by(species) %>%
    summarise(mean_per_plot = mean(n_live_trees), .groups = "drop") %>%
    slice_max(mean_per_plot, n = 20),
  aes(x = reorder(species, mean_per_plot),
      y = mean_per_plot)
) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    x = "species",
    y = "average # live trees per plot",
    title = "average abundance of species per plot"
  ) +
  theme_bw()

# boxplots of tree abundance per species per plot
WA_TREE_post_fire_plots %>%
  group_by(plot_id, SPCD) %>%
  summarise(n_trees = n(), .groups = "drop") %>%
  left_join(
    REF_SPECIES %>% dplyr:: select(SPCD, GENUS, SPECIES),
    by = "SPCD"
  ) %>%
  mutate(species = paste(GENUS, SPECIES, sep = "_")) %>%
  ggplot(aes(x = species, y = n_trees)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6) +
  coord_flip() +
  labs(
    x = "species",
    y = "trees per plot",
    title = "tree abundance per species per plot"
  ) +
  scale_y_continuous(
    breaks = function(x) seq(0, max(x, na.rm = TRUE), by = 5)
  ) +
  theme_bw()

```

## seedling
```{r}
# is there recruitment in post-fire surveys?

# seedling counts are under TREECOUNT in CA_SEEDLING
# TREECOUNT is for  (plot × subplot/microplot × species × inventory)
# in the CA_SEEDLING table, SUBP = microplot since seedlings are inventoried at the microplot within a subplot

# create unique plot ids for CA_SEEDLING
WA_SEEDLING_IDs <- WA_SEEDLING %>%
  dplyr::mutate(
    plot_id = paste(STATECD, UNITCD, COUNTYCD, PLOT, sep = "_")
  )

#filter CA_SEEDLING to only keep the unique plot ids from CA_PLOT_by_COND
WA_SEEDLING_filtered <- WA_SEEDLING_IDs %>%
  filter(plot_id %in% WA_PLOT_by_COND$plot_id)

# lets look at trees in the plots with both pre and post fire 
# pull the plot_ids for status = pre_and_post from CA_prepost_fire_summary

WA_SEEDLING_prepost_fire <- WA_SEEDLING_filtered %>%
  semi_join(WA_prepost_fire_summary %>% filter(status == "pre_and_post"),
    by = "plot_id") %>%
  left_join(WA_prepost_fire_summary %>%
      filter(status == "pre_and_post") %>%
      dplyr::select(plot_id, fire_year, status),
    by = "plot_id")

# filter to post-fire plots only
WA_SEEDLING_post_fire_plots <- WA_SEEDLING_prepost_fire %>%
  filter(INVYR > fire_year)

length(unique(WA_SEEDLING_post_fire_plots$plot_id)) # 68
length(unique(WA_TREE_post_fire_plots$plot_id)) # 155
# 68 unique plots (why is this different from the 155 for the TREE analysis ??)

# number of seedlings per plot (aggregated across SUBP/microplot)
WA_SEEDLING_per_plot <- WA_SEEDLING_post_fire_plots %>%
  group_by(plot_id, INVYR, fire_year) %>%
  summarise(
    n_seedlings = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop"
  )

hist(WA_SEEDLING_per_plot$n_seedlings, breaks=30)
summary(WA_SEEDLING_per_plot$n_seedlings)

# number of seedlings per species per plot (aggregated across SUBP/microplot)
WA_SEEDLING_species_summary <- WA_SEEDLING_post_fire_plots %>%
  group_by(plot_id, INVYR, fire_year, SPCD) %>%
  summarise(n_seedlings = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop") %>%
  left_join(REF_SPECIES %>% dplyr::select(SPCD, GENUS, SPECIES),
    by = "SPCD") %>%
  mutate(species = paste(GENUS, SPECIES, sep = "_"))

WA_SEEDLING_species_summary %>%
  group_by(plot_id) %>%
  summarise(n_species = n_distinct(SPCD)) %>%
  ggplot(aes(x = n_species)) +
  geom_histogram(binwidth = 1, fill = "darkgreen", color = "white") +
  theme_bw() +
  labs(
    x = "Number of seedling species per plot",
    y = "Number of plots"
  )

# number of species of seedlings per plot 
WA_SEEDLING_plot_summary <- WA_SEEDLING_post_fire_plots %>%
  group_by(plot_id, INVYR, fire_year) %>%
  summarise(
    n_seedling_species = n_distinct(SPCD),
    n_seedlings = sum(TREECOUNT, na.rm = TRUE),
    .groups = "drop"
  )

# average number of seedlings per species per plot
ggplot(
  WA_SEEDLING_species_summary %>%
    group_by(species, plot_id) %>%
    summarise(n_seedlings = sum(n_seedlings), .groups = "drop") %>%
    group_by(species) %>%
    summarise(mean_per_plot = mean(n_seedlings), .groups = "drop") %>%
    slice_max(mean_per_plot, n = 20),
  aes(
    x = reorder(species, mean_per_plot),
    y = mean_per_plot
  )
) +
  geom_col(fill = "darkgreen") +
  coord_flip() +
  scale_y_continuous(
    breaks = function(x) seq(0, max(x, na.rm = TRUE), by = 5)
  ) +
  labs(
    x = "species",
    y = "average # seedlings per plot",
    title = "average seedling abundance per species per plot"
  ) +
  theme_bw()

# how many plots have 0 species?
WA_SEEDLING_post_fire_plots %>%
  group_by(plot_id) %>%
  summarise(n_species = n_distinct(SPCD[TREECOUNT > 0])) %>%
  filter(n_species == 0)
# no plots have zero species as seedlings

# boxplots of seedling abundances per species per plot
WA_SEEDLING_species_summary %>%
  group_by(plot_id, species) %>%
  summarise(n_seedlings = sum(n_seedlings), .groups = "drop") %>%
  
  mutate(n_seedlings_capped = pmin(n_seedlings, 50)) %>%
  
  ggplot(aes(x = species, y = n_seedlings_capped)) +
  geom_boxplot(fill = "darkgreen", alpha = 0.6) +
  coord_flip() +
  labs(
    x = "species",
    y = "seedlings per plot (capped at 50)",
    title = "seedling abundance per species per plot"
  ) +
  theme_bw()

```

# species list
```{r}
# now lets aggregate a species list for each state
# STATE_TREE links to STAT_COND via the common identifier PLT_CN
# take the PLT_CN codes from forest_type_combined and filter STATE_TREE
CA_TREE_fire_only <- CA_TREE %>%
  filter(PLT_CN %in% CA_forest_type_combined$PLT_CN)

WA_TREE_fire_only <- WA_TREE %>%
  filter(PLT_CN %in% WA_forest_type_combined$PLT_CN)

OR_TREE_fire_only <- OR_TREE %>%
  filter(PLT_CN %in% OR_forest_type_combined$PLT_CN)

ID_TREE_fire_only <- ID_TREE %>%
  filter(PLT_CN %in% ID_forest_type_combined$PLT_CN)

MT_TREE_fire_only <- MT_TREE %>%
  filter(PLT_CN %in% MT_forest_type_combined$PLT_CN)

CO_TREE_fire_only <- CO_TREE %>%
  filter(PLT_CN %in% CO_forest_type_combined$PLT_CN)

NV_TREE_fire_only <- NV_TREE %>%
  filter(PLT_CN %in% NV_forest_type_combined$PLT_CN)

UT_TREE_fire_only <- UT_TREE %>%
  filter(PLT_CN %in% UT_forest_type_combined$PLT_CN)

WY_TREE_fire_only <- WY_TREE %>%
  filter(PLT_CN %in% WY_forest_type_combined$PLT_CN)

# attach SCIENTIFIC_NAME from REF_SPECIES to SPCD in CA_TREE_fire_only_spp_ID
CA_TREE_fire_only_spp_ID <- CA_TREE_fire_only %>%
  left_join(REF_SPECIES %>% dplyr::select(SPCD, SCIENTIFIC_NAME), by = "SPCD")

WA_TREE_fire_only_spp_ID <- WA_TREE_fire_only %>%
  left_join(REF_SPECIES %>% dplyr::select(SPCD, SCIENTIFIC_NAME), by = "SPCD")

OR_TREE_fire_only_spp_ID <- OR_TREE_fire_only %>%
  left_join(REF_SPECIES %>% dplyr::select(SPCD, SCIENTIFIC_NAME), by = "SPCD")

ID_TREE_fire_only_spp_ID <- ID_TREE_fire_only %>%
  left_join(REF_SPECIES %>% dplyr::select(SPCD, SCIENTIFIC_NAME), by = "SPCD")

MT_TREE_fire_only_spp_ID <- MT_TREE_fire_only %>%
  left_join(REF_SPECIES %>% dplyr::select(SPCD, SCIENTIFIC_NAME), by = "SPCD")

CO_TREE_fire_only_spp_ID <- CO_TREE_fire_only %>%
  left_join(REF_SPECIES %>% dplyr::select(SPCD, SCIENTIFIC_NAME), by = "SPCD")

NV_TREE_fire_only_spp_ID <- NV_TREE_fire_only %>%
  left_join(REF_SPECIES %>% dplyr::select(SPCD, SCIENTIFIC_NAME), by = "SPCD")

UT_TREE_fire_only_spp_ID <- UT_TREE_fire_only %>%
  left_join(REF_SPECIES %>% dplyr::select(SPCD, SCIENTIFIC_NAME), by = "SPCD")

WY_TREE_fire_only_spp_ID <- WY_TREE_fire_only %>%
  left_join(REF_SPECIES %>% dplyr::select(SPCD, SCIENTIFIC_NAME), by = "SPCD")

# create a list of tree species in fire only plots
CA_TREE_fire_only_spp_list <- unique(CA_TREE_fire_only_spp_ID$SCIENTIFIC_NAME)

WA_TREE_fire_only_spp_list <- unique(WA_TREE_fire_only_spp_ID$SCIENTIFIC_NAME)

OR_TREE_fire_only_spp_list <- unique(OR_TREE_fire_only_spp_ID$SCIENTIFIC_NAME)

ID_TREE_fire_only_spp_list <- unique(ID_TREE_fire_only_spp_ID$SCIENTIFIC_NAME)

MT_TREE_fire_only_spp_list <- unique(MT_TREE_fire_only_spp_ID$SCIENTIFIC_NAME)

CO_TREE_fire_only_spp_list <- unique(CO_TREE_fire_only_spp_ID$SCIENTIFIC_NAME)

NV_TREE_fire_only_spp_list <- unique(NV_TREE_fire_only_spp_ID$SCIENTIFIC_NAME)

UT_TREE_fire_only_spp_list <- unique(UT_TREE_fire_only_spp_ID$SCIENTIFIC_NAME)

WY_TREE_fire_only_spp_list <- unique(WY_TREE_fire_only_spp_ID$SCIENTIFIC_NAME)

# can we now take the species lists for each state and find the intersection?
#(i.e. we want a list of total unique species across states)

# merge all states species lists and keep only unique species
total_unique_species <- unique(c(
  CA_TREE_fire_only_spp_list,
  WA_TREE_fire_only_spp_list,
  OR_TREE_fire_only_spp_list,
  ID_TREE_fire_only_spp_list,
  MT_TREE_fire_only_spp_list,
  CO_TREE_fire_only_spp_list,
  NV_TREE_fire_only_spp_list,
  UT_TREE_fire_only_spp_list,
  WY_TREE_fire_only_spp_list
))

write.csv(total_unique_species, file = "fire-severity-unique-species-list.csv")
```


# plot locations/distribution
```{r}
# try plotting locations on a map of the US
# color code by burn severity to see the spatial distribution of high and low severity

#extract plot locations from CA_PLOT or 'CA' for the PLT_CN's on CA_fire_only

CA_burned_plot_coordinates <- CA_TREE_fire_only %>%
  left_join(
    CA %>% dplyr::select(CN, LAT, LON),
    by = c("PLT_CN" = "CN")
  )


# Convert to sf object using LAT and LON columns
CA_burned_sf <- st_as_sf(
  CA_burned_plot_coordinates,
  coords = c("LON", "LAT"),  # Order matters: (lon, lat)
  crs = 4326  # WGS84 coordinate system
)

# Get US states shapefile
usa_shapefile <- ne_states(country = "United States of America", returnclass = "sf")

# Filter for western states
western_states <- usa_shapefile %>%
  filter(name %in% c(
    "California", "Oregon", "Washington", "Nevada", "Idaho", "Utah",
    "Arizona", "New Mexico", "Montana", "Wyoming", "Colorado"
  ))


# Join disturbance code into your plots dataframe
CA_burned_plot_coordinates_DSTRBCD1 <- CA_burned_plot_coordinates %>%
  left_join(
    forest_type_combined %>% dplyr::select(PLT_CN, DSTRBCD1, DSTRBYR1),
    by = "PLT_CN"
  )

# Convert to sf with disturbance code included
CA_burned_sf <- st_as_sf(
  CA_burned_plot_coordinates_DSTRBCD1,
  coords = c("LON", "LAT"),
  crs = 4326
)

ggplot() +
  geom_sf(data = western_states, fill = "gray95", color = "black") +
  geom_sf(
    data = CA_burned_sf %>% filter(DSTRBCD1 %in% c(31, 32)),
    aes(color = factor(DSTRBCD1)),
    size = 1, alpha = 0.8
  ) +
  scale_color_manual(
    values = c("31" = "orange", "32" = "red"),
    labels = c("31" = "Disturbance 31", "32" = "Disturbance 32"),
    name = "Disturbance Code"
  ) +
  coord_sf(xlim = c(-125, -100), ylim = c(30, 45)) +
  theme_minimal() +
  labs(
    title = "Burned plots in low-moderate and high severity",
    x = "Longitude",
    y = "Latitude"
  )

# lets add other states now

# 1. Join coordinates from the state-level plot table
OR_burned_plot_coordinates <- OR_TREE_fire_only %>%
  left_join(
    OR %>% dplyr::select(CN, LAT, LON),
    by = c("PLT_CN" = "CN")
  )

# 2. Join in the disturbance code from forest_type_combined
OR_burned_plot_coordinates_DSTRBCD1 <- OR_burned_plot_coordinates %>%
  left_join(
    OR_forest_type_combined %>% dplyr::select(PLT_CN, DSTRBCD1, DSTRBYR1),
    by = "PLT_CN"
  )

# 3. Convert to sf object with disturbance info
OR_burned_sf <- st_as_sf(
  OR_burned_plot_coordinates_DSTRBCD1,
  coords = c("LON", "LAT"),
  crs = 4326
)


# 1. Join coordinates from the state-level plot table
WA_burned_plot_coordinates <- WA_TREE_fire_only %>%
  left_join(
    WA %>% dplyr::select(CN, LAT, LON),
    by = c("PLT_CN" = "CN")
  )

# 2. Join in the disturbance code from forest_type_combined
WA_burned_plot_coordinates_DSTRBCD1 <- WA_burned_plot_coordinates %>%
  left_join(
    WA_forest_type_combined %>% dplyr::select(PLT_CN, DSTRBCD1, DSTRBYR1),
    by = "PLT_CN"
  )

# 3. Convert to sf object with disturbance info
WA_burned_sf <- st_as_sf(
  WA_burned_plot_coordinates_DSTRBCD1,
  coords = c("LON", "LAT"),
  crs = 4326
)

# 1. Join coordinates from the state-level plot table
ID_burned_plot_coordinates <- ID_TREE_fire_only %>%
  left_join(
    ID %>% dplyr::select(CN, LAT, LON),
    by = c("PLT_CN" = "CN")
  )

# 2. Join in the disturbance code from forest_type_combined
ID_burned_plot_coordinates_DSTRBCD1 <- ID_burned_plot_coordinates %>%
  left_join(
    ID_forest_type_combined %>% dplyr::select(PLT_CN, DSTRBCD1),
    by = "PLT_CN"
  )

# 3. Convert to sf object with disturbance info
ID_burned_sf <- st_as_sf(
  ID_burned_plot_coordinates_DSTRBCD1,
  coords = c("LON", "LAT"),
  crs = 4326
)


# 1. Join coordinates from the state-level plot table
MT_burned_plot_coordinates <- MT_TREE_fire_only %>%
  left_join(
    MT %>% dplyr::select(CN, LAT, LON),
    by = c("PLT_CN" = "CN")
  )

# 2. Join in the disturbance code from forest_type_combined
MT_burned_plot_coordinates_DSTRBCD1 <- MT_burned_plot_coordinates %>%
  left_join(
    MT_forest_type_combined %>% dplyr::select(PLT_CN, DSTRBCD1),
    by = "PLT_CN"
  )

# 3. Convert to sf object with disturbance info
MT_burned_sf <- st_as_sf(
  MT_burned_plot_coordinates_DSTRBCD1,
  coords = c("LON", "LAT"),
  crs = 4326
)

# 1. Join coordinates from the state-level plot table
CO_burned_plot_coordinates <- CO_TREE_fire_only %>%
  left_join(
    CO %>% dplyr::select(CN, LAT, LON),
    by = c("PLT_CN" = "CN")
  )

# 2. Join in the disturbance code from forest_type_combined
CO_burned_plot_coordinates_DSTRBCD1 <- CO_burned_plot_coordinates %>%
  left_join(
    CO_forest_type_combined %>% dplyr::select(PLT_CN, DSTRBCD1),
    by = "PLT_CN"
  )

# 3. Convert to sf object with disturbance info
CO_burned_sf <- st_as_sf(
  CO_burned_plot_coordinates_DSTRBCD1,
  coords = c("LON", "LAT"),
  crs = 4326
)

# 1. Join coordinates from the state-level plot table
UT_burned_plot_coordinates <- UT_TREE_fire_only %>%
  left_join(
    UT %>% dplyr::select(CN, LAT, LON),
    by = c("PLT_CN" = "CN")
  )

# 2. Join in the disturbance code from forest_type_combined
UT_burned_plot_coordinates_DSTRBCD1 <- UT_burned_plot_coordinates %>%
  left_join(
    UT_forest_type_combined %>% dplyr::select(PLT_CN, DSTRBCD1),
    by = "PLT_CN"
  )

# 3. Convert to sf object with disturbance info
UT_burned_sf <- st_as_sf(
  UT_burned_plot_coordinates_DSTRBCD1,
  coords = c("LON", "LAT"),
  crs = 4326
)


# 1. Join coordinates from the state-level plot table
WY_burned_plot_coordinates <- WY_TREE_fire_only %>%
  left_join(
    WY %>% dplyr::select(CN, LAT, LON),
    by = c("PLT_CN" = "CN")
  )

# 2. Join in the disturbance code from forest_type_combined
WY_burned_plot_coordinates_DSTRBCD1 <- WY_burned_plot_coordinates %>%
  left_join(
    WY_forest_type_combined %>% dplyr::select(PLT_CN, DSTRBCD1),
    by = "PLT_CN"
  )

# 3. Convert to sf object with disturbance info
WY_burned_sf <- st_as_sf(
  WY_burned_plot_coordinates_DSTRBCD1,
  coords = c("LON", "LAT"),
  crs = 4326
)

###### this chunk for NV didnt work, see below instead
# 1. Join coordinates from the state-level plot table
NV_burned_plot_coordinates <- NV_TREE_fire_only %>%
  left_join(
    NV %>% select(CN, LAT, LON),
    by = c("PLT_CN" = "CN")
  )

# 2. Join in the disturbance code from forest_type_combined
NV_burned_plot_coordinates_DSTRBCD1 <- NV_burned_plot_coordinates %>%
  left_join(
    NV_forest_type_combined %>% select(PLT_CN, DSTRBCD1),
    by = "PLT_CN"
  )

# 3. Convert to sf object with disturbance info
NV_burned_sf <- st_as_sf(
  NV_burned_plot_coordinates_DSTRBCD1,
  coords = c("LON", "LAT"),
  crs = 4326
)

######
NV_fire_coords <- NV_fire_only %>%
  left_join(NV %>% dplyr::select(CN, LAT, LON), by = c("PLT_CN" = "CN"))

NV_fire_sf <- st_as_sf(NV_fire_coords, coords = c("LON", "LAT"), crs = 4326)



NV_burned_plot_coordinates <- NV_fire_only %>%
  left_join(NV %>% dplyr::select(CN, LAT, LON), by = c("PLT_CN" = "CN"))

# Step 3: Convert to sf object
NV_burned_sf <- st_as_sf(NV_burned_plot_coordinates, coords = c("LON", "LAT"), crs = 4326)

normalize_types <- function(x) {
  dplyr::mutate(
    x,
    dplyr::across(
      where(~ !inherits(.x, "sfc")),
      as.character
    )
  )
}



all_burned_sf <- dplyr::bind_rows(
  normalize_types(CA_burned_sf),
  normalize_types(OR_burned_sf),
  normalize_types(WA_burned_sf),
  normalize_types(ID_burned_sf),
  normalize_types(MT_burned_sf),
  normalize_types(CO_burned_sf),
  normalize_types(UT_burned_sf),
  normalize_types(NV_burned_sf),
  normalize_types(WY_burned_sf)
)




ggplot() +
  geom_sf(data = western_states, fill = "gray95", color = "black") +
  geom_sf(
    data = all_burned_sf %>% filter(DSTRBCD1 %in% c(31, 32)),
    aes(color = factor(DSTRBCD1)),
    size = 1, alpha = 0.8
  ) +
  scale_color_manual(
    values = c("31" = "orange", "32" = "red"),
    name = "Disturbance Code"
  ) +
  coord_sf(xlim = c(-125, -100), ylim = c(30, 50)) +
  theme_minimal() +
  labs(
    x = "Longitude",
    y = "Latitude"
  )
```
# plot level summary stats
```{r}
# lets visualize some summary stats about our burned and unburned plots

h1 <- ggplot(
  data = CA_burned_plot_coordinates_DSTRBCD1 %>%
    filter(DSTRBYR1 <= 2025) %>%
    group_by(PLT_CN) %>%
    slice_min(DSTRBYR1, with_ties = FALSE) %>%
    ungroup(),  # keep one row per unique plot
  aes(x = DSTRBYR1)
) +
  geom_histogram(binwidth = 1, boundary = 1990, color = "black", fill = "steelblue") +
  scale_x_continuous(breaks = seq(1990, 2025, by = 1)) +
  labs(
    title = "Histogram of Disturbance Years in CA (using first year fire was recorded)",
    x = "Disturbance Year",
    y = "Count (Unique PLT_CN)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

h1

h2 <- ggplot(
  data = OR_burned_plot_coordinates_DSTRBCD1 %>%
    filter(DSTRBYR1 >= 1990, DSTRBYR1 <= 2025) %>%
    group_by(PLT_CN) %>%
    slice_min(DSTRBYR1, with_ties = FALSE) %>%
    ungroup(),
  aes(x = DSTRBYR1)
) +
  geom_histogram(binwidth = 1, boundary = 1990, color = "black", fill = "steelblue") +
  scale_x_continuous(breaks = seq(1990, 2025, by = 1)) +
  labs(
    title = "Histogram of Disturbance Years in OR (using first year fire was recorded)",
    x = "Disturbance Year",
    y = "Count (Unique Plots)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))


h2



h3 <- ggplot(
  data = WA_burned_plot_coordinates_DSTRBCD1 %>%
    filter(DSTRBYR1 <= 2025) %>%
    group_by(PLT_CN) %>%
    slice_min(DSTRBYR1, with_ties = FALSE) %>%
    ungroup(),  # keep one row per unique plot
  aes(x = DSTRBYR1)
) +
  geom_histogram(binwidth = 1, boundary = 1990, color = "black", fill = "steelblue") +
  scale_x_continuous(breaks = seq(1990, 2025, by = 1)) +
  labs(
    title = "Histogram of Disturbance Years in WA (using first year fire was recorded)",
    x = "Disturbance Year",
    y = "Count (Unique Plots)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))



h3
grid.arrange (h1, h2, h3)

h4 <- ggplot(
  data = CA_burned_plot_coordinates_DSTRBCD1 %>%
    filter(!is.na(DSTRBYR1), DSTRBYR1 >= 1990, DSTRBYR1 <= 2025) %>%
    group_by(PLT_CN) %>%
    slice_min(DSTRBYR1, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(TimeSinceFire = 2025 - DSTRBYR1),
  aes(x = TimeSinceFire)
) +
  geom_histogram(binwidth = 1, boundary = 0, color = "black", fill = "darkorange") +
  scale_x_continuous(breaks = seq(0, 35, by = 1)) +
  labs(
    title = "CA",
    x = "Time Since Fire (Years)",
    y = "Count (Unique Plots)"
  ) +
  theme_minimal()
h4

h5 <- ggplot(
  data = OR_burned_plot_coordinates_DSTRBCD1 %>%
    filter(!is.na(DSTRBYR1), DSTRBYR1 >= 1990, DSTRBYR1 <= 2025) %>%
    group_by(PLT_CN) %>%
    slice_min(DSTRBYR1, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(TimeSinceFire = 2025 - DSTRBYR1),
  aes(x = TimeSinceFire)
) +
  geom_histogram(binwidth = 1, boundary = 0, color = "black", fill = "darkorange") +
  scale_x_continuous(breaks = seq(0, 35, by = 1)) +
  labs(
    title = "OR",
    x = "Time Since Fire (Years)",
    y = "Count (Unique Plots)"
  ) +
  theme_minimal()

h5

h6 <- ggplot(
  data = WA_burned_plot_coordinates_DSTRBCD1 %>%
    filter(!is.na(DSTRBYR1), DSTRBYR1 >= 1990, DSTRBYR1 <= 2025) %>%
    group_by(PLT_CN) %>%
    slice_min(DSTRBYR1, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(TimeSinceFire = 2025 - DSTRBYR1),
  aes(x = TimeSinceFire)
) +
  geom_histogram(binwidth = 1, boundary = 0, color = "black", fill = "darkorange") +
  scale_x_continuous(breaks = seq(0, 35, by = 1)) +
  labs(
    title = "WA",
    x = "Time Since Fire (Years)",
    y = "Count (Unique Plots)"
  ) +
  theme_minimal()

h6

grid.arrange (h4, h5, h6)


# 600+ plots in ground fire and 400+ plots in canopy fire in CA
ggplot(dplyr::filter(CA_cond, DSTRBCD1 %in% c(30, 31, 32)),
       aes(x = factor(DSTRBCD1))) +
  geom_bar() +
  labs(x = "Disturbance Code", y = "Count",
       title = "Fire-Related Disturbances (Codes 30–32)") +
  theme_minimal()
```

# climate extraction 
```{r}
```

# understory
```{r}
# lets look at the understory too...

# attach SCIENTIFIC_NAME from REF_SPECIES to CA_veg_understory_fire_only by matching VEG_SPCD
# to SPECIES_SYMBOL
CA_veg_understory_fire_only_spp_ID <- CA_veg_understory_fire_only %>%
  left_join(REF_SPECIES %>%
              select(SPECIES_SYMBOL, SCIENTIFIC_NAME),
            by = c("VEG_SPCD" = "SPECIES_SYMBOL")) ## this has the be redone, only gives trees

# take the PLT_CN codes from forest_type_combined and filter CA_veg_understory to those PLT_CN's
CA_veg_understory <- CA_P2VEG_SUBPLOT_SPP

CA_veg_understory_fire_only <- CA_veg_understory %>%
  filter(PLT_CN %in% forest_type_combined$PLT_CN)

# CA_veg_understory_fire_only uses VEG_SPCD instead of SPCD



# create a list of understory species in fire only plots



# how many resurveys do fire only mixed conifer plots in CA have?
#CA_PLOT (or just CA here) links to CA_COND via the common identifier PLT_CN


#If INVYR (plot visit year) < DSTRBYR1 → then this visit happened before the fire.

#If INVYR > DSTRBYR1 → then this visit was after the fire.

#If equal, the fire occurred in the same year as the plot visit

# if we want to link CA_PLOT (or CA) with CA_cond that has been filtered by disturbance codes
# and forest types then we want to lin forest_type_combined with CA_PLOT or CA but PLT_CN in
#CA_PLOT of CA PLT_CN is the same as CN

inventory_year <- CA_cond %>%
  filter(DSTRBCD1 %in% c(31,32) | DSTRBCD2 %in% c(31,32) | DSTRBCD3 %in% c(31,32)) %>%
  left_join(CA, by = "PLT_CN") %>%
  mutate(fire_year = coalesce(DSTRBYR1, DSTRBYR2, DSTRBYR3),
         pre_fire = INVYR < fire_year,
         post_fire  = INVYR > fire_year)


resurveyed_fire_only_plots <- CA_TREE_fire_only %>%
  group_by(PLT_CN) %>%
  summarise(n_years = n_distinct(INVYR))
# all plots only have 1 visit?

resurveyed_fire_only_plots <- CA_TREE_fire_only %>%
  group_by(PLT_CN) %>%
  summarise(INVYR_list = list(sort(unique(INVYR))))


```



